Activity modulation in anaerobic ribonucleotide reductases: nucleotide binding to the ATP-cone mediates long-range order-disorder transitions in the active site

Version 1

Abstract

Abstract

A small, nucleotide-binding domain, the ATP-cone, is found at the N-terminus of most ribonucleotide reductase (RNR) catalytic subunits. By binding ATP or dATP it regulates the enzyme activity of all classes of RNR. Functional and structural work on aerobic RNRs has revealed a plethora of ways in which dATP inhibits activity by inducing oligomerization and preventing a productive radical transfer from one subunit to the active site in the other. Anaerobic RNRs, on the other hand, store a stable glycyl radical next to the active site and the basis for their dATP-dependent inhibition is completely unknown. We present biochemical, biophysical and structural information on the effects of ATP and dATP binding to the anaerobic RNR from Prevotella copri. The enzyme exists in a dimer-tetramer equilibrium biased towards dimers when two ATP molecules are bound and tetramers when two dATP molecules are bound. In the presence of ATP, P. copri NrdD is active and has a fully ordered glycyl radical domain (GRD) in one monomer of the dimer. Binding of dATP to the ATP-cone results in loss of activity and disordering of the GRD. The glycyl radical is formed even in the dATP-bound form, but the substrate does not bind, suggesting that dATP inhibition in anaerobic RNRs acts by disordering of the GRD more than 30 Å away from the dATP molecule, thereby preventing both substrate binding and radical mobilisation. The structures implicate a complex network of activity regulation involving the GRD, the allosteric substrate specificity site and a conserved but previously unseen flap over the active site.

Introduction

Ribonucleotide reductases (RNRs) are a family of enzymes with sophisticated radical chemistries and allosteric regulation. RNRs produce all four deoxyribonucleotides and are the only enzymes providing de novo building blocks for DNA replication and repair in all free-living organisms. Virtually all RNRs possess an allosteric site regulating substrate specificity (the s-site), a crucial aspect of RNR that ultimately provides a balanced supply of dNTPs (Mathews, 2016, 2018). A second allosteric site (the a-site) is located in an N-terminal domain called the ATP-cone (Aravind et al, 2000), found in the majority of RNRs (Jonna et al, 2015). Whereas the s-site binds all deoxyribonucleoside triphosphates (dNTPs) and usually ATP, the a-site in the ATP-cone can only bind ATP and dATP. Binding of ATP activates RNR, and binding of dATP inhibits it (Hofer et al, 2012; Martínez-Carranza et al, 2020).

The RNR family consists of three evolutionarily related classes (I to III) with a common radical-based reaction mechanism but differing in the mode of radical generation and in quaternary structure (Högbom et al, 2020; Lundin et al, 2015). Class I, strictly aerobic RNRs, are heterotetramers consisting of two α subunits (NrdA) with binding sites for substrates and allosteric nucleotides and two β subunits (NrdB) harbouring a stable radical. Class II RNRs are either dimers or monomers with an α subunit (NrdJ) that, in addition to binding sites for substrate and allosteric nucleotides, also harbours the radical initiator cofactor adenosylcobalamin. Class III RNRs, which only function anaerobically, are dimers of an α subunit (NrdD) with a stable radical close to the active site, and binding sites for allosteric nucleotides. The radical is introduced via encounter with a specific radical-SAM enzyme called NrdG, and once activated NrdD can perform multiple turnovers in the absence of NrdG (Backman et al, 2017; Torrents et al, 2001).

The dATP-dependent inhibition has been biochemically characterised in representatives of all three RNR classes, but only structurally studied in class I. A common denominator of all inhibited class I RNRs is oligomerization, leading to disturbed radical transfer between the α and β subunits. Hitherto four different inhibitory oligomerization mechanisms have been identified: heterooctameric α4β4 complexes in Escherichia coli, Clostridium botulinum and Neisseria gonorrhoeae (Ando et al, 2011; Martínez-Carranza et al, 2020; Wei et al, 2014), an α4 complex in Pseudomonas aeruginosa (Johansson et al, 2016), α6 complexes in the eukaryotes Homo sapiens, Saccharomyces cerevisiae and Dictyostelium discoideum (Ando et al, 2016; Brignole et al, 2018; Crona et al, 2013; Fairman et al, 2011), and a β4 complex in the bacterium Leeuwenhoekiella blandensis (Rozman Grinberg et al, 2018a). All these class I oligomers involve protein-protein interactions mediated by the ATP-cone domain.

The ATP-cone is in essence restricted to RNRs and the RNR-specific repressor NrdR (Rozman Grinberg et al, 2022). Among RNRs it is found in about 80% of class III enzymes and about 50% of class I enzymes, but less than 10% of class II enzymes (Jonna et al., 2015). The presence and function of the ATP-cone domain distinguish anaerobic RNRs from the other members of the large glycyl radical enzyme (GRE) family that are otherwise structurally and mechanistically related (Backman et al., 2017). The allosteric regulation mechanisms of class III RNRs (NrdDs) from E. coli, bacteriophage T4 and Lactococcus lactis were determined several decades ago (Andersson et al, 2000; Eliasson et al, 1994; Torrents et al, 2000). T4NrdD was the first class III RNR structure solved (Logan et al, 1999), and later, structures of Thermotoga maritima NrdD (TmNrdD) have been published (Aurelius et al, 2015; Wei et al, 2014). However, these two NrdDs lack an ATP-cone and the structural basis for allosteric activity regulation in the class III RNRs is still an outstanding question. Here, we present structural, biochemical and biophysical studies on the ATP-cone-containing NrdD from the human pathogen Prevotella copri (PcNrdD), showing that the binding of dATP causes complete disordering of the C-terminal glycyl radical-bearing domain, as well as a flap-like loop that binds across the top of the active site in the active form. This disordering is coupled to enzymatic inactivation by inhibition of substrate binding and radical transfer. The final outcome of dATP inhibition is thus blocked radical transfer in both class I and class III RNRs, but it is achieved by completely different mechanisms, i.e. oligomerization prevents radical transfer between subunits in class I RNRs but long-range induced order-disorder transitions prevent substrate binding in class III RNRs.

Results

Allosteric activity regulation by ATP and dATP

The amino acid sequence of PcNrdD suggests that it belongs to the formate-requiring class III RNRs (Burnim et al, 2022b; Wei et al., 2014), and initial optimization of the assay composition showed that this was the case (Figure 1 - figure supplement 1). ATP-cone mediated activation of PcNrdD enzyme activity by ATP or inhibition by dATP was tested with two different substrates. With GTP as substrate the s-site was filled with dTTP as effector, and in absence of any a-site effector the basal level of GTP reduction was approximately 25 nmol/min•mg. An increasing concentration of ATP stimulated activity to a Vmax of 470 nmol/min•mg, with an apparent KL of 0.78 mM (Fig. 1a). On the other hand, an increasing concentration of dATP resulted in an abrupt inhibition of enzyme activity with a Ki of 26 µM (Fig. 1b). Hence, the Ki value for dATP inhibition via the ATP-cone is 30-fold lower compared to the KL value for ATP activation via the ATP-cone.

Activity assays of Prevotella copri NrdD in the presence of ATP or dATP.

A 2.5-fold excess of holoNrdG over apoNrdD was used to study the allosteric regulation effect of ATP and dATP on the a-site. GTP reduction was monitored with up to 1 mM dTTP as effector in the s-site and titrated with ATP (a) or dATP (b) in the a-site. CTP reduction assays were titrated with ATP (c) or dATP (d), in this case acting both as s-site effectors and a-site regulators. Curve fits for calculation of KL and Ki used equations given in Materials and Methods, for (c) and (d) assuming a start activity of 50 nmol/min/mg; curve fit in (d) used a combination of the two equations.

The activation/inhibition experiments were also carried out with CTP as substrate. Specificity effectors for CTP reduction are ATP or dATP, so in these experiments both the s- and a-sites were conceivably filled with ATP or dATP, respectively. The basal level of CTP reduction in the absence of effector nucleotide was approximately 50 nmol/min•mg. An increasing concentration of ATP again stimulates enzyme activity to a Vmax of 273 nmol/min•mg with an apparent KL of 0.53 mM (Fig. 1c). Addition of dATP initially results in a stimulation of enzyme activity, conceivably when the s-site is filled with dATP, after which competing inhibition appears when dATP also binds to the ATP-cone (Fig. 1d). All in all, these results suggest that the ATP-cone in PcNrdD responds similarly to inhibition by dATP as has been observed before for several class I RNRs.

Binding of nucleotides to the ATP-cone

To confirm binding of ATP or dATP to the ATP-cone we used isothermal titration calorimetry (ITC) and microscale thermophoresis (MST) (Fig. 2). Substrate GTP and s-site effector dTTP were used to fill up the other nucleotide binding sites in PcNrdD. Binding of dATP occurred with a KD of 6 µM, whereas binding of ATP was 4-fold weaker with a KD of 26 µM (Fig. 2a-c). The number of bound dATP molecules observed with ITC was almost 2-fold higher for dATP compared to ATP, but both values were below 1. However, in cryo-EM experiments where higher nucleotide concentrations can be used, we show that the ATP-cone can bind two ATP or two dATP molecules (see below), as also expected based on its sequence. Nucleotide binding measured with MST confirmed our results for ATP (Fig. 2d) that had a KD of 25 µM, and dATP that with this method had a KD of 23 µM (Fig. 2e).

Binding of nucleotides to the ATP-cone.

Binding assay using ITC: (a) Binding of ATP to 50 µM PcNrdD; (b) binding of dATP to 200 µM PcNrdD; (c) thermodynamic parameters of ligand binding. All titrations were performed in the presence of 1 mM GTP substrate and 1 mM s-site effector dTTP at 20 °C. Binding of nucleotides to the ATP-cone using MST: (d) Binding of ATP, (e) binding of dATP. All MST binding experiments were performed in the presence of 1 mM s-site effector dTTP and 5 mM GTP substrate at room temperature.

It is noteworthy that the KD values for ATP and dATP do not differ by orders of magnitude, as has been found for several class I RNRs (Ando et al, 2016; Birgander et al, 2004; Ormö & Sjöberg, 1990; Rozman Grinberg et al., 2018a; Rozman Grinberg et al, 2018b; Torrents et al, 2006). Collectively these two binding methods instead suggested that ATP and dATP may interact at two distinct sites in the ATP-cone for the functional readout of dATP inhibition, as has earlier been observed in the transcriptional repressor NrdR (Rozman Grinberg et al., 2022). To test this hypothesis, apo-PcNrdD was incubated with ATP only, dATP only or a combination of ATP and dATP, then desalted, boiled, centrifuged and the nucleotide content of the supernatant analysed by HPLC. Figure 2 - figure supplement 1 shows that only dATP was bound when PcNrdD was incubated with a combination of ATP and dATP. This was also the case when incubations were performed in the presence of s-site effector and substrate, and also when the desalting was performed in the presence of s-site allosteric effector and substrate. In incubations with only ATP, s-site effector and substrate were needed during the entire work-up procedure for ATP to be retained by the protein. In conclusion, the binding experiments suggest that the PcNrdD ATP-cone can bind either ATP or dATP but not both simultaneously. Importantly, the results point to allosteric communication between the s-site, the active site and the a-site in the ATP-cone, as dTTP and GTP are required for any ATP to be retained. Both of these observations are consistent with the structural analyses shown below.

Modulation of the oligomeric state by ATP and dATP in the presence of allosteric effectors

Binding of dATP to the ATP-cone in all class I RNRs studied to date results in formation of higher oligomers unable to perform catalysis (Martínez-Carranza et al., 2020). We therefore next asked whether a similar mechanism would be valid for dATP-mediated inhibition of a class III RNR. To elucidate the oligomeric states of PcNrdD we used gas-phase electrophoretic mobility molecular analysis (GEMMA) (Kaufman et al, 1996), a method based on time-of-flight mass spectrometry. It requires low protein and nucleotide concentrations as well as volatile buffers.

When loaded with ATP, PcNrdD was in a dimer-tetramer equilibrium shifted towards dimers (Fig. 3, left). Addition of the CTP substrate or the s-site effector dTTP resulted in a similar equilibrium. In contrast, the dimer-tetramer equilibrium of dATP-loaded PcNrdD was shifted towards tetramers, and this equilibrium was likewise not affected by addition of the CTP substrate (Fig. 3, right). Hence, dATP-inhibition of enzyme activity seems to work differently in PcNrdD than the clearcut and drastic change in oligomeric structure observed for class I RNRs, since both ATP-loaded and dATP-loaded forms of PcNrdD were in dimer-tetramer equilibria.

Oligomeric states of Prevotella copri NrdD in the presence of nucleotides determined by GEMMA.

(a) Apo-PcNrdD (2 µM) loaded with the activator ATP (50-100 µM) (black and red) in the presence of CTP (100 µM) as a substrate (cyan) or dTTP (100 µM) (orange) as the allosteric effector. (b) apo-PcNrdD (2 µM) loaded with the inhibitor dATP (50-100 µM) (black and red) in the presence of CTP (50 µM) (cyan) as the substrate.

Glycyl radical formation in presence of ATP or dATP

Next, we asked whether dATP inhibition was mediated by prevention of glycyl radical formation. Apo-PcNrdD was incubated with NrdG and S-adenosylmethionine in presence of ATP or dATP. Whereas reduction of substrate requires the presence of formate, the formation of a glycyl radical does not. We therefore incubated with or without formate as well as with or without substrate. Figure 4 shows that the glycyl radical was readily formed in presence of dATP both with and without the CTP substrate, and the result was similar in the absence of formate (Figure 4 - figure supplement 1). Interestingly, under all conditions the glycyl radical content was higher in the presence of dATP compared to with ATP, and it was also more stable than in absence of any effector nucleotides. In addition, the shape of the doublet EPR signal suggests that the environment of the radical was similar under all conditions studied. Collectively, these results show that the glycyl radical is readily formed in the presence of a dATP concentration that inhibits RNR activity.

Glycyl radical formation after 20 min in presence of formate and ATP±CTP or dATP±CTP.

Nucleotide concentrations were: 1.5 mM ATP, 1 mM dATP, and 1 mM CTP. Traces are arbitrarily moved to increase visibility and scaled to identical units (Y-axes).

Substrate binding to PcNrdD

To study substrate binding to activated and inhibited PcNrdD we used MST with labelled protein and GTP or CTP as substrate. The results were clearcut in both sets of experiments (Fig. 5). ATP-loaded PcNrdD bound GTP in the presence of s-site effector dTTP (Fig. 5a), and it bound CTP in the presence of only ATP (Fig. 5b), whereas dATP-loaded PcNrdD could bind neither GTP nor CTP under similar conditions (Fig 5c,d). These results show that the dATP-inhibition of enzyme activity in a class III RNR is mediated by inhibited binding of substrate.

Binding of substrate to ATP- or dATP-loaded PcNrdD.

Binding of GTP (a), and binding of CTP (b) to ATP-loaded PcNrdD. (c) Binding of GTP, and (d) binding of CTP to dATP-loaded PcNrdD. No additional nucleotides were present in CTP binding experiments, whereas binding of GTP was performed in the presence of 1 mM S-site effector dTTP. Fitted KDs are 2.8 ± 0.5 and 11 ± 2.4 mM for GTP and CTP binding respectively. Experiments were performed at room temperature.

Cryo-EM structure of the ATP-CTP bound PcNrdD dimer

2D classes of particles extracted from micrographs of PcNrdD incubated with effector ATP and substrate CTP indicated only dimers, with no appreciable amounts of tetramer. From a set of 570 730 particles we obtained a map from cryoSPARC at 3.0 Å. Application of deep learning local sharpening in DeepEMhancer (Sanchez-Garcia et al, 2021) improved the quality of the map, and we were able to build a reliable model for the great majority of the structure. As the ATP-cone density still did not permit reliable modelling, 3D classification was carried out on this particle set, enabling the identification of a subset of 291 231 particles that gave a volume at 3.2 Å with better-defined ATP-cones (Figure 6a). This map was used for refinement of the final model (Table 1; Table 1 – table supplement 1).

Structure of the PcNrdD dimer in complex with effector ATP and substrate CTP.

a) cryo-EM map with C1 symmetry showing the best-ordered ATP-cones after post-processing using DeepEMhancer. For the right-hand (active) monomer, the ATP-cone (residues 1-91) and the linking helix (92-104) are coloured olive, the linker and NxN flap region pale green, the glycyl radical loop red and the C-terminal extended region orange. The loop in the middle of the α/β barrel containing the radical initiator cysteine Cys416 at its tip is yellow. b) Overview of the PcNrdD dimer with ATP and CTP. The two monomers of the dimer are coloured in different shades of grey. The ATP and CTP molecules are shown as sticks. c) Schematic of the domain organisation of PcNrdD with the same colour scheme as a) and b); d) Closeup view of the binding of four ATP molecules to the dimer of ATP-cones in PcNrdD. The view is from the bottom of the molecule as seen in a) and b). Disordered loops are shown as dotted lines. e) Closeup view of the active site including the cryo-EM map for CTP. Residues within 4 Å of CTP are shown as sticks and polar interactions as dotted lines. f) Sequence logos of NrdD sequence motifs. A) ATP-cone plus downstream interaction partners, B) central parts of sequence and C) C-terminal parts. The numbering is from PcNrdD. Segments were selected to illustrate amino acids discussed in the text.

PcNrdD is a compact dimer (Fig. 6b). Each monomer consists of 739 residues, of which residues 1-91 constitute the ATP-cone, 92-110 a linker region, 111-121 a flap that folds over the substrate in the active site, 122-671 are the core domain with its 10-stranded α/β barrel fold typical of the RNR/glycyl radical enzyme (GRE) family, and 676-739 are the C-terminal glycyl radical domain (GRD) that contains a structural Zn site followed by the buried loop containing the radical glycine residue at position Gly711 and a C-terminal tail that extends across the surface of the protein (Figure 6b,c). The closest structural neighbours as identified by the DALI server (Holm, 2022) are the NrdDs from bacteriophage T4 (Logan et al., 1999) (1H7A, 2.5 Å root mean square deviation [rmsd] in Cα positions for 535 residues, 27% sequence identity) and T. maritima (Aurelius et al., 2015; Wei et al., 2014) (4U3E, 3.0 Å rmsd for 508 residues, 16% sequence identity; Figure 6 - figure supplement 1). The linker, flap and GRD are well-ordered in one of the monomers (dark blue in Fig. 6), while in the other monomer this region is less well-ordered. We will refer to these as the “active” and “inactive” monomers respectively.

At the concentration of ATP used (1 mM), ATP is not observed to bind to the s-sites near the dimer interface. The ATP-cones containing the a-site are flexible, but the reconstruction allowed the modelling of two ATP molecules per ATP-cone and many of the most important side chains. The ATP molecules are bound similarly to the twin dATP molecules seen in the ATP-cones of P. aeruginosa NrdA (PaNrdA) and L. blandensis NrdB (LbNrdB) (Johansson et al, 2016; Rozman Grinberg et al., 2018a), and to dATP bound to PcNrdD itself (see below). However, the ATP-cones in PcNrdD exhibit a hitherto unobserved relative orientation in which the four triphosphate groups are projected towards each other. (Fig. 6b,d). The large negative charge is compensated for by the proximity of as many as 12 Arg and Lys side chains from both monomers. The ATP-cones interact by contacts involving the “roof” of the domain (residues 1-18). Arg11 of each ATP-cone reaches over to interact with the α-phosphate group of one of the ATP molecules in the other ATP-cone, and a salt bridge is formed between Asp9 at the end of the lid and Arg91 at the end of the last helix of the other ATP-cone. Strikingly, the dimer of ATP-cones is slightly offset from the dimer axis of the core protein, such that one of them approaches the core domain of the monomer with ordered C-terminus, while the other is more distant (Fig. 6a,b). However, the closest atom of any ATP molecule is over 30 Å from the substrate CTP in the active site.

The glycyl radical site is found at Gly711, at the tip of a loop projected from the C-terminal region of the enzyme into the barrel, where it approaches the radical initiator cysteine Cys416 at the tip of a loop projected through the barrel (Fig. 6b,e). The second cysteine necessary for the radical mechanism of PcNrdD is Cys171 on the first β-strand of the barrel (Fig. 6e). Consistent with the fact that PcNrdD belongs to the group of anaerobic RNRs that use formate as overall reductant (Burnim et al, 2022a; Mulliez et al, 1995; Wei et al., 2014), there is no third cysteine residue on the sixth β-strand, and its place is taken by Asn435. As usual for GREs, Gly711 is completely buried within the barrel, with no solvent-accessible surface area. The substrate CTP is bound to the active site. The base, phosphate and most of the ribose are well-defined but the density is weaker around the 5’ C-atom (Fig. 6e).

After the glycyl radical loop emerges from the α/β barrel, it traverses the top of the barrel, forming a short helix from residues 723-730 and ending in an extended tail. This conformation of the C-terminus is very similar to the one seen for TmNrdD in PDB entry 4U3E (Wei et al., 2014). This tail makes few specific interactions with other residues in the core.

When compared to previously determined NrdD structures, the first common structural element is a long α-helix in the core domain stretching from Thr123 to Leu141. Between this and the ATP-cone are a linker (91-110), followed by a “flap” over the top of the active site (111-122), which passes the substrate in the active site. Two residues from the flap make H-bonds to CTP: Asn114 to the γ-phosphate group and Asn116 to the amino group of the cytosine base. We call this region the “NxN flap”, as the NxN motif (114-116) is one of the most highly conserved sequence motifs in the NrdD family (Fig. 6f). This is the first time such interactions have been observed in an NrdD, as the flap corresponds to a disordered segment of 17 residues in the structures of TmNrdD and T4NrdD. However, an AlphaFold2 (Jumper et al, 2021) model of TmNrdD (entry Q9WYL) suggests that this linker potentially has the same conformation in TmNrdD and possibly all other NrdDs, whether or not they have an ATP-cone. As seen previously in TmNrdD, the triphosphate moiety is recognised through the conserved HxHD motif (157-160), H-bonds to the main chain atoms of a loop (577-581) following the third-last strand of the α/β barrel and the dipole of the following helix (Fig. 6f).

Significantly, the NxN flap forms the nexus of a network of interactions (Fig. 7) between the substrate, the linker to the ATP-cone, the C-terminal GRD, and “loop 2” (residues 304-311) that is responsible for communicating the substrate specificity signal from the s-site at the dimer interface to the active site (Aurelius et al., 2015; Larsson et al, 2001; Logan et al., 1999). Important interactions linking the NxN flap to the C-terminus include H-bonds from Tyr712, which immediately follows the radical Gly711, to Glu581, which bridges to both the substrate and Ala115 in the flap, as well as from Arg719 in the GRD to the main chain carbonyl groups of Asn114 and Met117 in the flap. Tyr712 is highly conserved in the NrdD family (Fig. 6f and 7). This nexus is crucial for allosteric activity regulation, as will be discussed later.

The intricate network of interactions between the NxN flap, C-terminal GRD, loop 2 and substrate in the ATP-CTP complex of PcNrdD.

The NxN flap is coloured light green, the GRD orange, loop 2 dark red and the Cys radical loop yellow. The Gly radical loop is semi-transparent for clarity. The CA atom of Gly711 is indicated by an orange sphere. Important hydrogen bonds are shown as dotted lines.

As the NxN flap appears to close off the top of the active site, we used the web server Caver Web v1.1 (Stourac et al, 2019) to analyse substrate access to the active site in the ATP-bound form, and identified a tunnel ∼12 Å long leading from the surface of the protein to the active site that is 5.6 Å in diameter at its narrowest point (Figure 7 - figure supplement 1). Thus, in principle a fresh substrate could diffuse into the active site on each catalytic cycle.

Cryo-EM structure of the ATP-dTTP-GTP dimer

In order to probe the generality of the conformations observed in the ATP-CTP complex, we also solved the structure of PcNrdD in complex with ATP in the a-site, specificity effector dTTP in the s-site, and substrate GTP in the active site (c-site). Again, 2D classes show a predominantly dimeric form with no more than ∼10% tetramers. From a set of 437 886 particles selected from 2D classes with the best apparent density for the ATP-cone region, we obtained a volume at 2.40 Å resolution for the dimer with C1 symmetry that was further improved using DeepEMhancer (Table 1; Table 1 - table supplement 2). dTTP is clearly visible in the s-site (Figure 8 - figure supplement 1a) as is GTP in the active site (Figure 8 - figure supplement 2). A Mg2+ ion can be modelled in the s-site, coordinated by Glu290 and all three phosphate groups of dTTP. A β-hairpin loop between residues 185-192 that is disordered in the ATP-CTP structure in the absence of an s-site nucleotide is here ordered, due to an H-bond to the 2’-OH group of dTTP. Readout of the s-site nucleotide’s identity is achieved through an H-bond to Asn298 (Figure 8 - figure supplement 1).

Structure of PcNrdD in complex with a-site effector ATP, s-site effector dTTP and substrate GTP.

a) Volume for the PcNrdD dimer in complex with ATP, dTTP and GTP. The volume is contoured at a low level to emphasise the weak density for the more ordered ATP-cone domain in the active monomer. b) Ribbon diagram of the ATP-dTTP-GTP complex coloured as in Figure 6b, except that the most N-terminal helix (olive) now extends from residues 85-104. c) Zoom in on panel b) to illustrate the highly asymmetrical loop 2 conformations in the active and inactive monomers. Loop 2 in the inactive monomer (light blue) is coloured dark blue, while in the active monomer (grey) it is dark red. The conformation of loop 2 that forms a complementary cradle for the guanosine base of the substrate GTP in the active monomer induces a conformation of the other loop 2 that precludes substrate binding. d) Severe steric clash of loop 2 of the inactive monomer with the substrate and NxN flap. The view is rotated approximately 180° from panels a-c. Loop 2 from the inactive monomer (wheat) is superimposed on the active site of the active monomer (dark blue). The molecular surface of the loop is shown to emphasise that this conformation is incompatible with an ordered NxN flap (light green) and substrate binding. The Cys radical loop is yellow and Cys416 is marked by a sphere. The Gly radical loop is omitted for clarity.

Like the ATP-CTP complex, the ATP-dTTP-GTP complex can also be partitioned into an active and an inactive monomer. The ATP-cone region is very disordered, but the last helix of the ATP-cone in the active monomer can be traced back to residue 85 (Fig. 8a,b). The helix is not kinked at residue 91 as in the ATP-CTP dimer but forms an uninterrupted helix between residues 85-104. The volume for the ATP-cone region lies entirely on one side of the PcNrdD dimer axis, as is also apparent in some of the 2D classes (Figure 8 – figure supplement 3). The ATP-cone cannot be modelled in detail, but if the volume is contoured at very low level, there is almost enough density for one ATP-cone. Taken together, this suggests that there is one partially disordered ATP-cone and one completely disordered in the ATP-dTTP-GTP complex. In the active monomer, the NxN flap and entire C-terminus are ordered while in the inactive monomer, the entire assembly of linker and NxN flap up to residue 120 and the entire GRD are disordered.

Intriguingly, binding of dTTP to the s-sites at the dimer interface induces two distinct conformations of loop 2 (Fig. 8c). In the active monomer, loop 2 is projected towards the aforementioned network of interactions between NxN flap, GRD and substrate, packing against the flap, though no polar interactions are apparent. However, loop 2 also makes extensive interactions with its equivalent in the inactive monomer, which forces the latter into a conformation where it would sterically clash with an ordered NxN flap (Fig. 8d). Furthermore, Phe412, which stacks on the substrate base in the active monomer, and Arg305, which is projected from the substrate-distal side of loop 2, sterically prevent the substrate binding in the inactive monomer. These may be important contributing factors to the disorder of the NxN flap and GRD. Finally, Gln309, which interacts with the substrate in the active monomer (Figure 8 - figure supplement 2), is oriented away from the substrate through its interactions with the “active” loop 2.

Cryo-EM structure of the ATP-dGTP dimer

As a third insight into the ATP-bound forms of PcNrdD, we solved the structure of PcNrdD in complex with specificity effector dGTP in the s-site and ATP at a concentration where it would both act as allosteric effector at the a-site and as the cognate substrate for dGTP (Table 1; Table 1 - table supplement 3). Density for dGTP is clearly visible in the s-site (Figure 8 - figure supplement 1d) and for ATP in the active site (Figure 8 - figure supplement 2). Again, the two monomers can be divided into an active one and an inactive one. The conformations of loop 2 are once again asymmetric (Figure 8 - figure supplement 4). The conformation in the active monomer is well-defined, while loop 2 in the inactive monomer is more flexible than in the ATP-dTTP-GTP complex, but still compatible with the hypothesis that the asymmetric arrangement of loop 2 is responsible for disorder of the NxN-GRD-substrate network in the inactive monomer.

Cryo-EM structure of the dATP-bound dimer

In micrographs of PcNrdD samples in the presence of 0.5 mM dATP, we observed an approximately 1:1 mixture of dimeric and tetrameric particles, consistent with results from GEMMA. To resolve whether dATP inhibition was linked to oligomerisation, we resolved the structures of each of these oligomers separately. We made a reconstruction of the dATP-bound PcNrdD dimer from 1 009 021 particles to 2.7 Å resolution that was further improved by DeepEMhancer postprocessing. The core domain is very similar to that of the ATP-CTP complex, with an rmsd in Cα positions of 1.2 Å for 1089 Cα atoms considering each dimer as a whole (Table 1; Table 1 - table supplement 4). dATP is clearly visible in the specificity site (Figure 8 - figure supplement 1b).

Strikingly, in contrast to the ATP complexes, the entire GRD is disordered in both monomers of the dATP-bound dimer. No structure is visible after residue 676 at the end of the last strand of the barrel. This suggests that inhibition by binding of dATP to the ATP-cone is coupled to disordering of the C-terminal domain. The NxN flap that covers the top of the substrate in the ATP/CTP-bound form is also disordered, as are the ATP-cones themselves.

Cryo-EM structure of the dATP-bound tetramer

From the same micrographs that revealed the dATP-bound dimers we also resolved dATP-bound tetramers. We obtained a map from cryoSPARC at 2.8 Å, which was again improved with DeepEMhancer (Fig. 9a). Two dATP molecules are bound to each ATP-cone and one at each of the specificity sites, giving a total of 12 visible dATPs per tetramer. The tetramer reveals an arrangement of two dimers (chains A/B and C/D respectively) in which the ATP-cones are all well-ordered and one ATP-cone from each pair mediates interactions between the dimers (Fig. 9b). The ATP-cones have an almost domain-swapped arrangement within each dimer relative to the core domains. This oligomeric arrangement has not previously been observed within the RNR family.

Structure of the dATP-bound PcNrdD tetramer.

a) cryo-EM reconstruction coloured by chain: dimer 1 light blue and pale blue; dimer 2 light green and pale green. The four ATP-cones are at the bottom of the figure. b) View rotated by 90° around a horizontal axis relative to a) and thus viewed from the bottom, showing the ATP-cones. c-d) Cartoon representation of the tetramer from the same angles as panels a) and b) respectively and with the same chain colouring. The two-fold symmetry axis relating the two dimers of the tetramer is shown as a yellow line. The 12 dATP molecules in the tetramer are shown as sticks. The dimer axis of the left-hand dimer’s core domains is shown as a dark blue line and the dimer axis of the ATP-cone pair as a black line. c) Details of the interaction of dATP molecules at the allosteric activity site in the ATP-cone. d) Close-up of the interaction area between the roof of the ATP-cone of one dimer and the core domain of the other dimer (marked with a black box in panel d). The most significant residues are labelled.

dATP binds to the specificity site as in the dimeric dATP-bound form (Figure 8 - figure supplement 1c). The two dATP molecules bound to each ATP-cone do so very similarly to the dATP-loaded ATP-cones of the RNR class I active site subunit from PaNrdA (PDB ID 5IM3) and the class I radical generating subunit from LbNrdB (5OLK) to which the PcNrdD ATP-cones have 29% and 33% sequence identity respectively (Figure 9 – figure supplement 1). The DALI server reveals an rmsd in Cα positions of 1.9 Å and 1.7 Å respectively. The next most similar ATP-cone is that of the RNR transcriptional regulator NrdR from Streptomyces coelicolor in its dodecameric form (Rozman Grinberg et al., 2022) binding two ATP molecules, with 82 aligned residues, rmsd 2.4 Å and sequence identity 17%. Structural similarity to other ATP-cones is listed in Figure 9 – figure supplement 1.

The two dATP molecules bind with their negatively charged triphosphate moieties oriented towards each other (Fig. 9c). No Mg2+ ion is observed between the triphosphate tails, but this may be an artefact of the cryo-EM method, as the PaNrdA and LbNrdB crystal structures suggest that this ion is necessary for charge neutralisation. As in the ATP-bound form, further charge compensation is achieved by extensive coordination by, or proximity of, basic side chains, e.g. Lys7, Arg8, Arg87 and Arg91. As previously observed in PaNrdA and LbNrdB, the adenosine base of the “inner” dATP (dATP1) fits into a pocket under the β-hairpin roof of the ATP-cone and the base of the “outer” nucleotide (dATP2) is sandwiched between an aromatic residue (Tyr83) and the side of the first α-helix in the domain. No direct contacts are seen between the dATP molecules bound to the ATP-cone and residues in the core domain.

Figure 9 – figure supplement 2 shows the differences in binding modes of dATP in the dATP-bound tetramer and ATP in the ATP-CTP dimer. The high mobility of the ATP-cone in the latter precludes a detailed analysis. However, it is clear that the triphosphate tails have very different orientations. In the ATP complex the tail of ATP2 is oriented away from the C-terminal helix towards the base of ATP1, such that the β-phosphate moiety may even H-bond to the base of ATP1. The side chain of Arg87 fills the space vacated by the ψ-phosphate of ATP2. Instead of H-bonding to the ψ-phosphate of ATP2, Arg91 swings away and interacts with the equivalent moiety of ATP2 in the ATP-cone of the other chain. The structures suggest that the additional 2’-OH group in ATP may cause conformational changes in the ribose ring that propagate to the phosphate tail, but further analysis requires improved resolution.

As in PaNrdA and LbNrdB, the ATP-cones in the dATP-bound tetramer self-interact through a pair of α-helices at the C-terminal end of the ATP-cone. Remarkably, the pair of ATP-cones is highly asymmetrically positioned with respect to the 2-fold axis of its core dimer, being displaced as a rigid body towards the other dimer (Fig. 9b). Both pairs of ATP-cones are displaced in the same way, giving the complex overall C2 symmetry around an axis bisecting the pair of dimers. Consequently, the 2-fold axes of the ATP-cone and the core dimer are not aligned (Fig. 9b). The asymmetric arrangement is associated with two distinct conformations of the linker 93-103 between the fourth helix of the ATP-cone (Figure 9 – figure supplement 3a,b) and the core domain. In chains A and C, a small kink leads from helix 4 into a short helical segment that packs antiparallel to helix 122-142 in the core domain. In contrast, in chains B and D, the short helix is parallel to helix 122-142 and is joined to helix 4 at a 45° angle. The rest of the polypeptide between residues 103-121, including the NxN flap, is disordered in both linkers.

Interestingly, all contacts between the two dimers of the tetramer are mediated by interactions between one of the ATP-cones in one dimer (B or D) and one of the core domains of the other dimer (C or A respectively). Figure 9d shows the interactions of the roof of the ATP-cone with two outer helices from the C-terminal half of the barrel domain (residues 458-485 and 550-569, preceding the sixth and seventh strands). The interactions are mostly hydrophobic but are reinforced by several H-bonds, e.g. between Gln3D-Gln458A, Ser53D–Gln458A, Arg11D-Asp468A, the main chain amide of Ile12D and Tyr557A. The amount of buried surface area is small: 717 Å2, or around 1% of the total surface area of each dimer. For comparison, the monomer-monomer interactions within each dimer bury 11.7 % of the total area. The residues on the core domain of PcNrdD involved in interactions with the ATP-cone of the other dimer do not show high sequence conservation, even when the alignments are restricted to the most similar sequences.

Cryo-EM structures of PcNrdD-dATP complexes produced in the presence of CTP

To investigate whether substrate had any allosteric effect on the conformations of the ATP-cone, flap and GRD, we determined separate cryo-EM structures of tetrameric and dimeric PcNrdD, from the same grid, in the presence of 0.5 mM dATP and 0.5 mM CTP, achieving a resolution of 2.6 Å from 1 105 348 particles before post-processing for the tetramer and 2.6 Å from 1 132 695 particles for the dimer (Table 1; Table 1 - table supplement 5). Both forms were almost identical to those seen in the presence of dATP alone, and no CTP could be seen bound to the active site (Figure 6 - figure supplement 2). This strongly suggests that substrate binding is dependent on ordering of the flap region by binding of ATP to the ATP-cone.

Discussion

Ribonucleotide reductase is an essential enzyme for all free-living organisms. Its complex regulation at the levels of transcription, overall activity and substrate specificity makes it quite unique in Nature (Mathews, 2016, 2018). A fundamental understanding of this regulation could e.g. pave the way for better antibiotics. The ATP-cone is a small, genetically mobile domain frequently found at the N-terminus of RNR catalytic subunits of all major classes, as well as in the bacterial RNR-specific transcription factor NrdR (Rozman Grinberg et al., 2022). Its role in allosteric regulation of class I aerobic RNRs has been extensively studied. The ATP-cone mediates the formation of a surprising variety of different oligomeric complexes that nevertheless have in common that they block the formation of an (observed or inferred) active complex between the catalytic and radical generating subunits that permits proton-coupled electron transfer on each catalytic cycle (Martínez-Carranza et al., 2020).

Despite their catalytic subunits being evolutionarily related to those of class I, anaerobic class III RNRs have a fundamentally different activation mechanism that is likely common to all glycyl radical enzymes (Backman et al., 2017; Lundahl et al, 2022). Once generated on the catalytic subunit by encounter of the reductase with its radical SAM family activase, the glycyl radical can catalyse hundreds of cycles of turnover before being exhausted (Torrents et al., 2001). Allosteric inhibition could thus in principle proceed by two mechanisms: by blocking the initial encounter of the reductase and activase, or by preventing subsequent transfer of the radical to the substrate. The first of these hypotheses is intuitively more similar to the steric blocking mechanisms encountered in class I RNRs; the latter is more challenging to understand, as the glycine and cysteine are buried in close proximity in the active site.

In this work we present the first biochemical, biophysical and structural characterisation of an anaerobic, class III RNR containing an ATP-cone, providing the first picture of the structural basis for allosteric activity regulation in this large family. We show that the ATP-cone can bind two molecules of ATP or dATP, but not both simultaneously. As expected, the presence of an ATP-cone confers inhibition of enzyme activity at high dATP concentrations. GEMMA experiments show that PcNrdD exists in a mixture of dimeric and tetrameric states under most experimental conditions studied, with the equilibrium shifted towards dimers in the presence of ATP. Addition of inhibiting concentrations of dATP shifts the equilibrium towards tetramers. The equilibria are not affected by substrate or specificity effector. EPR spectroscopy clearly shows that the glycyl radical is formed even in the presence of dATP. Together, these results show that dATP inhibition in class III RNRs does not occur by blocking the initial encounter of reductase and activase. Instead, our nucleotide binding results show that dATP inhibition prevents binding of substrates.

The seven cryo-EM structures of PcNrdD in a variety of nucleotide complexes presented here suggest a mechanism in which binding of ATP or dATP modulates activity by affecting the degree of order of a tightly knit network of structural elements including the critical glycyl radical domain in the C-terminal region of the enzyme. In the presence of ATP and substrate CTP, the GRD is fully ordered in one monomer and Gly711 is found at its expected position proximal to Cys416, to which it will deliver its radical to initiate each catalytic cycle and accept it back at the end of the cycle. The CTP substrate is bound in this monomer and a conserved but previously unobserved flap containing two Asn residues (NxN flap) forms over the top of the substrate, making H-bonding interactions with it. Loop 2, which mediates between the substrate specificity site and the active site, is fully ordered. In contrast, in the other monomer, no substrate is bound, loop 2 is disordered and the flap and C-terminus are less well-ordered. At the same time, the ATP-cones themselves are relatively flexible. This is not unexpected for this type of ATP-cone, which binds two nucleotides and was first observed in the class I RNR PaNrdA (Johansson et al., 2016). SAXS studies showed that the ATP-cone in PaNrdA is flexible in the presence of ATP, while in its dATP-bound form it forms a symmetric interaction with an ATP-cone on another dimer, causing the enzyme to tetramerise. A similar behaviour was seen even when this type of ATP-cone was evolutionarily “transplanted” to the N-terminus of the radical-generating subunit, as seen in LbNrdB (Rozman Grinberg et al., 2018a).

By careful isolation of a subset of particles, we were able to characterise an unusual dimer of ATP-cones in the ATP-CTP complex, where four ATP molecules bind with all their triphosphate tails pointing towards each other. Such an orientation is unprecedented for RNRs, but such tail-to-tail orientations, compensated by extensive positive charges are found in e.g. adenylate kinases (Berry, 1994). The relevance of this arrangement remains to be elucidated. Interestingly, when dTTP was added to the s-site in the ATP-dTTP-CTP complex, the ATP-cones became much more disordered and could only be observed in one monomer of the dimer as diffuse density. This behaviour was reproduced with dGTP in the s-site. Both complexes with s-site effector are functionally more relevant forms than the ATP-CTP complex, as in the cell, a form with empty s-site is unlikely to exist. The partitioning of monomers into active and inactive is amplified by the presence of s-site effector (dTTP or dGTP) in both complexes with ATP. Only one of the active sites is occupied by substrate, with ordered NxN flap and C-terminal GRD. This is correlated with two distinct conformations of loop 2 in the two monomers. In the inactive monomer, the NxN flap and GRD are completely disordered as loop 2 obviates their ordering and concomitantly substrate binding. These results point to only one of the PcNrdD monomers being active at a given time and also to a significant cross-talk between the a-site, s-site and active site. This is corroborated by the finding that an s-site effector is necessary for retention of ATP by the ATP-cone when PcNrdD is first incubated with these nucleotides then purified without them.

In the presence of inhibitory concentrations of dATP we were able to isolate structures of both dimers and tetramers from the same cryo-EM sample, in agreement with the GEMMA results that show a shift towards the tetrameric form. In both forms, the C-terminal GRD is completely disordered in all monomers. In the tetramers, pairs of ATP-cones asymmetrically disposed relative to the 2-fold axes of their respective PcNrdD dimers associate through the “roof” of the “inner” dATP binding site to the core domain of the other dimer. While this allowed us to build complete models for the ATP-cones and analyse dATP binding, the role of tetramerisation in the inhibition mechanism is unclear. The tetrameric structures appear to be facilitated by a dATP-induced “stiffening” of the pair of ATP-cones, similar to what was seen in PaNrdA and LbNrdB, and their interaction with a second dimer of PcNrdD, but the interaction area is small (1% of the total surface area) and the residues involved are not highly conserved in NrdDs, even from the most similar sequences. Furthermore, the GRD is disordered in all complexes with dATP in the a-site, whether dimeric or tetrameric. Since both the ATP-cone and the GRD are disordered in the dATP-bound dimers, the exact molecular mechanism by which dATP induces disorder of the GRD remains somewhat elusive.

Taken together, our results show a clear correlation between ATP or dATP binding to the ATP-cone and activity or inhibition respectively. The order-disorder transitions involve a previously uncharacterised but highly conserved structural element, the NxN flap, that folds over the substrate and furthermore acts as a structural bridge to the C-terminal GRD. Disordering of these elements prevents substrate binding and transfer of the radical from the GRD to the substrate. Interestingly, a similar flap is formed over the top of the active site in the E. coli active complex, but it is formed by part of the C-terminal region of the separate radical generating subunit (Kang, 2020). Thus, a locking of the substrate in the active site appears to be a conserved feature of active class I and class III RNRs. However, in class I RNR there is no tunnel allowing substrate access to the active site in the locked state. This is consistent with the fact that the class I subunits dissociate on each catalytic cycle, while in class III the glycyl radical can catalyse many substrate reductions before having to be regenerated. It remains enigmatic exactly how the tiny chemical difference between dATP and ATP (the 2’-OH group) results in such major conformational changes, as in the ATP-CTP complex, where the ATP-cones are visible, the a-site in the ATP-cone is separated from the NxN flap by at least 30 Å and in the most biologically relevant complexes with additional s-site effector, the ATP-cone is very disordered. Nevertheless, the present results give first insights into allosteric activity regulation in anaerobic RNRs and add yet another aspect to the surprisingly wide range of allosteric conformational changes that can be mediated by the binding of the two highly similar nucleotides ATP and dATP to a small, evolutionarily mobile protein domain.

Materials and Methods

Cloning of nrdD and nrdG from P. copri

The genes encoding the reductase (NrdD) and activase (NrdG) proteins from P. copri were synthesised by GenScript with codon optimization for E. coli and subcloned into the pBG102 plasmid (pET27 derivative) (Center for Structural Biology, Vanderbilt University) between the BamHI and EcoRI restriction sites to produce His6-SUMO-NrdD and His6-SUMO-NrdG protein constructs.

Overexpression and purification of PcNrdD and PcNrdG

Plasmids containing the nrdD gene or the nrdG gene were transformed into E. coli BL21 (DE3) star and E. coli BL21 (DE3) competent cells, respectively. Cells (6 L) were grown at 37 °C in Luria Broth supplemented with kanamycin (50 µg/mL) to an A600 of 1.2. Protein expression was then induced with 0.5 mM IPTG and incubation was extended overnight at 20 °C. After centrifugation, pellets were suspended in 60 mL of lysis buffer 1a (50 mM Tris-HCl pH 8, 500 mM KCl, 0.5 mM TCEP) for PcNrdD and lysis buffer 2a (50 mM Tris-HCl pH 8, 500 mM NaCl, 0.5 mM TCEP) for PcNrdG, supplemented with lysozyme (0.5 mg/mL), and disrupted by sonication. Cell debris was removed by ultracentrifugation at 210 000 x g for 1 h at 4 °C. The supernatant was then loaded on an immobilised metal affinity Ni-NTA column (HisTrap 5 mL; Cytiva) equilibrated with buffer 1b (50 mM Tris-HCl, pH 8, 300 mM KCl) or buffer 2b (50 mM Tris-HCl, pH 8, 300 mM NaCl), supplemented with 50 mM imidazole as appropriate. The column was washed extensively with the corresponding buffer containing 50 mM imidazole and the His6-SUMO tagged proteins were eluted using 300 mM imidazole. The proteins were collected and dialyzed overnight at 4 °C against buffer 1b or 2b supplemented with 1 mM 1,4-dithiothreitol (DTT) in the presence of the PreScission Protease (150 µM) to cleave the affinity-solubility tag. The GST–3C-protease (PreScission) was expressed using pGEX-2T recombinant plasmids. After induction at 25 °C with 0.1 mM IPTG for 20 h, the PreScission Protease was purified using glutathione–Sepharose chromatography. After the dialysis, the cleaved PcNrdD and PcNrdG proteins were centrifuged at 4 °C for 10 min and loaded onto a HisTrap column equilibrated with buffer 1b or 2b, supplemented with 50 mM imidazole and the flow-through containing the cleaved protein was collected. Following cleavage of the His6-SUMO tag, the proteins incorporated a non-native N-terminal Gly-Pro-Gly-Ser sequence. The purified preparations are called as-purified PcNrdD and as-purified PcNrdG, respectively.

A fraction of as-purified PcNrdD was precipitated by addition of HCl, then centrifuged at 15 000 x g for 5 min. UV-visible spectra of the supernatant were recorded before and after protein precipitation to estimate the amount of nucleotide contamination. PcNrdD preparations with ≤5% nucleotide contamination are referred to as apo-PcNrdD.

Fractions containing apo-PcNrdD and as-purified PcNrdG were concentrated and loaded on a gel filtration column (Hiload 16/60 Superdex S200; Cytiva) in 25 mM Tris-HCl, pH 8, 250 mM NaCl, and 5 mM DTT. The purified protein was concentrated to 20 mg/mL with an Amicon ultrafiltration device (100 kDa cutoff for NrdD and 10 kDa cutoff for NrdG; Millipore), frozen in liquid nitrogen, and stored at −80 °C.

[Fe-S] cluster reconstitution and purification of holo-PcNrdG

The reconstitution of the [4Fe-4S] cluster and purification of PcNrdG containing an iron-sulfur cluster (holo-PcNrdG) were performed under strict anaerobic conditions in an Mbraun glove box kept at 18°C and containing less than 0.5 ppm O2. PcNrdG was treated with 5 mM DTT for 10 min then incubated for 3 h with a fivefold molar excess of ferrous ammonium sulphate and L-cysteine in the presence of 2 μM E. coli cysteine desulfurase CsdA. The holo-PcNrdG was then loaded onto a Superdex 200 10/300 gel filtration column (Cytiva) equilibrated in 25 mM Tris-HCl, 250 mM NaCl, and 5 mM DTT. The peak containing the soluble holo-PcNrdG was then concentrated to 10 mg/mL on a Vivaspin concentrator (10 kDa cutoff) (Figure 1 - figure supplement 2).

In vitro enzymatic assay and HPLC analysis

Activity assays were performed under strict anaerobic conditions inside a glove box in 25 mM Tris pH 8, 50 mM KCl, 10 mM DTT in a volume of 100 μL. A standard pre-reaction mixture contained 1 μM apo-PcNrdD, 2.5 μM holo-PcNrdG, 500 µM S-adenosyl-methionine (SAM), 5 mM MgCl2, 1 mM GTP (or 1 mM CTP), and different allosteric effectors dTTP, ATP or dATP. The pre-reaction mixture was incubated for 5 min at 37 °C before adding simultaneously 10 mM sodium formate and sodium dithionite (12.5 molar excess). CTP was used as a substrate, ATP and dATP were titrated and when dTTP was used as an s-site effector, GTP was used as substrate.

The reaction was incubated for 10 min at 37 °C and stopped by the addition of 2.5 μL of 3 M formic acid. Product formation was analysed by HPLC using an Agilent ZORBAX RR StableBond (C18, 4.6 x 150 mm, 3.5 µm pore size) equilibrated with buffer A (10% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Samples of 10 μL were injected and eluted at 0.5 mL/min with a step isocratic flow of 40%-100% buffer B (30% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Compound identification and product quantification were performed by external calibration and NrdD specific activity was determined.

From a direct plot of activity versus concentration of effector, the KL values for binding of effectors to the a-site, were calculated in KaleidaGraph using the equation:

and Ki for non-competitive dATP inhibition was calculated in KaleidaGraph using the equation:

Isothermal titration calorimetry experiments

ITC experiments were carried out at 20°C on a MicroCal PEAQ-ITC system (Malvern Instruments Ltd) in a buffer containing 25 mM HEPES, pH 7.5, 100 mM KCl, 10 mM MgCl2 and 0.5 mM TCEP with a stirring speed of 1000 rpm. An initial injection volume was 0.6 µl over a duration of 1.2 s. All subsequent injection volumes were 1.7–2.5 µl over 3.4–5 s with a spacing of 160 s between the injections. Data for the initial injection were not considered. PcNrdD was incubated with 1 mM dTTP, 1 mM GTP and 10 mM MgCl2 at room temperature during 5 min. For dATP binding analysis, the concentration of PcNrdD in the cell was 50 to 100 µM and dATP concentration in the syringe 0.6 to 1 mM. For ATP titration into PcNrdD, protein concentration in the cell was 200 µM and ATP concentration in the syringe was 0.6 to 1.5 mM. The data were analysed using the one set of sites model of the MicroCal PEAQ-ITC analysis software (Malvern Panalytical). Standard deviations in thermodynamic parameters, N and KD were estimated from the fits of three different titrations.

Microscale thermophoresis

Both binding of nucleotides to the ATP-cone in PcNrdD and binding of substrates CTP and GTP to PcNrdD were assessed using MST. PcNrdD was labelled using Monolith Protein Labeling Kit RED-NHS 2nd generation according to the manufacturer’s protocol. MST was performed using the Monolith NT.115 instrument (NanoTemper Technologies GmbH) at room temperature. Binding of GTP and CTP was assayed in MST buffer containing 40 mM Hepes pH 7, 50 mM KCl, 50 and 100 mM MgCl2, 5 mM DTT, 0.1% Tween-20, 1 mM dTTP (only for GTP binding) and either 5 mM ATP or 1 mM dATP. The 16 reaction tubes were prepared by the addition of 2 μl MST buffer (concentrated 5 times), 5 μl GTP or CTP of the desired concentration and 1 μl RED dye-labelled NrdD in a total volume of 10 μl. Final NrdD concentration in the binding reaction was 11 nM while the binding partner concentrations were between 70 mM and 2 μM for CTP and 50 mM and 1.5 μM for GTP. Binding of ATP and dATP to the ATP-cone was assayed in MST buffer containing 25 mM Hepes pH 7.5, 100 mM KCl, 10 mM MgCl2, 0.5 mM TCEP, 0.1% Tween-20, 1 mM dTTP and 5 mM GTP. A 16-step dilution series of the binding partners was prepared by adding 5 μl buffer to 15 tubes. ATP or dATP was added to the first tube to the final concentration of 4 mM and 5 μl was transferred to the second tube and mixed by pipetting (1:1 dilution series). To each tube of the dilution series 5 μl of PcNrdD was added to reach a final concentration of 11 nM. The samples were transferred to Monolith™ NT.115 Series Premium Capillaries (NanoTemper Technologies GmbH), which were scanned using the MST instrument (100% excitation power, medium MST power). Obtained MST data was analysed and fitted using the MO. Affinity Analysis v2.3 software (NanoTemper Technologies) with default parameters. KD and standard deviation for GTP and CTP binding to PcNrdD in the presence of ATP and for ATP and dATP binding to the ATP-cone were calculated using fits from at least three individual titrations. No binding was detected for CTP or GTP to PcNrdD in the presence of dATP.

Nucleotide binding to the PcNrdD ATP-cone

PcNrdD was incubated with nucleotides at room temperature for 30 minutes and then desalted on a NAP-5 column. The desalted protein was boiled for 10 min, centrifuged, and the supernatant was loaded on a Agilent ZORBAX RR StableBond (C18, 4.6 x 150 mm, 3.5 µm pore size) HPLC column to evaluate the amount of nucleotides retained by the protein. Samples of 100 μL were injected on the column equilibrated with buffer A (10% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide) and eluted at 1 mL/min with a gradient of 40%-100% buffer B (30% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Compound identification and product quantification were performed by comparison with injected standards. Three different sets of experiments were performed, each set consisting of incubation with only ATP, only dATP, and a combination of ATP and dATP. In the first set of experiments 75 µM apo-NrdD was incubated with 1 mM ATP or 1 mM dATP, and when 1 mM each of ATP and dATP was used, dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer without nucleotides. In the second set of experiments 75 µM apo-NrdD in presence of 1 mM dTTP and 5 mM GTP was incubated with 3 mM ATP or 1 mM dATP, and when both nucleotides were combined dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer without nucleotides. In the third set of experiments 75 µM apo-NrdD in presence of 2 mM dTTP and 5 mM GTP was incubated with 3 mM ATP or 1 mM dATP, and when both nucleotides were combined dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer containing 0.1 mM dTTP and 1 mM GTP. It was possible to separate all phosphorylation levels of adenosine and deoxyadenosine nucleotides in the first set of experiments, but not in the second and third sets due to overlapping peaks of the s-site and substrate nucleotides (Figure 2 - figure supplement 1).

Gas-phase electrophoretic mobility molecular analysis

The GEMMA instrumental setup and general procedures were as described previously (Kaufman et al., 1996; Rofougaran et al, 2008). Apo-PcNrdD was equilibrated into a buffer containing 100 mM ammonium acetate pH 7.8, then applied onto a Sephadex G-25 chromatography column. In addition, 5 mM DTT was added to the PcNrdD protein solution to increase the protein stability. Prior to analysis, the protein sample was diluted to 2 µM in a buffer containing 100 mM ammonium acetate, pH 7.8, 0.005% (v/v) Tween 20, magnesium acetate (equimolar to the total nucleotide concentration) and the corresponding nucleotides (where indicated). The protein samples were then incubated for 2 min at room temperature, centrifuged and applied to the GEMMA instrument for data collection. The experiments were performed with a flow rate driven by 2 psi to minimise noise signals that may appear with elevated nucleotide or protein concentrations at the manufacturer’s recommended flow rate driven by 3.7 psi.

Glycyl radical characterization by EPR

EPR samples were handled under strictly anaerobic conditions inside a glovebox in 25 mM Tris-HCl pH 8, 50 mM KCl, 5 mM DTT in a final volume of 100 μL. A standard pre-reaction mixture containing 75 μM holo-PcNrdG, 1 mM S-adenosyl-methionine, was treated with a 12.5 excess of dithionite and was incubated for 20 min at 37 °C. The pre-mixture was added to a mixture of 50 µM apo-PcNrdD, 10 mM sodium formate, supplemented or not with 1.5 mM ATP or 1 mM dATP and in the presence of 1 mM CTP. The final reaction was incubated for 20 min at 37 °C and 100 µL sample was transferred into a standard EPR tube (Wilmad LabGlass 707-SQ-250M) and stored in liquid nitrogen before recording the EPR spectrum.

X-band CW-EPR spectra of PcNrdD were recorded on a Bruker ELEXYS-E500 spectrometer operating at microwave frequency of 9.38 GHz, equipped with a SuperX EPR049 microwave bridge and a cylindrical TE011 ER 4122SHQE cavity in connection with an Oxford Instruments continuous flow cryostat. Measuring temperature at 30 K was achieved using liquid helium flow through an ITC 503 temperature controller (Oxford Instruments). For achieving non-saturating conditions, a microwave power of 10 µW was applied for all measurements. Other EPR settings were: modulation frequency of 100 kHz; modulation amplitude of 8 G; and accumulation of 16 scans for an optimal S/N ratio. Double integrated spectra were used for intensity analysis.

Cryo-EM sample preparation and data acquisition

The proteins for cryo-EM analysis were all prepared in the same manner. Apo-PcNrdD was incubated with 1 mM NTP or dNTP at room temperature for 5 min, then centrifuged before injection onto a Superdex S200 10/300 column pre-equilibrated with a buffer containing 25 mM HEPES-NaOH pH 7.5, 100 mM KCl, 0.5 mM TCEP, 10 mM MgCl2 and supplemented with varying concentrations of 0.5 mM NTP or 0.2 mM dNTP. The relevant peak was collected, then diluted at 0.5 mg/mL and the corresponding nucleotides were added to different final concentrations (0.5-5 mM NTP or 0.5-1 mM dNTP) (Table 1).

The grids were glow-discharged for 60 s at 20 mA using a GloQube (Quorum) instrument, then prepared in the following manner: 3 µL of sample at 0.5 mg/ml were applied on Quantifoil 1.2, 1.3 300 mesh Cu grids or Quantifoil 2.1, 300 mesh Au grids and plunge-frozen in liquid nitrogen-cooled liquid ethane using a Vitrobot Mark IV (Thermo Fisher Scientific) with a blot force of 1 and followed by 5 s blot time, at 4 °C and 100% humidity. The grids used for each sample are specified in Table 1.

Grids were clipped and loaded into a 300 kV Titan Krios G2 microscope (Thermo Fisher Scientific, EPU 2.8.1 software) equipped with a Gatan BioQuantum energy filter and a K3 Summit direct electron detector (AMETEK). Grids were screened for quality based on particle distribution and density, and images from the best grid were recorded. Micrographs were recorded at a nominal magnification of 105,000. Details of the other data collection parameters used for each sample are given in Table 1.

Cryo-EM data processing

Data processing was performed in cryoSPARC (Punjani et al, 2017). The first steps in processing of all datasets were patch motion correction and patch CTF estimation, followed by curation of exposures based on ice thickness, poor defocus estimation, etc. Other steps are as detailed below. Resolution estimation in cryoSPARC was done using a gold standard Fourier shell correlation value of 0.143 (Rosenthal & Henderson, 2003). Map post-processing was done using DeepEMhancer (Sanchez-Garcia et al., 2021) and post-processing resolution estimation was performed using DeepRes (Ramirez-Aportela et al, 2019). Models were placed in the maps using either phenix.dock_in_map (Liebschner et al, 2019) or Molrep (Vagin & Teplyakov, 1997) in the CCP-EM package (Burnley et al, 2017). Model building was done by alternating rounds of manual building in Coot (Emsley et al, 2010) with real space refinement in phenix.refine (Liebschner et al., 2019). Secondary structure restraints were used throughout. The inclusion of riding hydrogen atoms reduced the number of bad contacts in all models. An AlphaFold2 (Jumper et al., 2021) model of PcNrdD was used in the later stages of building the first model to resolve some ambiguous loops. All final models have good correlations between map and model and good stereochemical properties (Table 1). All structures have been deposited in the Protein Databank and the corresponding volumes deposited in the Electron Microscopy Data Bank with the accession numbers listed in Table 1.

PcNrdD-ATP-CTP complex

Data processing

A total of 17 033 movies were used after curation. A low-pass filtered volume was made in Chimera (Pettersen et al, 2021) from the crystal structure of NrdD from T. maritima (TmNrdD, PDB ID 4COI). From this volume, a set of templates was created and used for template-based picking with a diameter of 100 Å. Due to the large number of micrographs, they were split into two sets. The largest set (16 219 micrographs) was analysed first. About 8.6 million particles were extracted with a 350-pixel box size and classified into 60 2D classes. Eight representative classes containing different orientations (∼3.3 million particles) were used to generate four ab-initio models without symmetry (C1) and particles were subjected to three-dimensional (3D) heterogeneous refinement using four classes. The most populated model containing ∼1 million particles was subjected to homogeneous and non-uniform refinement, which gave a map with an overall FSC resolution of 2.94 Å. A further 435 695 particles were then extracted from the remaining 810 micrographs. These were subjected to 2D classification, and the resulting classes were merged with the larger set, giving 1 199 575 particles. The combined particles were subjected to homogeneous, non-uniform refinement and local refinement with a predefined mask, which gave a map at 2.87 Å (Table 1 - table supplement 1).

Generation of map with better ATP-cone density

The particles were further classified into 50 2D classes, from which 23 classes having 1 040 266 particles were selected. Four ab-initio models were generated and 570 730 particles with slightly better ATP-cone density were refined without symmetry. The particles were then subjected to 3D classification using ten classes of ∼57 000 particles each. The classes having slightly better ATP-cone density were selected for further processing and final refinement without symmetry gave a map with an overall FSC resolution of 3.17 Å from 291 231 particles, which was post-processed using DeepEMhancer (Table 1 - table supplement 1).

Model building and refinement

A partially complete model for one dimer of the dATP complex (see below) was placed in the map using phenix.dock_in_map. The ATP-cones were built based on the ones from the dATP-only tetramers (see below).

PcNrdD-dATP complexes

Tetramers

Data processing for first part

A set of 11 161 movies was used after curation, template-based picking based on the structure of TmNrdD and particle extraction using a box size of 448 pixels gave 5 331 420 particles. Seven 2D classes having 1 171 839 particles were used to make ab-initio models, followed by heterogeneous refinement. The best class having 461 020 particles and volume was used for homogeneous refinement, which gave a map at 3.3 Å. Non-uniform refinement followed by local refinement with a pre-determined mask gave a map at 3.1 Å with no imposed symmetry.

Data processing for second part

A set of 16 667 movies was used after curation. Particles were picked and extracted using the same box size as in the first part, which gave 7 889 234 particles. Ten 2D classes having 1 881 057 particles were used to make two ab-initio models, followed by heterogeneous refinement. The best class with 1 083 657 particles was selected and subjected to homogeneous refinement followed by non-uniform refinement and a local new refinement with a pre-determined mask, which gave a map at 2.9 Å with no imposed symmetry.

Merging of datasets and CTF refinement

The best particles from the first dataset were classified into 80 2D classes from which 36 having 399 523 particles were selected. The particles from the second dataset resulting in the best volume were classified into 100 2D classes, of which 31 having 949 610 particles were selected. These particle sets were merged (giving 1 349 133 particles) and refined against the best 3D volume which gave a map at 2.78 Å. Iterative refinement and NU refinement jobs, ultimately using C2 symmetry, and local CTF refinement with a defocus search range of +/- 2000, gave a map at 2.8 Å with C2 symmetry that was post-processed using DeepEMhancer (Table 1 - table supplement 4).

Model building and refinement

Model building was initially done for one dimer by fitting a homology model generated using SwissModel with T4NrdD as a template to the map using phenix.dock_in_map. Similar homology models for the four ATP-cones were placed by hand into the density followed by rounds of real space refinement. After almost completely rebuilding one dimer, the second dimer was placed into the density and model building and refinement were continued.

Dimers

The first part of the data processing was shared with the PcNrdD-dATP tetramers. In the same micrographs there were a number of 2D class averages that represented dimers. Eight such class averages having 1 916 387 particles were used to make two ab-initio models. This was followed by rounds of heterogeneous refinement from where the best model having 1 009 021 particles was refined with no symmetry to give a map at 2.8 Å. A final non-uniform refinement with C1 symmetry gave a map at 2.6 Å that was post-processed using DeepEMhancer (Table 1 - table supplement 4).

PcNrdD-dATP-CTP complexes

Tetramers

Blob picking was carried out from 11 780 curated micrographs with a maximum diameter of 200 Å and a minimum diameter of 90. In total 4 646 479 particles were then extracted with a box size of 448 pixels. This was followed by 2D classification, ab-initio model generation and heterogeneous refinements. The best class with 1 105 348 particles was used for non-uniform refinement with C2 symmetry, which gave a final map at 2.59 Å resolution that was post-processed using DeepEMhancer (Table 1 - table supplement 5).

Dimers

The initial data processing steps were the same as above. Template based picking from 8 796 curated micrographs using a low-pass filtered volume of the dATP-only tetramer gave 6 836 668 particles after extraction with a box size of 300 pixels. This was followed by 2D classifications, selection of classes representing dimers, ab-initio model generations and refinement of the best class. The final round of non-uniform refinement with 1 132 695 particles with C1 symmetry gave a map of 2.53 Å resolution. The estimated median resolution after post-processing was 2.1 Å (Table 1 - table supplement 5). Model building used the coordinates of the dATP-only dimer as a starting model.

PcNrdD-ATP-dTTP-GTP complex

Data processing

A total of 12 501 movies were processed and 6 105 movies were used after curation. Template-based picking was used with a diameter of 150 pixels. About 3.9 million particles were extracted from 5 952 movies with a 400 pixel box size and used for two-dimensional (2D) classification into 100 classes. Eleven representative 2D classes containing different orientations (∼2.4 million particles) were used to generate an ab-initio model and particles were subjected to three-dimensional (3D) heterogeneous refinement using ten classes, resulting in the most populated model containing 437 866 particles. This model was subjected to homogeneous and non-uniform refinement, which gave a map with an overall FSC resolution of 2.47 Å. The particles were then subjected to 3D classification and the resulting classes were merged for further processing and final refinement without symmetry (C1) gave a map with an overall FSC resolution of 2.4 Å that was post-processed using DeepEMhancer (Table 1 - table supplement 2).

Model building and refinement

A complete model for the dimer of the ATP-CTP structure (see above) was placed in the map using phenix.dock_in_map. The ATP-cones were removed as they were not visible in the reconstruction. After model building and refinement, the final map-to-model correlation value was X. Model quality statistics are presented in Table 1.

PcNrdD-ATP-dGTP complex

Data processing

Template-based picking was carried out on 14 373 curated micrographs and 4 522 028 particles were extracted with a box size of 350 pixels. This was followed by multiple rounds of 2D classifications, ab-initio model generation and refinements. C1 symmetry was applied on the best volume which gave a map with 589 345 particles having a final resolution of 2.58 Å. Model building and refinement used the PcNrdD-ATP/CTP complex as a starting model. (Table 1 - table supplement 3)

Bioinformatics

To construct an HMM logo of NrdD sequences, a representative selection of NrdD sequences from all domains of life including viruses were collected and aligned with Clustal Omega (Sievers et al, 2011). Subsequently an HMM model was built with hmmbuild from the HMMER suite (Eddy, 2011) and a logo was created using the Skylign web service using the “Information Content - Above Background” option (Wheeler et al, 2014). Conserved parts of the logo were extracted and displayed here.

eLife Assessment

This study advances our understanding of the allosteric regulation of anaerobic ribonucleotide reductases (RNRs) by nucleotides, providing valuable new structural insight into class III RNRs containing ATP cones. The cryo-EM structural characterization of the system is solid, but other aspects of the manuscript, which are incomplete, could be improved by including additional functional characterization and more evidence for the proposed mechanism of inhibition by dATP. The work will be of interest to biochemists and structural biologists working on ribonucleotide reductases and other allosterically regulated enzymes.

Reviewer #1 (Public Review):

The goal of this study is to understand the allosteric mechanism of overall activity regulation in an anaerobic ribonucleotide reductase (RNR) that contains an ATP-cone domain. Through cryo-EM structural analysis of various nucleotide-bound states of the RNR, the mechanism of dATP inhibition is found to involve order-disorder transitions in the active site. These effects appear to prevent substrate binding and a radical transfer needed to initiate the reaction.

Strengths of the manuscript include the comprehensive nature of the work - including numerous structures of different forms of the RNR and detailed characterization of enzyme activity to establish the parameters of dATP inhibition. The manuscript could be improved, however, by performing additional experiments to establish that the mechanism of inhibition can be observed in other contexts and it is not an artifact of the structural approach. Additionally, some of the presentations of biochemical data could be improved to comply with standard best practices.

The work is impactful because it reports initial observations about a potentially new mode of allosteric inhibition in this enzyme class. It also sets the stage for future work to understand the molecular basis for this phenomenon in more detail.

General comments:

1. It would be ideal to perform an additional experiment of some type to confirm the order-disorder phenomena observed in the cryo-EM structures to rule out the possibility that it is an artifact of the structure determination approach. Circular dichroism might be a possibility?

2. Does the disordering phenomenon of one subunit in the ATP-bound structures have any significance - could it be related to half-of-sites activity? Does this RNR exhibit half-of-sites activity?

3. Does the disordering of the GRD with dATP bound have any long-term impact on the stability of the Gly radical? I realize that the authors tested the ability to form the Gly radical in the presence of dATP in Fig. 4 of the manuscript. But it looks like they only analyzed the samples after 20 min of incubation. Were longer time points analyzed?

4. Did the authors establish whether the effect of dATP inhibition on substrate binding is reversible? If dATP is removed, can substrates rebind?

5. In some figures (Fig. 6e, for example), the cryo-EM density map for the nucleotide component of the model is not continuous over the entire molecule. Can the authors comment on the significance of this phenomenon? Were the ligands validated in any way to ensure that the assignments were made correctly?

Reviewer #2 (Public Review):

This manuscript describes the functional and structural characterization of an anaerobic (Class III) ribonucleotide reductase (RNR) with an ATP cone domain from Prevotella copri (PcNrdD). Most significantly, the cryo-EM structural characterization revealed the presence of a flap domain that connects the ATP cone domain and the active site and provides structural insights about how nucleotides and deoxynucleotides bind to this enzyme. The authors also demonstrated the catalytic functions and the oligomeric states. However, many of the biochemical characterizations are incomplete, and it is difficult to make mechanistic conclusions from the reported structures. The reported nucleotide-binding constants may not be accurate because of the design of the assays, which complicates the interpretation of the effects of ATP and dATP on PcNrdD oligomeric states. Importantly, statistical information was missing in most of the biochemical data. Also, while the authors concluded that the dATP binding makes the GRD flexible based on the absence of cryo-EM density for GRD in the dATP-bound PcNrdD, no other supports were provided. There was also a concern about the relevance of the proposed GRD flexibility and the stability of Gly radical. Overall, the manuscript provides structural insights about Class III RNR with ATP cone domain and how it binds ATP and dATP allosteric effectors. However, ambiguity remains about the molecular mechanism by which the dATP binding to the ATP cone domain inhibits the Class III RNR activity.

Strengths:

1. The manuscript reports the first near-atomic resolution of the structures of Class III RNR with ATP domain in complex with ATP and dATP. These structures revealed the NxN flap domain proposed to form an interaction network between the substrate, the linker to the ATP cone domain, the GRD, and loop 2 important for substrate specificity. The structures also provided insights into how ATP and dATP bind to the ATP cone domain of Class III RNR. Also, the structures suggested that the ATP cone domain is directly involved in the tetramer formation by forming an interaction with the core domain in the presence of dATP. These observations serve as an important basis for future study on the mechanism of Allosteric regulation of Class III RNR.

2. The authors used a wide range of methodologies including activity assays, nucleotide binding assays, oligomeric state determination, and cryo-EM structural characterization, which were impressive and necessary to understand the complex allosteric regulation of RNR.

3. The activity assays demonstrated the catalytic function of PcNrdD and its ability to be activated by ATP and low-concentration dATP and inhibited by high-concentration dATP.

4. ITC and MST were used to show the ability of PcNrdD to bind NTP and dATP.

5. GEMMA was used successfully to determine the oligomeric state of PcNrdD, which suggested that PcNrdD exists in dimeric and tetrameric forms, whose ratio is affected by ATP and/or dATP.

Weaknesses:

1. Activity assays.

The activity assays were performed under conditions that may not represent the nucleotide reduction activity. The authors initiated the Gly radical formation and nucleotide reduction simultaneously. The authors also showed that the amount of Gly radical formation was different in the presence of ATP vs dATP. Therefore, it is possible that the observed Vmax is affected by the amount of Gly radical. In fact, some of the data fit poorly into the kinetic model. Also, the number of biological and technical replicates was not described, and no statistical information was provided for the curve fitting.

2. Binding assays.

The interpretation of the binding assays is complicated by the fact that dATP binds both a- and s-sites and ATP binds a- and active sites. dATP may also bind the active site as the product. It is unknown if ATP binds s-site in PcNrdD. Despite this complexity, the binding assays were performed under the condition that all the binding sites were available. Therefore, it is not clear which event these assays are reporting.

3. Oligomeric states.

Due to the ambiguity in the kinetic parameters and the binding constants determined above, the effects of ATP and dATP on the oligomeric states are difficult to interpret. The concentrations of ATP used in these experiments (50 and 100 uM) were significantly lower than KL determined by the activity assays (780 uM), while it is close to the Kd values determined by ITC or MST (~25 uM). Since it is unclear what binding events ITC and MST are reporting, the data in Figure 3 does not provide support for the claimed effects of ATP binding. For the effects of dATP, the authors did not observe a significant difference in oligomeric states between 50 or 100 uM dATP alone vs 50 uM dATP and 100 uM CTP. The former condition has dATP ~ 2x higher than the Kd and KL (Figure 1b) and therefore could be considered as "inhibited". On the other hand, NrdD should be fully active under the latter condition. Therefore, these observations show no correlation between the oligomeric state and the catalytic activity.

4. Effects of dATP binding on GRD structure

One of the key conclusions of this manuscript is that dATP binding induces the dissociation of GRD from the active site. However, the structures did not provide an explanation for how the dATP binding affects the conformation of GRD or whether the dissociation of GRD is a direct consequence of dATP binding or it is due to the absence of nucleotide substrate. Also, Gly radical is unlikely to be stable when it is not protected from the bulk solvent. Therefore, it is unlikely that the GRD dissociates from the active site unless the inhibition by dATP is irreversible. Further evidence is needed to support the proposed mechanism of inhibition by dATP.

5. Functional support for the observed structures.

Evidence for connecting structural observations and mechanistic conclusions is largely missing. For example, the authors proposed that the interactions between the ATP cone domain and the core domain are responsible for tetramer formation. However, no biochemical evidence was provided to support this proposal. Similarly, the functional significance of the interaction through the NxN flap domain was not proved by mutagenesis experiments.

Reviewer #3 (Public Review):

The manuscript by Bimai et al describes a structural and functional characterization of an anaerobic ribonucleotide reductase (RNR) enzyme from the human microbe, P. copri. More specifically, the authors aimed to characterize the mechanism by how (d)ATP modulates nucleotide reduction in this anaerobic RNR, using a combination of enzyme kinetics, binding thermodynamics, and cryo-EM structural determination. One of the principal findings of this paper is the ordering of a NxN 'flap' in the presence of ATP that promotes RNR catalysis and the disordering of both this flap and the glycyl radical domain (GRD) when the inhibitory effector, dATP, binds. The latter is correlated with a loss of substrate binding, which is the likely mechanism for dATP inhibition. It is important to note that the GRD is remote (>30 Ang) from the binding site of the dATP molecule, suggesting long-range communication of the structural (dis)ordering. The authors also present evidence for a shift in oligomerization in the presence of dATP. The work does provide evidence for new insights/views into the subtle differences of nucleotide modulation (allostery) of RNR through long-range interactions.

The strengths of the work are the impressive, in-depth structural analysis of the various regulated forms of PcRNR by (d)ATP using cryo-EM. The authors present seven different models in total, with striking differences in oligomerization and (dis)ordering of select structural features, including the GRD that is integral to catalysis. The authors present several, complementary biochemical experiments (ITC, MST, EPR, kinetics) aimed at resolving the binding and regulatory mechanism of the enzyme by various nucleotides. The authors present a good breadth of the literature in which the focus of allosteric regulation of RNRs has been on the aerobic orthologues.

Given the resolution of some of the structures in the remote regions that appear to be of importance, the rigor of the work could have been improved by complementing this experimental studies with molecular dynamics (MD) simulations to reveal the dynamics of the GRD and loops/flaps at the active site. The biochemical data supporting the loss of substrate binding with dATP association is compelling, but the binding studies of the (d)ATP regulatory molecules are not; the authors noted less-than-unity binding stoichiometries for the effectors. Also, the work would benefit from additional support for oligomerization changes using an additional biochemical/biophysical approach.

Overall, the authors have mostly achieved their overall aims of the manuscript. With focused modifications, including additional control experiments, the manuscript should be a welcomed addition to the RNR field.

Author Response

The following is the authors’ response to the original reviews.

We thank the Reviewing Editor and two additional reviewers for the insightful input they gave us on the first version of our manuscript on allosteric activity regulation of the anaerobic ribonucleotide reductase from Prevotella copri. We have revised the manuscript in the light of the reviewers' comments. In particular, we have added additional experiments using hydrogen-deuterium exchange mass spectrometry (HDX-MS) to probe the accessibility and mobility of different parts of the protein structure in the apo-state and in the presence of dATP/CTP and ATP/CTP. The results strongly confirm the binding of nucleotides to the activity and specificity sites, as seen biochemically and structurally. In the question of mobility of the glycyl radical domain the HDX-MS experiments suggest an increased mobility in the presence of dATP, though the results are not as clear-cut as for the nucleotide binding. The HDX-MS analyses are complicated by the fact that they reflect all species in solution, which are evidently multiple for all states of PcNrdD. Finally, we have rephrased key parts of the results and discussion, and modified the title, to avoid any implication that we believe the glycyl radical domain becomes extensively disordered, rather that it becomes more mobile to the extent that it cannot be seen in the cryo-EM structures.

eLife assessment

This study advances our understanding of the allosteric regulation of anaerobic ribonucleotide reductases (RNRs) by nucleotides, providing valuable new structural insight into class III RNRs containing ATP cones. The cryo-EM structural characterization of the system is solid, but other aspects of the manuscript, which are incomplete, could be improved by including additional functional characterization and more evidence for the proposed mechanism of inhibition by dATP. The work will be of interest to biochemists and structural biologists working on ribonucleotide reductases and other allosterically regulated enzymes.

Public Reviews:

Reviewer #1 (Public Review):

The goal of this study is to understand the allosteric mechanism of overall activity regulation in an anaerobic ribonucleotide reductase (RNR) that contains an ATP-cone domain. Through cryo-EM structural analysis of various nucleotide-bound states of the RNR, the mechanism of dATP inhibition is found to involve order-disorder transitions in the active site. These effects appear to prevent substrate binding and a radical transfer needed to initiate the reaction.

Strengths of the manuscript include the comprehensive nature of the work - including numerous structures of different forms of the RNR and detailed characterization of enzyme activity to establish the parameters of dATP inhibition. The manuscript could be improved, however, by performing additional experiments to establish that the mechanism of inhibition can be observed in other contexts and it is not an artifact of the structural approach. Additionally, some of the presentations of biochemical data could be improved to comply with standard best practices.

The work is impactful because it reports initial observations about a potentially new mode of allosteric inhibition in this enzyme class. It also sets the stage for future work to understand the molecular basis for this phenomenon in more detail.

We thank the editor and reviewers for their positive evaluation of the potential impact of our work. We completely agree that hypotheses based on structural data require orthogonal experimental verification. However, the number and consistency of the cryo-EM structures speak in favour of the data being representative of conditions in solution. We feel that in particular cryo-EM data should be relatively free of artefacts, e.g. biased or incorrect relative domain orientations, compared to crystallography, where crystal packing effects can affect these parameters. As we write in response to Reviewer #2, it has been difficult to propose a direct structural mechanism for transmission of the allosteric signal from the a-site in the ATP-cone to the active site and GRD given that the ATP-cones and linker are disordered in the dATP-bound dimers and only partly ordered in the dATP-bound tetramers. Further verification experiments will be performed in future but are outside the scope of the present article.

We will improve the presentation of the biochemical data in a revised version.

General comments:

(1) It would be ideal to perform an additional experiment of some type to confirm the orderdisorder phenomena observed in the cryo-EM structures to rule out the possibility that it is an artifact of the structure determination approach. Circular dichroism might be a possibility?

Circular dichroism reports only on the approximate relative proportions of helix, sheet and loop structure in a protein, thus we believe that it would not be a sensitive enough tool to distinguish between ordered and disordered states. We are considering what alternative methods might be appropriate.

(2) Does the disordering phenomenon of one subunit in the ATP-bound structures have any significance - could it be related to half-of-sites activity? Does this RNR exhibit half-of-sites activity?

Half-of-sites activity has not been biochemically proven in any ribonucleotide reductase in spite of the fact that it was first suggested in 1987 (PMID: 3298261). However, strong structural indication was recently published in the form of the holo-complex of the class Ia ribonucleotide reductase from Escherichia coli, which is highly asymmetrical and in which productive contacts forming an intact proton-coupled electron transfer pathway are only formed between one of two pairs of monomers (PMID: 32217749). We have not been able to prove half-of-sites activity for PcNrdD due to low overall radical content, but the structural results are indeed consistent with such an activity.

(3) Does the disordering of the GRD with dATP bound have any long-term impact on the stability of the Gly radical? I realize that the authors tested the ability to form the Gly radical in the presence of dATP in Fig. 4 of the manuscript. But it looks like they only analyzed the samples after 20 min of incubation. Were longer time points analyzed?

Radical content was measured after 5 min and 20 min incubation; 5 min incubations (not included in the manuscript) consistently gave higher radical content compared to 20 min incubation. Longer time points were not analysed, as we assumed that the radical content would be even lower after 20 min.

(4) Did the authors establish whether the effect of dATP inhibition on substrate binding is reversible? If dATP is removed, can substrates rebind?

This is an interesting question. We measured KDs for dATP in the micromolar range and are hence confident that dATP binding is reversible. Our measurements do not, however, directly prove that inhibition of the enzyme is reversible. Nevertheless, it is worth noting that the protein as purified was precipitated and analysed by the UV-visible spectrum. The aspurified PcNrdD contained 30% nucleotide contamination. The as-purified sample was then analysed by HPLC and we identified a major peak, corresponding to dATP/dADP. Therefore, purification conditions had to be optimised to remove the nucleotides. This is evidence that PcNrdD that has “seen” dATP can subsequently bind substrates in the presence of ATP. We will describe the purification more clearly in a revision.

(5) In some figures (Fig. 6e, for example), the cryo-EM density map for the nucleotide component of the model is not continuous over the entire molecule. Can the authors comment on the significance of this phenomenon? Were the ligands validated in any way to ensure that the assignments were made correctly?

Indeed we sometimes saw discontinuous density for the nucleotides, both in the active site and in the specificity site. However, the break was almost always near the C5’ carbon atom, which is common to all nucleotides. While we cannot readily explain this phenomenon, the nucleotides refined well with full occupancy, giving B-factors similar to those of the surrounding protein atoms. The identity of the nucleotide could always be inferred from a) the size of the base (purine or pyrimidine); b) the known nucleotide combinations added to the protein before grid preparation; c) prior knowledge on the combinations of effector and substrate that have been found valid for all RNRs since the first studies of allosteric specificity regulation.

Reviewer #2 (Public Review):

This manuscript describes the functional and structural characterization of an anaerobic (Class III) ribonucleotide reductase (RNR) with an ATP cone domain from Prevotella copri (PcNrdD). Most significantly, the cryo-EM structural characterization revealed the presence of a flap domain that connects the ATP cone domain and the active site and provides structural insights about how nucleotides and deoxynucleotides bind to this enzyme. The authors also demonstrated the catalytic functions and the oligomeric states. However, many of the biochemical characterizations are incomplete, and it is difficult to make mechanistic conclusions from the reported structures. The reported nucleotide-binding constants may not be accurate because of the design of the assays, which complicates the interpretation of the effects of ATP and dATP on PcNrdD oligomeric states. Importantly, statistical information was missing in most of the biochemical data. Also, while the authors concluded that the dATP binding makes the GRD flexible based on the absence of cryo-EM density for GRD in the dATP-bound PcNrdD, no other supports were provided. There was also a concern about the relevance of the proposed GRD flexibility and the stability of Gly radical. Overall, the manuscript provides structural insights about Class III RNR with ATP cone domain and how it binds ATP and dATP allosteric effectors. However, ambiguity remains about the molecular mechanism by which the dATP binding to the ATP cone domain inhibits the Class III RNR activity.

Strengths:

(1) The manuscript reports the first near-atomic resolution of the structures of Class III RNR with ATP domain in complex with ATP and dATP. These structures revealed the NxN flap domain proposed to form an interaction network between the substrate, the linker to the ATP cone domain, the GRD, and loop 2 important for substrate specificity. The structures also provided insights into how ATP and dATP bind to the ATP cone domain of Class III RNR. Also, the structures suggested that the ATP cone domain is directly involved in the tetramer formation by forming an interaction with the core domain in the presence of dATP. These observations serve as an important basis for future study on the mechanism of Allosteric regulation of Class III RNR.

(2) The authors used a wide range of methodologies including activity assays, nucleotide binding assays, oligomeric state determination, and cryo-EM structural characterization, which were impressive and necessary to understand the complex allosteric regulation of RNR.

(3) The activity assays demonstrated the catalytic function of PcNrdD and its ability to be activated by ATP and low-concentration dATP and inhibited by high-concentration dATP.

(4) ITC and MST were used to show the ability of PcNrdD to bind NTP and dATP.

(5) GEMMA was used successfully to determine the oligomeric state of PcNrdD, which suggested that PcNrdD exists in dimeric and tetrameric forms, whose ratio is affected by ATP and/or dATP.

Weaknesses:

(1) Activity assays.

The activity assays were performed under conditions that may not represent the nucleotide reduction activity. The authors initiated the Gly radical formation and nucleotide reduction simultaneously. The authors also showed that the amount of Gly radical formation was different in the presence of ATP vs dATP. Therefore, it is possible that the observed Vmax is affected by the amount of Gly radical. In fact, some of the data fit poorly into the kinetic model. Also, the number of biological and technical replicates was not described, and no statistical information was provided for the curve fitting.

The highest turnover activity of PcNrdD measured in presence of ATP was 1.3 s-1 (470 nmol/min/mg), a kcat comparable to recently reported values for anaerobic and aerobic RNRs from Neisseria bacilliformis, Leeuwenhoekiella blandensis, Facklamia ignava, Thermus virus P74-23, and Aquifex aeolicus (PMID: 25157154, PMID: 29388911, PMID: 30166338, PMID: 34314684, PMID: 34941255). The general trend illustrated in Figure 1 is that ATP has an activating effect on enzyme activity, whereas high concentrations of dATP have an inactivating effect on activity, which cannot be explained by suboptimal assay conditions since our EPR results consistently show that more radical is formed in incubations with dATP compared to incubations with ATP. Curve fitting methods used are listed in Materials and Methods (as specified in the Figure 1 legend), and standard errors for all specified curve fitting results (from triplicate experiments) are shown in Figure 1.

(2) Binding assays.

The interpretation of the binding assays is complicated by the fact that dATP binds both a- and s-sites and ATP binds a- and active sites. dATP may also bind the active site as the product. It is unknown if ATP binds s-site in PcNrdD. Despite this complexity, the binding assays were performed under the condition that all the binding sites were available.

Therefore, it is not clear which event these assays are reporting.

Both ITC and MST experiments involving ATP and dATP binding to the a-site were performed in the presence of at least 1 mM GTP substrate (5 mM in MST) to fill the active site, and 1 mM dTTP effector to fill the s-site (specified in the legend to Figure 2). These conditions enable binding of ATP or dATP only to the a-site in the ATP-cone.

(3) Oligomeric states.

Due to the ambiguity in the kinetic parameters and the binding constants determined above, the effects of ATP and dATP on the oligomeric states are difficult to interpret. The concentrations of ATP used in these experiments (50 and 100 uM) were significantly lower than KL determined by the activity assays (780 uM), while it is close to the Kd values determined by ITC or MST (~25 uM). Since it is unclear what binding events ITC and MST are reporting, the data in Figure 3 does not provide support for the claimed effects of ATP binding. For the effects of dATP, the authors did not observe a significant difference in oligomeric states between 50 or 100 uM dATP alone vs 50 uM dATP and 100 uM CTP. The former condition has dATP ~ 2x higher than the Kd and KL (Figure 1b) and therefore could be considered as "inhibited". On the other hand, NrdD should be fully active under the latter condition. Therefore, these observations show no correlation between the oligomeric state and the catalytic activity.

The results in Figure 3 show that at in presence of 100 µM ATP plus 100 µM CTP the oligomeric equilibrium is 64% dimers plus 36% tetramers, and in presence of 50-100 µM dATP the oligomeric equilibrium is 32% dimers and 68% tetramers. We agree that there is no clear and strong correlation between oligomeric state and inhibition. We will also try to make it clearer in a revised version. Meanwhile, in order to add some clarity to our observations, SEC experiments at higher nucleotide concentrations will be done to strengthen our observations.

(4) Effects of dATP binding on GRD structure

One of the key conclusions of this manuscript is that dATP binding induces the dissociation of GRD from the active site. However, the structures did not provide an explanation for how the dATP binding affects the conformation of GRD or whether the dissociation of GRD is a direct consequence of dATP binding or it is due to the absence of nucleotide substrate. Also, Gly radical is unlikely to be stable when it is not protected from the bulk solvent. Therefore, it is unlikely that the GRD dissociates from the active site unless the inhibition by dATP is irreversible. Further evidence is needed to support the proposed mechanism of inhibition by dATP.

We admit that it has been difficult to propose a direct structural mechanism for transmission of the allosteric signal from the a-site in the ATP-cone to the active site and GRD given that the ATP-cones and linker are disordered in the dATP-bound dimers and that the linker can only be partly modelled in the dATP-bound tetramers. Most likely dATP binding causes a change in the dynamics of the linker region and NxN flap that directly affects substrate binding and simultaneously causes disorder of the GRD, given that all are part of a connected system (described as “nexus” in the manuscript). The structures determined in the presence of dATP and CTP show that CTP cannot bind in the absence of an ordered NxN flap.

In any case a major conclusion of the work is that dATP does not inhibit the anaerobic RNR by prevention of glycyl radical formation but by prevention of its subsequent transfer. We agree that further evidence is required to support the proposed mechanism, but given the extent of the data already presented in the manuscript, we feel that such studies should be the subject of a future publication.

(5) Functional support for the observed structures.

Evidence for connecting structural observations and mechanistic conclusions is largely missing. For example, the authors proposed that the interactions between the ATP cone domain and the core domain are responsible for tetramer formation. However, no biochemical evidence was provided to support this proposal. Similarly, the functional significance of the interaction through the NxN flap domain was not proved by mutagenesis experiments.

We did actually make mutants to verify the observed interactions, but several of them did not behave well in our hands, e.g. with regard to protein stability. Since we have no evidence that oligomerisation is coupled to inhibition, and since we did not observe any conservation between protein sequences in the interaction area, we chose not to pursue this point further. The main merit of the tetramer structures is that they allowed a high-resolution view of dATP binding to the ATP-cone and a comparison to previously-observed ATP-cones. Nevertheless, mutation experiments, also including the NxN flap, could be the subject of future work.

Reviewer #3 (Public Review):

The manuscript by Bimai et al describes a structural and functional characterization of an anaerobic ribonucleotide reductase (RNR) enzyme from the human microbe, P. copri. More specifically, the authors aimed to characterize the mechanism by how (d)ATP modulates nucleotide reduction in this anaerobic RNR, using a combination of enzyme kinetics, binding thermodynamics, and cryo-EM structural determination. One of the principal findings of this paper is the ordering of a NxN 'flap' in the presence of ATP that promotes RNR catalysis and the disordering of both this flap and the glycyl radical domain (GRD) when the inhibitory effector, dATP, binds. The latter is correlated with a loss of substrate binding, which is the likely mechanism for dATP inhibition. It is important to note that the GRD is remote (>30 Ang) from the binding site of the dATP molecule, suggesting long-range communication of the structural (dis)ordering. The authors also present evidence for a shift in oligomerization in the presence of dATP. The work does provide evidence for new insights/views into the subtle differences of nucleotide modulation (allostery) of RNR through long-range interactions.

The strengths of the work are the impressive, in-depth structural analysis of the various regulated forms of PcRNR by (d)ATP using cryo-EM. The authors present seven different models in total, with striking differences in oligomerization and (dis)ordering of select structural features, including the GRD that is integral to catalysis. The authors present several, complementary biochemical experiments (ITC, MST, EPR, kinetics) aimed at resolving the binding and regulatory mechanism of the enzyme by various nucleotides. The authors present a good breadth of the literature in which the focus of allosteric regulation of RNRs has been on the aerobic orthologues.

Given the resolution of some of the structures in the remote regions that appear to be of importance, the rigor of the work could have been improved by complementing this experimental studies with molecular dynamics (MD) simulations to reveal the dynamics of the GRD and loops/flaps at the active site.

We have discussed with expert colleagues the possibility of carrying out MD simulations on the different states in order to study the differential effects of ATP and dATP binding on the dynamics of the GRD. However, they felt that the chance of obtaining meaningful results was low, particularly since some structural elements are missing from the models for both forms, in particular the linker between the ATP-cone and the core.

The biochemical data supporting the loss of substrate binding with dATP association is compelling, but the binding studies of the (d)ATP regulatory molecules are not; the authors noted less-than-unity binding stoichiometries for the effectors.

Most of the methods used measure only binding strength, not the number of binding sites (N), whereas ITC also measures number of sites. N is dependent on the integrity of the protein, i.e. the number of protein molecules in a preparation that are involved in binding, and quite often gives lower values than the theoretical number of binding sites.

Also, the work would benefit from additional support for oligomerization changes using an additional biochemical/biophysical approach.

SEC (chromatography), GEMMA (mass spectrometry) and cryo-EM were used to study oligomerization. Since each method has restrictions on nucleotide concentrations as well as protein concentrations that can be used, the results are not directly comparable, but all three methods indicate nucleotide dependent oligomerization changes. The SEC results will be included in a revised version.

Overall, the authors have mostly achieved their overall aims of the manuscript. With focused modifications, including additional control experiments, the manuscript should be a welcomed addition to the RNR field

Recommendations for the authors: Reviewer #1 (Recommendations For The Authors):

(1) The last sentence of the abstract is not complete. The structures implicate a complex network of interactions in ... ? What do they implicate?

A couple of words seem to have been missed from the abstract. We have rewritten the end of the abstract to emphasise better that the dynamical transitions involve a linked network of interactions and not just the GRD.

(2) A reference is needed in the second sentence of the introduction.

We have added a reference as requested.

(3) Page 2, paragraph 2. The authors state "two beta subunits (NrdB) harboring a stable radical." This is not accurate. First of all, each beta subunit harbors its own cysteine oxidant.

And in several subclasses, that oxidant is not a stable radical but an oxidized metal cluster. Please revise to improve accuracy and also provide appropriate references.

We have revised the description and added a recent reference.

(4) Page 4, Fig. 1, panels C and D. The fit of the curve to the data is pretty poor. Is there an explanation? Could the data be improved in some way? In general, it is also best practice nowadays to show the individual data points in addition to the error bars in plots like the ones shown in Figure 1. Please modify the plots to include the individual data points in this figure - and probably also the subsequent figures showing binding data.

We have modified relevant panels in Figures 1, 2 and 5 as requested.

(5) Page 12, first paragraph. The authors state that one of the monomers in the ATP-CTP structure is well ordered and the other is less ordered. It would be ideal to show in a figure the basis for this conclusion using the cryo-EM maps. The "less ordered" monomer appears to be fully modeled.

Since the 2-fold axis of the dimer is vertical, the GRD of the left-hand monomer is hidden from view at the back of the molecule in Figure 6. For this monomer there was a small amount of density that allowed modelling of part of the glycyl radical loop (though not the tip containing the radical Gly itself) and the NxN flap, albeit with significantly higher mobility. We have illustrated this through an additional supplement for Figure 6 (figure supplement 2) in which the B-factors of the residues are shown both as a ribbon with radius proportional to the B-factor and through colouring. We hope that the four views in Figure 6 (figure supplement 2) together illustrate the relative mobility of different parts of the dimer.

It would also be ideal to show the basis for the conclusion that the entire GRD is disordered in the dATP-bound dimer structure.

Thank you for this suggestion. We have added a fifth supplement to Figure 8 in which we show the cryo-EM reconstruction for the dATP-bound dimer in two orientations, with the ATP-CTP-bound structure superimposed, which clearly shows that the entire GRD, the ATPcones, linker and NxN flap are all disordered in both monomers.

Reviewer #2 (Recommendations For The Authors):

(1) Units to describe enzyme activity.

The unit for the specific activity in the main text (nmol/min•mg) is unusual. It is most likely a typo of nmol/min/mg or nmol/(min•mg).

We have changes to nmol/min/mg in the text.

The unit for the Vmax is unusual and should not be confused with the specific activity. By definition, Vmax is the velocity of a reaction at a defined enzyme concentration/amount. For example, if an assay of 10 mg enzyme yielded 470 nmol of product in 1 min, Vmax is 470 nmol/min, whereas the specific activity is 47 nmol/min/mg.

The velocity as calculated above is ca 1.3 s-1. We have added kcat values to accompany the specific activities given.

(2) Steady-state kinetic analysis.

The steady-state kinetic analysis in Figure 1 needs to be repeated. While the nonlinear curve fitting for Figure 1a is reasonable, those in Figures 1b, 1c, and 1d were outside the error range. Consequently, the reported kinetic parameters are unlikely accurate. The authors should repeat the assays with different enzyme preparation to account for all the errors. If the fit curve is still outside the error range, the kinetic model is likely incorrect, and the authors need to investigate different kinetic models.

The replotted Figure 1 now includes two different experiments for 1b (four replicates in total).

The authors should report the number of replicates and the statistical data for the curve fitting.

The figure legend has been updated with statistical data for all curve fits, and the number of replicates has been added.

The authors should report Vmax, Ki, and KL for Figure 1d.

Results in Figures 1c and 1d are less straightforward than those in Figures 1a and 1b where the s-site is filled with dTTP, favouring binding of GTP to the active site. The curve fit in Figure 1c is disturbed at high concentrations of ATP, which plausibly competes with the CTP substrate and results in inhibition by formed dATP. The curve fit in Figure 1d is less certain since reduction of substrate is low due to intrinsic CTP reduction in absence of effector and partially overlapping activation and inhibition effects of dATP.

The authors should consider presenting the data in a log scale because of the complex nature of the activation/inhibition at the lower concentrations of dATP.

Log scale plots are included as insets in Figures 1b and 1d.

• The basal level of CPT reduction in the absence of an effector nucleotide should be reported with an error.

The error value has been added in the figure legend for the basal level of CTP reduction in the absence of effector.

(3) Equations for the kinetic analysis.

-The equations should be numbered and referred to in the Figure 1 legend.

All equations are specified and numbered in Materials and Methods. The equation used for each curve fit in the panels in Figure 1 is specified in the figure legend.

-KL must be defined in the main text. I suppose this is Kd for ATP or dATP. The equation for KL determination is missing brackets for dNTP.

KL (the concentration of an allosteric effector that gives half maximal enzyme activity) is defined in Materials and Methods where the equation is described. KL is not the same as KD (the dissociation constant for a ligand and its receptor). Brackets have been added to equation 1.

I believe dNTP in the first equation is incorrect because ATP was the ligand for Figures 1A and 1C.

[dNTP] in the first equation has been changed to [NTP/dNTP] to indicate that both ribonucleotides and deoxyribonucleotides can bind.

The second equation can be expressed as dATP as I believe this is the only ligand that inhibits the enzyme.

We prefer to keep the more general [dNTP] in the equation.

The equation used for the fitting in Figure 1d must be defined more clearly than "a combination of the two equations".

The equation used for the curve fit in Figure 1d has been specified as equation 3 in Materials and Methods.

(4) Design of the activity assays

It is not clear if the activity assays report the rate of glycyl radical formation or nucleotide reduction. The authors mixed NrdD and NrdG and initiated the reaction by adding formate (essential for nucleotide reduction) and dithionite (Gly radical formation). The Gly radical formation is slow (in min time scale). The authors reported that ATP/dATP affected the rate of Gly radical formation and in the presence of ATP, Gly radical formation was incomplete even after 20 min. Therefore, it is possible that within the timescale of the activity assays (5 min), the reactions could be partially limited by the Gly radical formation, which may be the reason for the poor curve fitting.

Activity assays were performed with 5 min pre-incubation without dithionite and formate (no glycyl radical formation) and 10 min incubation after addition of dithionite and formate (glycyl radical formation plus substrate reduction). During earlier tests, NrdD and NrdG were first preincubated in the presence of dithionite (glycyl radical formation) and after addition of formate the substrate reduction was monitored during 20 min. These experiments resulted in lower enzyme activity, whereas higher activity was achieved only upon formate addition to the preincubation reaction. We suppose that the presence of dithionite, which is a strong reducing agent, affected NrdD stability and the reaction was stabilised by the presence of formate at an earlier stage of the reaction. For the EPR conditions used in the paper, 5 min incubation gave higher radical content compared to 20 min, and the reported activity assay gave highest activity after 10 min incubation; kcat of 1.3 s-1.

(5) Methods section for the activity assays.

The concentration of dTTP, ATP, and dATP used in the assays must be described.

We thank the reviewer for pointing out this omission and we have now specified the concentrations used.

Although the authors mentioned that they changed the concentration of dTTP, such data were not presented. Is this correct? Did the authors fix the dTTP concentration for the GTP reduction?

We apologise for the ambiguity and have specified that the dTTP concentration was fixed at 1 mM in the GTP experiments and that only the ATP or dATP concentrations were varied.

(6) Discrepancy between Ki/KL and Kd.

There is a significant ambiguity remaining about the binding event that the ITC and MST results are reporting. Although dATP binds to both a- and s-sites and ATP binds to both active site and a-site, only a single binding event was observed in both cases. To distinguish the dATP binding to a- and s-sites and the active site, the authors should perform binding assays using mutant enzymes with only one of the binding sites available for dATP/ATP binding.

MST and ITC were performed in presence of substrate (1 mM GTP) and s-site effector (1 mM dTTP in ITC experiments, and 5 mM dTTP in MST experiments), thus dATP is blocked from binding to the s-site and ATP from binding to the active site.

There are significant differences between Kd determined by MST or ITC and Ki/KL determined by the activity assays. Kd measurements were performed in the absence of the substrate nucleotides, while the assays required substrates. There may be complications from the presence of NrdG and the Gly radical formation. The authors must clearly describe all these complications and the discrepancy between Kd and Ki/KL.

MST, ITC and enzyme assays were all performed in the presence of substrate, and enzyme assays also contained NrdG, which was not present in the MST and ITC analyses. While KD is a thermodynamic constant representing the affinity of ligand to its binding site - in our case an effector nucleotide to the ATP-cone, KL is a kinetic constant (the allosteric effector concentration that gives half maximal activity) representing the relationship between the effector concentration and the reaction speed and is affected by the enzyme turnover number (kcat). The relationship between KD, KL and Ki is further complicated by conformational and possibly oligomeric state changes of NrdD upon binding of allosteric effectors, which occurs on a slower time scale than the rapid exchange of nucleotides in allosteric sites.

The results of ATP/dATP copurification experiments shown in Figure 2 - figure supplement 1 show the preference of dATP binding over ATP. However, the results do not necessarily support the competition between ATP and dATP for binding to the ATP cone domain. It is still possible that dATP binding to the s-site diminishes the binding of ATP to the a-site.

Our aim was to exclude the possibility that ATP and dATP can bind to the ATP-cone at the same time and not to study competition between the two. Nevertheless, to eliminate the possibility that dATP binding to the s-site could affect nucleotide binding to the a-site, in two out of three conditions described in the supplementary figure, the experiments were performed in the presence of dTTP to prevent binding of dATP to the s-site.

(7) Oligomeric states.

The authors must present the GEMMA results without ATP or dATP. Otherwise, the effects of ATP and dATP on the oligomeric state are not clear.

We cannot report GEMMA results without ATP or dATP because apo-PcNrdD was unstable in the GEMMA buffer and clogged the capillaries. Instead, SEC analysis was performed on apo-PcNrdD in a more suitable buffer and showed a homogeneous peak corresponding to a dimer (included as Figure 3 - figure supplement 1).

Figure 3 does not support the induction of a2 upon ATP binding. The concentrations of ATP used in these experiments (50 and 100 uM) were significantly lower than KL determined by the activity assays (780 uM), while it is close to the Kd values determined by ITC or MST (~25 uM). Since it is unclear what binding events ITC and MST are reporting, the data in Figure 3 does not provide support for the claimed effects of ATP binding.

MST and ITC were performed in the presence of substrate (1 mM GTP) and s-site effector (1 mM dTTP in ITC experiments, and 5 mM dTTP in MST experiments), and they thus measure binding of ATP or dATP to the ATP cone. SEC analysis with 2 µM apo-PcNrdD and higher nucleotide concentrations (1 mM) was performed, confirming the presence of both dimers and tetramers in solution at different ratios depending on the addition of ATP or dATP. The SEC analysis, included as Figure 3 - figure supplement 1, confirms the existence of an equilibrium in solution.

The effects of dATP must be presented more clearly. The authors did not observe a significant difference in oligomeric states between 50 or 100 uM dATP vs 50 uM dATP and 100 uM CTP. The former condition has dATP ~ 2x higher than the Kd and KL (Figure 1b) and therefore could be considered as "inhibited". On the other hand, NrdD should be fully active under the latter condition. The absence of difference in the oligomeric states between these two different conditions suggested to me that the oligomeric state does not regulate the NrdD activity. The authors seemed to indicate the same conclusion, but did not describe it clearly.

We agree that the oligomeric state most likely does not regulate the NrdD activity and hope to have explained this better in the revised version.

Figure 3 legend mentioned a and b, but the figure was not labeled.

We have corrected this.

The authors should triplicate the analysis and report the errors.

Five scans were added for each trace to increase the signal-to-noise level (included in figure legend).

(8) EPR characterization of Gly radical

The amount of Gly radical must be quantified by EPR. The authors must report how much NrdD has Gly radical.

The concentration of NrdD (1 µM) in the activity assays is too low to be quantified by EPR. In the EPR experiment the glycyl radical content is given in the figure legend.

The authors claim that the Gly radical environment was similar based on the doublet feature. However, the double feature comes from the hyperfine splitting with α proton whose orientation relative to the radical p-orbital would not be affected by the conformation or the environment. Thus, this conclusion is incorrect and must be removed.

We thank the reviewer for the clarifying comment and have removed our suggestion in the text.

(9) Gly711 should be shown in Fig. 6e to help readers understand the last paragraph on page 12.

The figure reference has been changed to Fig. 7, where this is shown more clearly. In Fig. 6e, inclusion of Gly711 would obscure other important information.

(10) GRD structure with dATP

The disorder of GRD in the presence of dATP does not agree with the formation of Gly radical under the same conditions. Gly radical is unlikely stable if it is extensively exposed to solvent. Most likely, the observed cryo-EM structures represent the conformation irrelevant to Gly radical formation.

We agree that the glycyl radical is unlikely to be stable if exposed to solvent. We believe that the GRD is not completely disordered but most likely made more mobile through rigid body movements of the domain to an extent that makes it invisible in the cryo-EM maps. It is most likely still in the vicinity of the active site, shielding the glycyl radical. Our new HDX-MS results show a small but tangible increase in mobility of the GRD in the presence of dATP compared to ATP. Of course the differences in dynamics remain to be confirmed. It is worth noting that the group of Catherine Drennan at MIT published a conference abstract more than a year ago that suggested a similar pattern of ordered/dynamic GRDs, based on crystal structures, though the details have not yet been published (https://doi.org/10.1096/fasebj.2022.36.S1.R3407).

We also agree that the cryo-EM structures do not show the GRD conformation relevant to Gly radical formation, as this has been shown spectroscopically for the GRE pyruvate formate lyase to require large conformational changes in the GRD and also the presence of the activase. However, revealing this conformation would be a completely different project. We postulate that inactivation proceeds by prevention of radical transfer to the substrate, not by prevention of its formation.

We have altered the wording in several places in the revised manuscript, including the title, to avoid using the term “disorder”, as this may imply (partial) unfolding, and we certainly do not wish to imply that.

(11) The difference between dATP and ATP binding

From the presented structures, it was not clear how the absence of 2'-OH affects the oligomeric state and the structure of the GRD. The low resolution of the ATP-bound structure precluded the comparison between the ATP and dATP-bound structures.

We agree that a detailed analysis of the differences between ATP- and dATP-bound structures requires higher resolution structures, particularly of the ATP-bound form. This will be the subject of future studies.

(12) Conclusion about the disordered GRD.

-The authors should describe the reason why the dATP binding affected the structure of GRD. The authors did not discuss why dATP binding affected the folding or mobility of GRD. Since this is the key conclusion of this manuscript and the authors are making this conclusion based on the absence of the ordered GRD structure (hence the negative results), the authors should carefully describe why the dATP binding does not allow the binding/folding of GRD in the position observed in the ATP-bound structure.

As mentioned in our response to point 4 in this reviewer’s Public Review, it is difficult to propose a direct structural mechanism for transmission of the allosteric signal from the a-site in the ATP-cone to the active site and GRD given that the ATP-cones and linker are disordered in the dATP-bound dimers and that the linker cannot be completely modelled even in the dATP-bound tetramers. Our first hypotheses were that the ATP-cone might work by a steric occlusion mechanism, but the reality appears more complex. Most likely dATP binding causes a change in the dynamics of the linker region and NxN flap that directly affects substrate binding and simultaneously causes higher mobility of the GRD, given that all are part of a connected system. The structures determined in the presence of dATP and CTP show that CTP cannot bind in the absence of an ordered NxN flap. We hope that future structural studies of NrdDs from other organisms may shed further light on this mechanism.

The authors should test if the dATP inhibition is reversible for PcNrdD. If dATP binding induces dissociation of GRD from the active site and makes GRD flexible, Gly radical would most likely be quenched by formate or other components in the assay solution. If dATP inhibition is reversible, it is hard to believe that Gly radical dissociates completely from the active site.

As-purified PcNrdD contains dATP and can after removal of bound nucleotides bind substrate in presence of ATP. The as-purified PcNrdD protein contained 30% nucleotide contamination. After precipitation, HPLC analysis identified a major peak corresponding to dATP/dADP. Purification conditions were optimised to remove the nucleotides and we have added this information to the purification description.

(13) Functional support for the observed structures.

Similar to X-ray crystallography, cryo-EM is a highly selective method that requires the selection of particles that can be analyzed with sufficient resolution. This means that the analysis could be biased towards the protein conformations stable on the cryo-EM grid. Consequently, testing the structural observations by functional characterization of mutant enzymes is critical. However, the authors did not perform such functional characterizations and made conclusions purely based on the structural observations.

We acknowledge this limitation. We constructed several mutations located at the tetrameric interface between the ATP-cone and the core protein based on the cryo-EM structure of dATP loaded NrdD. Unfortunately, these mutant proteins were unstable and led to protein cleavage.

(14) Other minor points:

In the introduction, the authors stated "The presence and function of the ATP-cone domain distinguish anaerobic RNRs from the other members of the large glycyl radical enzyme (GRE) family that are otherwise structurally and mechanistically related (Backman et al., 2017)." This statement is misleading because GREs are functionally diverse.

We have removed the words “and mechanistically” to reduce ambiguity.

p. 12, e.g. should be removed.

We are not sure what is meant here. Does the reviewer mean p. 21 “The interactions are mostly hydrophobic but are reinforced by several H-bonds, e.g. between Gln3D-Gln458A, Ser53D–Gln458A, Arg11D-Asp468A, the main chain amide of Ile12D and Tyr557A.”?

Reviewer #3 (Recommendations For The Authors):

Overall, the work presents an impressive and in-depth structural view of the conformational changes stemming from the interactions of (d)ATP allosteric effector molecules that are interrelated to RNR function. The manuscript is written clearly and provides a solid overview of RNR chemistry. The cryo-EM data show striking differences between ATP and dATP bound forms, though in select regions, the resolution is not good enough for strong interpretations of the finer details.

(1) In cryo-EM structures, dATP appears to shift the oligomerization equilibrium from nearly all dimeric forms (absence of dATP) to a mixture of both dimeric and tetrameric species (presence of dATP). The examination of the oligomeric composition in solution using the GEMMA - a mass spectral technique - showed somewhat similar trends, though given the magnitude of the differences, it was less compelling. Have the authors considered a complementary solution technique, such as analytical SEC or dynamic light scattering that could provide further support for the change in oligomerization as observed in the cryo-EM?

SEC analysis with 2 µM apoPcNrdD and higher nucleotide concentrations (1 mM) was performed, confirming the presence of both dimer and tetramer in solution at different ratios depending on the addition of ATP or dATP. The SEC analysis, included as Figure 3 - figure supplement 1, confirms the existence of an equilibrium in solution.

(2) The protein as isolated from the final SEC shows a predominant peak corresponding to aggregate protein. It would be helpful if the authors ran an analytical SEC on the protein sample that is more refined to see how much soluble dimer/tetramer vs. aggregate protein there is. This could impact the kinetic and thermodynamic analysis of effector interactions. Further, the second major peak is labeled as 'monomer'. Is the protein isolated as a monomer and then forms dimer upon effector binding? It is unclear. The authors should consider presenting the SEC standards for the given column and buffer condition so that a reasonable estimate of the oligomerization status of the isolated protein can be assigned.

Can the reviewer possibly have believed that Figure 1 - supplementary figure 2a shows PcNrdD rather than PcNrdG? The figure supplement corresponds to the as-isolated SEC analysis of the activase (PcNrdG), which shows the presence of two main peaks of aggregates and monomer. The monomeric peak was reinjected and showed no presence of further aggregation states. Currently it is not known which oligomeric state the activase harbours upon binding to PcNrdD and glycyl radical formation. None of the other SEC figures in the MS has any predominant peak corresponding to aggregated protein.

(3) More details are needed for the ITC section. The ITC methods are not clear. What is the exact composition of the ligand solution being titrated into the protein solution? It is unclear how the less-than-unity binding stoichiometry was determined and what it means. Is the n value for the monomer, dimer, or tetramer forms? It is concerning that n < 1 is observed for dATP binding in the ITC whereas there are 3 dATP bound/subunit in the cryo-EM. For completeness, titration of a buffer into protein solution (no ligand) should be conducted and presented to demonstrate that the heats produced in Figure 2 correspond to the ligand only (and not a buffer mismatch).

ITC experiments were performed in the presence of 1 mM GTP (c-site) and 1 mM dTTP (ssite). Unlike other parameters in ITC analyses, the N value is usually the least accurate of all fitted parameters and strongly depends on the concentration of the active protein in the sample. N values described in the current study are in the same range as values reported for ATP-cones in other RNRs and NrdR (Rozman Grinberg & al 2018a, 2018b, 2022 McKethan and Spiro 2013). The results most likely reflect two high-affinity binding sites for dATP and one high affinity binding site for ATP. Different nucleotide concentrations were used in the cryoEM and ITC experiments.

(4) It is intriguing that the binding of dATP doesn't quell the glycyl radical. In fact, it appears that, as the authors suggest, the amount of glycyl radical might be increased in these samples. However, the cryo-EM data indicates that the GRD is disordered. It is unclear how these would be correlated, as one would not expect a disordered structural element to maintain such a potent oxidant.

As already written above, we do not wish to imply that the GRD is completely or even highly disordered, just that its dynamics increase in the presence of dATP. Otherwise we completely agree that a very exposed Gly radical is incompatible with its stability. It could be that the amount of disorder is exaggerated somewhat by the vitrification process in cryo-EM. We have tried to reword some of the text to emphasise higher mobility rather than disorder.

It has been difficult to propose a direct structural mechanism for transmission of the allosteric signal from the a-site in the ATP-cone to the active site and GRD given that the ATP-cones and linker are disordered in the dATP-bound dimers and that the linker can not be completely modelled even in the dATP-bound tetramers. We initially thought that a steric occlusion mechanism might be at play, but the reality appears more complex. Most likely dATP binding causes a change in the dynamics of the linker region and NxN flap that directly affects substrate binding and simultaneously causes higher mobility of the GRD, given that all are part of a connected system. The structures determined in the presence of dATP and CTP show that CTP cannot bind in the absence of an ordered NxN flap. We hope that future structural studies of NrdDs from other organisms may shed further light on this mechanism.

(5) It is a bit difficult to keep track of the myriad of structural information and differences amongst the various nucleotide-dependent conditions. It would be useful for the authors to add a summary figure that depicts the various oligomers, orientations, and (dis)ordered structural elements with cartoon representations.

Thank you for this suggestion. It has been added as Figure 11.

(6) The mechanism by which (d)ATP binding changes the (dis)ordering of select loops based on the current cryo-EM data is unclear (even the authors agree). The addition of molecular dynamics (MD) simulations on two different structures to reveal the network or structural communication would be a great addition to the work and validate the structural data.

We have discussed this with a colleague who is an expert in MD. Their advice was that such simulations would be very difficult given that some amino-acids are missing in both of the relevant starting structures (ATP-CTP and dATP-CTP dimer) and could give very variable results. Thus we chose to do complementary experiments with hydrogen-deuterium exchange mass spectrometry (HDX-MS) instead. The results are included in the revised manuscript.

Minor points

(1) There are some conflicting reports as to whether P. copri is considered a human 'pathogen'. According to Yeoh, et al Scientific Reports 2022, P. copri is one of the predominant microbes in the human gut and is linked to a positive impact on metabolism. Perhaps the addition of a citation that provides support for it as a pathogen would clarify the statement on p. 3.

We have added a recent reference (Nii T, Maeda Y, Motooka D, et al. (2023) Genomic repertoires linked with pathogenic potency of arthritogenic Prevotella copri isolated from the gut of patients with rheumatoid arthritis. Ann Rheum Dis 82: 621-629. doi: 10.1136/annrheumdis-2022-222881).

(2) In Figure 3, the number of dimers/tetramers for dATP (100 uM) does not add up to 100.

What is the other 2%?

Thank you for pointing this out - it has been corrected.

(3) The data in Figures 5C and D do show slight changes that could be fit and interpreted as a 'weak' interaction. Thus, the statement on p 9 "where dATP-loaded PcNrdD could bind neither GTP nor CTP" should be changed to indicate that the interactions are weak (or that the nucleotides weakly associate).

The text and the figure have been changed according to the reviewer’s suggestion.

Version 2

Abstract

Abstract

A small, nucleotide-binding domain, the ATP-cone, is found at the N-terminus of most ribonucleotide reductase (RNR) catalytic subunits. By binding ATP or dATP it regulates the enzyme activity of all classes of RNR. Functional and structural work on aerobic RNRs has revealed a plethora of ways in which dATP inhibits activity by inducing oligomerization and preventing a productive radical transfer from one subunit to the active site in the other. Anaerobic RNRs, on the other hand, store a stable glycyl radical next to the active site and the basis for their dATP-dependent inhibition is completely unknown. We present biochemical, biophysical and structural information on the effects of ATP and dATP binding to the anaerobic RNR from Prevotella copri. The enzyme exists in a dimer-tetramer equilibrium biased towards dimers when two ATP molecules are bound to the ATP-cone and tetramers when two dATP molecules are bound. In the presence of ATP, P. copri NrdD is active and has a fully ordered glycyl radical domain (GRD) in one monomer of the dimer. Binding of dATP to the ATP-cone results in loss of activity and increased dynamics of the GRD, such that it can not be detected in the cryo-EM structures. The glycyl radical is formed even in the dATP-bound form, but the substrate does not bind. The structures implicate a complex network of interactions in activity regulation that involve the GRD more than 30 Å away from the dATP molecules, the allosteric substrate specificity site and a conserved but previously unseen flap over the active site. Taken together, the results suggest dATP inhibition in anaerobic RNRs acts by increasing the flexibility of the flap and GRD, thereby preventing both substrate binding and radical mobilisation.

Introduction

Ribonucleotide reductases (RNRs) are a family of enzymes with sophisticated radical chemistries and allosteric regulation. RNRs produce all four deoxyribonucleotides and are the only enzymes providing de novo building blocks for DNA replication and repair in all free-living organisms (Mathews, 2016). Virtually all RNRs possess an allosteric site regulating substrate specificity (the s-site), a crucial aspect of RNR that ultimately provides a balanced supply of dNTPs (Mathews, 2016, 2018). A second allosteric site (the a-site) is located in an N-terminal domain called the ATP-cone (Aravind et al, 2000), found in the majority of RNRs (Jonna et al, 2015). Whereas the s-site binds all deoxyribonucleoside triphosphates (dNTPs) and usually ATP, the a-site in the ATP-cone can only bind ATP and dATP. Binding of ATP activates RNR, and binding of dATP inhibits it (Hofer et al, 2012; Martínez-Carranza et al, 2020).

The RNR family consists of three evolutionarily related classes (I to III) with a common radical-based reaction mechanism but differing in the mode of radical generation and in quaternary structure (Högbom et al, 2020; Lundin et al, 2015). Class I, strictly aerobic RNRs are heterotetramers consisting of two α subunits (NrdA) with binding sites for substrates and allosteric nucleotides and two β subunits (NrdB), that harbour a radical initiator in the form of a stable radical or an oxidized metal cluster (Högbom et al, 2020). Class II RNRs are either dimers or monomers with an α subunit (NrdJ) that, in addition to binding sites for substrate and allosteric nucleotides, also harbours the radical initiator cofactor adenosylcobalamin. Class III RNRs, which only function anaerobically, are dimers of an α subunit (NrdD) with a stable radical close to the active site, and binding sites for allosteric nucleotides. The radical is introduced via encounter with a specific radical-SAM enzyme called NrdG, and once activated NrdD can perform multiple turnovers in the absence of NrdG (Backman et al, 2017; Torrents et al, 2001).

The dATP-dependent inhibition has been biochemically characterised in representatives of all three RNR classes, but only structurally studied in class I. A common denominator of all inhibited class I RNRs is oligomerization, leading to disturbed radical transfer between the α and β subunits. Hitherto four different inhibitory oligomerization mechanisms have been identified: heterooctameric α4β4 complexes in Escherichia coli, Clostridium botulinum and Neisseria gonorrhoeae (Ando et al, 2011; Martínez-Carranza et al, 2020; Wei et al, 2014), an α4 complex in Pseudomonas aeruginosa (Johansson et al, 2016), α6 complexes in the eukaryotes Homo sapiens, Saccharomyces cerevisiae and Dictyostelium discoideum (Ando et al, 2016; Brignole et al, 2018; Crona et al, 2013; Fairman et al, 2011), and a β4 complex in the bacterium Leeuwenhoekiella blandensis (Rozman Grinberg et al, 2018a). All these class I oligomers involve protein-protein interactions mediated by the ATP-cone domain.

The ATP-cone is in essence restricted to RNRs and the RNR-specific repressor NrdR (Rozman Grinberg et al, 2022). Among RNRs it is found in about 80% of class III enzymes and about 50% of class I enzymes, but less than 10% of class II enzymes (Jonna et al., 2015). The presence and function of the ATP-cone domain distinguish anaerobic RNRs from the other members of the large glycyl radical enzyme (GRE) family that are otherwise structurally related (Backman et al., 2017). The allosteric regulation mechanisms of class III RNRs (NrdDs) from E. coli, bacteriophage T4 and Lactococcus lactis were determined several decades ago (Andersson et al, 2000; Eliasson et al, 1994; Torrents et al, 2000). T4NrdD was the first class III RNR structure solved (Logan et al, 1999), and later, structures of Thermotoga maritima NrdD (TmNrdD) have been published (Aurelius et al, 2015; Wei et al, 2014). However, these two NrdDs lack an ATP-cone and the structural basis for allosteric activity regulation in the class III RNRs is still an outstanding question. Here, we present structural, biochemical and biophysical studies on the ATP-cone-containing NrdD from the human pathogen (Nii & al., 2023) Prevotella copri (PcNrdD), showing that the binding of dATP causes increased flexibilityof the C-terminal glycyl radical-bearing domain, as well as a flap-like loop that binds across the top of the active site in the active form. This increase in dynamics is coupled to enzymatic inactivation by inhibition of substrate binding and radical transfer. The final outcome of dATP inhibition is thus blocked radical transfer in both class I and class III RNRs, but it is achieved by completely different mechanisms, i.e. oligomerization prevents radical transfer between subunits in class I RNRs but long-range modulation of the flexibility of key structural elements prevents substrate binding in class III RNRs.

Results

Allosteric activity regulation by ATP and dATP

The amino acid sequence of PcNrdD suggests that it belongs to the formate-requiring class III RNRs (Burnim et al, 2022b; Wei et al., 2014), and initial optimization of the assay composition showed that this was the case (Figure 1 - figure supplement 1). ATP-cone mediated activation of PcNrdD enzyme activity by ATP or inhibition by dATP was tested with two different substrates. With GTP as substrate the s-site was filled with dTTP as effector, and in absence of any a-site effector the basal level of GTP reduction was 21±2 nmol/min/mg. An increasing concentration of ATP stimulated activity to a kcat of 1.3 s-1 (478±26 nmol/min/mg), with a KL of 0.67±0.12 mM (Fig. 1a). On the other hand, an increasing concentration of dATP resulted in an abrupt inhibition of enzyme activity with a Ki of 74±24 µM (Fig. 1b). Hence, the Ki value for dATP inhibition via the ATP-cone is approximately 10-fold lower compared to the KL value for ATP activation via the ATP-cone.

Activity assays of Prevotella copri NrdD in the presence of ATP or dATP.

A 2.5-fold excess of holoNrdG over apoNrdD was used to study the allosteric regulation effect of ATP and dATP on the a-site. GTP reduction was monitored with 1 mM dTTP as effector in the s-site and titrated with ATP (a) or dATP (b) in the a-site. CTP reduction assays were titrated with ATP (c) or dATP (d), in this case acting both as s-site effectors and a-site regulators. Experiments were performed in triplicate for (a), (c), and (d) and with four replicates for (b). Insets in panels (b) and (d) show the results plotted in log scale. Curve fits for calculation of KL and Ki used equations (1) and (2), respectively, given in Materials and Methods, and in (d) equation (3). Curve fits for (c) and (d) used a start activity of 50 nmol/min/mg, and in (c) only results for 0-3 mM ATP were used. The R values for curve fits in panels (a) - (d) were 0.99, 0.95, 0.83, and 0.96, respectively.

The activation/inhibition experiments were also carried out with CTP as substrate. Specificity effectors for CTP reduction are ATP or dATP, so in these experiments both the s- and a-sites were conceivably filled with ATP or dATP, respectively. It is obvious from Figure 1c that increasing concentrations of ATP initially stimulate activity and that high levels of ATP cause inhibition, plausibly by ATP also acting as substrate. We therefore used only results up to 3 mM ATP to calculate an approximate kcat of 1.1 s-1 (385±116 nmol/min/mg) with an apparent KL of 0.67±0.52 mM (Fig. 1c). Addition of dATP initially results in a stimulation of enzyme activity, conceivably when the s-site is filled with dATP, after which competing inhibition appears when dATP also binds to the ATP-cone (Fig. 1d). All in all, these results suggest that the ATP-cone in PcNrdD responds similarly to inhibition by dATP as has been observed before for several class I RNRs.

Binding of nucleotides to the ATP-cone

To confirm binding of ATP or dATP to the ATP-cone we used isothermal titration calorimetry (ITC) and microscale thermophoresis (MST) (Fig. 2). Substrate GTP and s-site effector dTTP were used to fill up the other nucleotide binding sites in PcNrdD. Binding of dATP occurred with a KD of 6 µM, whereas binding of ATP was 4-fold weaker with a KD of 26 µM (Fig. 2a-c). The number of bound dATP molecules observed with ITC was almost 2-fold higher for dATP compared to ATP, but both values were below 1. Nucleotide binding measured with MST confirmed our results for ATP (Fig. 2d) that had a KD of 25 µM, and dATP that with this method had a KD of 23 µM (Fig. 2e). However, in cryo-EM experiments, where higher nucleotide concentrations can be used, we show that the ATP-cone can bind two ATP or two dATP molecules (see below), as also expected based on its sequence. The KL and Ki values reported above for ATP-activation and dATP-inhibition of GTP reduction conceivably reflect the requirement for two bound nucleotide molecules to achieve these effects, whereas the KD measurements via ITC and MST are restricted by the component concentrations that can be used and plausibly only reflect the first nucleotide bound.

Binding of nucleotides to the ATP-cone.

Binding assay using ITC: Representative thermograms of binding of ATP to 50 µM PcNrdD (a); and binding of dATP to 200 µM PcNrdD (b); (c) thermodynamic parameters of ligand binding. All titrations were performed in the presence of 1 mM GTP substrate and 1 mM s-site effector dTTP at 20 °C. Binding of nucleotides to the ATP-cone using MST: (d) Binding of ATP, (e) binding of dATP. All MST binding experiments were performed in the presence of 1 mM s-site effector dTTP and 5 mM GTP substrate at room temperature. Control experiments were done for ITC, by titrating the nucleotide into the buffer (not shown).

It is noteworthy that the KD values for ATP and dATP do not differ by orders of magnitude, as has been found for several class I RNRs (Ando et al, 2016; Birgander et al, 2004; Ormö & Sjöberg, 1990; Rozman Grinberg et al., 2018a; Rozman Grinberg et al, 2018b; Torrents et al, 2006). Collectively these two binding methods instead suggested that ATP and dATP may interact at two distinct sites in the ATP-cone for the functional readout of dATP inhibition, as has earlier been observed in the transcriptional repressor NrdR (Rozman Grinberg et al., 2022). To test this hypothesis, apo-PcNrdD was incubated with ATP only, dATP only or a combination of ATP and dATP, then desalted, boiled, centrifuged and the nucleotide content of the supernatant analysed by HPLC. Figure 2 - figure supplement 1 shows that only dATP was bound when PcNrdD was incubated with a combination of ATP and dATP. This was also the case when incubations were performed in the presence of s-site effector and substrate, and also when the desalting was performed in the presence of s-site allosteric effector and substrate. In incubations with only ATP, s-site effector and substrate were needed during the entire work-up procedure for ATP to be retained by the protein. In conclusion, the binding experiments suggest that the PcNrdD ATP-cone can bind either ATP or dATP but not both simultaneously. Importantly, the results point to allosteric communication between the s-site, the active site and the a-site in the ATP-cone, as dTTP and GTP are required for any ATP to be retained. Both of these observations are consistent with the structural analyses shown below.

Modulation of the oligomeric state by ATP and dATP in the presence of allosteric effectors

Binding of dATP to the ATP-cone in all class I RNRs studied to date results in formation of higher oligomers unable to perform catalysis (Martínez-Carranza et al., 2020). We therefore next asked whether a similar mechanism would be valid for dATP-mediated inhibition of a class III RNR. To elucidate the oligomeric states of PcNrdD we used gas-phase electrophoretic mobility molecular analysis (GEMMA) (Kaufman et al, 1996), a method similar to electrospray mass spectrometry. It requires low protein and nucleotide concentrations as well as volatile buffers.

When loaded with ATP, PcNrdD was in a dimer-tetramer equilibrium shifted towards dimers (Fig. 3, left; Fig. 3 - figure supplement 1). Addition of the CTP substrate or the s-site effector dTTP resulted in a similar equilibrium. In contrast, the dimer-tetramer equilibrium of dATP-loaded PcNrdD was shifted towards tetramers, and this equilibrium was likewise not affected by addition of the CTP substrate (Fig. 3, right; Fig. 3 - figure supplement 1). Hence, dATP-inhibition of enzyme activity seems to work differently in PcNrdD than the clearcut and drastic change in oligomeric structure observed for class I RNRs, since both ATP-loaded and dATP-loaded forms of PcNrdD were in dimer-tetramer equilibria.

Oligomeric states of Prevotella copri NrdD in the presence of nucleotides determined by GEMMA.

(a) Apo-PcNrdD (2 µM) loaded with the activator ATP (50-100 µM) (black and red) in the presence of CTP (100 µM) as a substrate (cyan) or dTTP (100 µM) (orange) as the allosteric effector. (b) apo-PcNrdD (2 µM) loaded with the inhibitor dATP (50-100 µM) (black and red) in the presence of CTP (50 µM) (cyan) as the substrate. Each sample was scanned 5 times to increase signal-to-noise level. (c) Calculated numbers of monomers based on measured molecular weight and fractions of dimers versus tetramers after conversion of experiments (a) and (b) to mass concentrations.

Glycyl radical formation in presence of ATP or dATP

Next, we asked whether dATP inhibition was mediated by prevention of glycyl radical formation. Apo-PcNrdD was incubated with NrdG and S-adenosylmethionine in presence of ATP or dATP. Whereas reduction of substrate requires the presence of formate, the formation of a glycyl radical does not. We therefore incubated with or without formate as well as with or without substrate. Figure 4 shows that the glycyl radical was readily formed in the presence of dATP both with and without the CTP substrate, and the result was similar in the absence of formate (Figure 4 - figure supplement 1). Interestingly, under all conditions the glycyl radical content was higher in the presence of dATP compared to with ATP, and it was also more stable than in absence of any effector nucleotides. In addition, the shape of the doublet EPR signal was similar under all conditions studied. Collectively, these results show that the glycyl radical is readily formed in the presence of a dATP concentration that inhibits RNR activity.

Glycyl radical formation after 20 min in presence of formate and ATP±CTP or dATP±CTP.

Nucleotide concentrations were: 1.5 mM ATP, 1 mM dATP, and 1 mM CTP. Traces are arbitrarily moved to increase visibility and scaled to identical units (Y-axes).

Substrate binding to PcNrdD

To study substrate binding to activated and inhibited PcNrdD we used MST with labelled protein and GTP or CTP as substrate. The results were clear-cut in both sets of experiments (Fig. 5). ATP-loaded PcNrdD bound GTP in the presence of s-site effector dTTP (Fig. 5a), and it bound CTP in the presence of only ATP (Fig. 5b), whereas dATP-loaded PcNrdD had extremely low affinity to GTP and CTP under similar conditions (Fig 5c,d). These results show that the dATP-inhibition of enzyme activity in a class III RNR is mediated by inhibited binding of substrate.

Binding of substrate to ATP- or dATP-loaded PcNrdD.

Binding of GTP (a), and binding of CTP (b) to ATP-loaded PcNrdD. (c) Binding of GTP, and (d) binding of CTP to dATP-loaded PcNrdD. No additional nucleotides were present in CTP binding experiments, whereas binding of GTP was performed in the presence of 1 mM S-site effector dTTP. Fitted KDs are 2.8 ± 0.5 and 11 ± 2.4 mM for GTP and CTP binding respectively in the presence of ATP. In the presence of dATP fitted KDs were ≥ 2833 mM and ≥ 803 mM for GTP and CTP binding, respectively. Experiments were performed at room temperature.

Cryo-EM structure of the ATP-CTP bound PcNrdD dimer

2D classes of particles extracted from micrographs of PcNrdD incubated with effector ATP and substrate CTP indicated only dimers, with no appreciable amounts of tetramer. From a set of 570 730 particles we obtained a map from cryoSPARC at 3.0 Å. Application of deep learning local sharpening in DeepEMhancer (Sanchez-Garcia et al, 2021) improved the quality of the map, and we were able to build a reliable model for the great majority of the structure. As the ATP-cone density still did not permit reliable modelling, 3D classification was carried out on this particle set, enabling the identification of a subset of 291 231 particles that gave a volume at 3.2 Å with better-defined ATP-cones (Figure 6a). This map was used for refinement of the final model (Table 1; Table 1 - table supplement 1).

Structure of the PcNrdD dimer in complex with effector ATP and substrate CTP.

a) cryo-EM map with C1 symmetry showing the best-ordered ATP-cones after post-processing using DeepEMhancer. For the right-hand (active) monomer, the ATP-cone (residues 1-91) and the linking helix (92-104) are coloured olive, the linker and NxN flap region pale green, the glycyl radical loop red and the C-terminal extended region orange. The loop in the middle of the α/β barrel containing the radical initiator cysteine Cys416 at its tip is yellow. b) Overview of the PcNrdD dimer with ATP and CTP. The two monomers of the dimer are coloured in different shades of grey. The ATP and CTP molecules are shown as sticks. c) Schematic of the domain organisation of PcNrdD with the same colour scheme as a) and b); d) Closeup view of the binding of four ATP molecules to the dimer of ATP-cones in PcNrdD. The view is from the bottom of the molecule as seen in a) and b). Disordered loops are shown as dotted lines. e) Closeup view of the active site including the cryo-EM map for CTP. Residues within 4 Å of CTP are shown as sticks and polar interactions as dotted lines. f) Sequence logos of NrdD sequence motifs. A) ATP-cone plus downstream interaction partners, B) central parts of sequence and C) C-terminal parts. The numbering is from PcNrdD. Segments were selected to illustrate amino acids discussed in the text.

PcNrdD is a compact dimer (Fig. 6b). Each monomer consists of 739 residues, of which residues 1-91 constitute the ATP-cone, 92-110 a linker region, 111-122 a flap that folds over the substrate in the active site, 123-671 are the core domain with its 10-stranded α/β barrel fold typical of the RNR/glycyl radical enzyme (GRE) family, and 676-739 are the C-terminal glycyl radical domain (GRD) that contains a structural Zn site followed by the buried loop containing the radical glycine residue at position Gly711 and a C-terminal tail that extends across the surface of the protein (Figure 6b,c). The closest structural neighbours as identified by the DALI server (Holm, 2022) are the NrdDs from bacteriophage T4 (Logan et al., 1999) (1H7A, 2.5 Å root mean square deviation [rmsd] in Cα positions for 535 residues, 27% sequence identity) and T. maritima (Aurelius et al., 2015; Wei et al., 2014) (4U3E, 3.0 Å rmsd for 508 residues, 16% sequence identity; Figure 6 - figure supplement 1). The linker, flap and GRD are well-ordered in one of the monomers (dark blue in Fig. 6) and substrate CTP is bound (Figure 6 - figure supplement 2a,b), while in the other monomer this region is less well-ordered and the active site is empty (Figure 6 - figure supplement 2c,d). We will refer to these as the “active” and “inactive” monomers respectively.

At the concentration of ATP used (1 mM), ATP is not observed to bind to the s-sites near the dimer interface. The ATP-cones containing the a-site are flexible, but the reconstruction allowed the modelling of two ATP molecules per ATP-cone and many of the most important side chains. The ATP molecules are bound similarly to the twin dATP molecules seen in the ATP-cones of P. aeruginosa NrdA (PaNrdA) and L. blandensis NrdB (LbNrdB) (Johansson et al, 2016; Rozman Grinberg et al., 2018a), and to dATP bound to PcNrdD itself (see below). However, the ATP-cones in PcNrdD exhibit a hitherto unobserved relative orientation in which the four triphosphate groups are projected towards each other. (Fig. 6b,d). The large negative charge is compensated for by the proximity of as many as 12 Arg and Lys side chains from both monomers. The ATP-cones interact by contacts involving the “roof” of the domain (residues 1-18). Arg11 of each ATP-cone reaches over to interact with the α-phosphate group of one of the ATP molecules in the other ATP-cone, and a salt bridge is formed between Asp9 at the end of the lid and Arg91 at the end of the last helix of the other ATP-cone. Strikingly, the dimer of ATP-cones is slightly offset from the dimer axis of the core protein, such that the ATP-cone of the inactive monomer approaches the core domain of the active monomer, while the ATP-cone of the active monomer is more distant from the core domain of the inactive one (Fig. 6a,b). However, the closest atom of any ATP molecule is over 30 Å from the substrate CTP in the active site.

The glycyl radical site is found at Gly711, at the tip of a loop projected from the C-terminal region of the enzyme into the barrel, where it approaches the radical initiator cysteine Cys416 at the tip of a loop stretching through the barrel from the other side (Fig. 7). The second cysteine necessary for the radical mechanism of PcNrdD is Cys171 on the first β-strand of the barrel (Fig. 6e). Consistent with the fact that PcNrdD belongs to the group of anaerobic RNRs that use formate as overall reductant (Burnim et al, 2022a; Mulliez et al, 1995; Wei et al., 2014), there is no third cysteine residue on the sixth β-strand, and its place is taken by Asn435. As usual for GREs, Gly711 is completely buried within the barrel, with no solvent-accessible surface area. The base, phosphate and most of the ribose of substrate CTP are well-defined but the density is weaker around the 5’ C-atom (Fig. 6e).

The intricate network of interactions between the NxN flap, C-terminal GRD, loop 2 and substrate in the ATP-CTP complex of PcNrdD.

The NxN flap is coloured light green, the GRD orange, loop 2 dark red and the Cys radical loop yellow. The Gly radical loop is semi-transparent for clarity. The CA atom of Gly711 is indicated by an orange sphere. Important hydrogen bonds are shown as dotted lines.

After the glycyl radical loop emerges from the α/β barrel, it traverses the top of the barrel, forming a short helix from residues 723-730 and ending in an extended tail. This conformation of the C-terminus is very similar to the one seen for TmNrdD in PDB entry 4U3E (Wei et al., 2014). This tail makes few specific interactions with other residues in the core.

When compared to previously determined NrdD structures, the first common structural element is a long α-helix in the core domain stretching from Thr123 to Leu141. Between this and the ATP-cone are the linker (92-110), followed by the flap over the substrate in the active site (111-122). Two residues from the flap make H-bonds to CTP: Asn114 to the γ-phosphate group and Asn116 to the amino group of the cytosine base. We call this region the “NxN flap”, as the NxN motif (114-116) is one of the most highly conserved sequence motifs in the NrdD family (Fig. 6f). This is the first time such interactions have been observed in an NrdD, as the flap corresponds to a disordered segment of 17 residues in the structures of TmNrdD and T4NrdD. However, an AlphaFold2 (Jumper et al, 2021) model of TmNrdD (entry Q9WYL) suggests that this linker potentially has the same conformation in TmNrdD and possibly all other NrdDs, whether or not they have an ATP-cone. As seen previously in TmNrdD, the triphosphate moiety is recognised through the conserved HxHD motif (157-160), H-bonds to the main chain atoms of a loop (577-581) following the third-last strand of the α/β barrel and the dipole of the following helix (Fig. 6f).

Significantly, the NxN flap forms the nexus of a network of interactions (Fig. 7) between the substrate, the linker to the ATP-cone, the C-terminal GRD, and “loop 2” (residues 304-311) that is responsible for communicating the substrate specificity signal from the s-site at the dimer interface to the active site (Aurelius et al., 2015; Larsson et al, 2001; Logan et al., 1999). Important interactions linking the NxN flap to the C-terminus include H-bonds from Tyr712, which immediately follows the radical Gly711, to Glu581, which bridges to both the substrate and Ala115 in the flap, as well as from Arg719 in the GRD to the main chain carbonyl groups of Asn114 and Met117 in the flap. Tyr712 is highly conserved in the NrdD family (Fig. 6f and 7). This nexus is crucial for allosteric activity regulation, as will be discussed later.

As the NxN flap appears to close off the top of the active site, we used the web server Caver Web v1.1 (Stourac et al, 2019) to analyse substrate access to the active site in the ATP-bound form, and identified a tunnel ∼12 Å long leading from the surface of the protein to the active site that is 5.6 Å in diameter at its narrowest point (Figure 7 - figure supplement 1). Thus, in principle a fresh substrate could diffuse into the active site on each catalytic cycle.

Cryo-EM structure of the ATP-dTTP-GTP dimer

In order to probe the generality of the conformations observed in the ATP-CTP complex, we also solved the structure of PcNrdD in complex with ATP in the a-site, specificity effector dTTP in the s-site, and substrate GTP in the active site (c-site). Again, 2D classes show a predominantly dimeric form with no more than ∼10% tetramers. From a set of 437 886 particles selected from 2D classes with the best apparent density for the ATP-cone region, we obtained a volume at 2.40 Å resolution for the dimer with C1 symmetry that was further improved using DeepEMhancer (Table 1; Table 1 - table supplement 2). dTTP is clearly visible in the s-site (Figure 8 - figure supplement 1a) as is GTP in the active site (Figure 8 - figure supplement 2). A Mg2+ ion can be modelled in the s-site, coordinated by Glu290 and all three phosphate groups of dTTP. A β-hairpin loop between residues 185-192 that is disordered in the ATP-CTP structure in the absence of an s-site nucleotide is here ordered, due to an H-bond to the 2’-OH group of dTTP. Readout of the s-site nucleotide’s identity is achieved through an H-bond to Asn298 (Figure 8 - figure supplement 1).

Structure of PcNrdD in complex with a-site effector ATP, s-site effector dTTP and substrate GTP.

a) Volume for the PcNrdD dimer in complex with ATP, dTTP and GTP. The volume is contoured at a low level to emphasise the weak density for the more ordered ATP-cone domain in the active monomer. b) Ribbon diagram of the ATP-dTTP-GTP complex coloured as in Figure 6b, except that the most N-terminal helix (olive) now extends from residues 85-104. c) Zoom in on panel b) to illustrate the highly asymmetrical loop 2 conformations in the active and inactive monomers. Loop 2 in the inactive monomer (light blue) is coloured wheat, while in the active monomer (dark blue) it is purple. The conformation of loop 2 that forms a complementary cradle for the guanosine base of the substrate GTP in the active monomer induces a conformation of the other loop 2 that precludes substrate binding. d) Severe steric clash of loop 2 of the inactive monomer with the substrate and NxN flap. The view is rotated approximately 180° from panels a-c. Loop 2 from the inactive monomer (wheat) is superimposed on the active site of the active monomer (dark blue). The molecular surface of the loop is shown to emphasise that this conformation is incompatible with an ordered NxN flap (light green) and substrate binding. The Cys radical loop is yellow and Cys416 is marked by a sphere. The Gly radical loop is omitted for clarity.

Like the ATP-CTP complex, the ATP-dTTP-GTP complex can also be partitioned into an active and an inactive monomer. The ATP-cone region is very flexible, but the last helix of the ATP-cone in the active monomer can be traced back to residue 85 (Fig. 8a,b). The helix is not kinked at residue 91 as in the ATP-CTP dimer but forms an uninterrupted helix between residues 85-104. The volume for the ATP-cone region lies entirely on one side of the PcNrdD dimer axis, as is also apparent in some of the 2D classes (Figure 8 - figure supplement 3). The ATP-cone cannot be modelled in detail, but if the volume is contoured at very low level, there is almost enough density for one ATP-cone. Taken together, this suggests that there is one partially disordered ATP-cone and one highly flexible in the ATP-dTTP-GTP complex. In the active monomer, the NxN flap and entire C-terminus are ordered while in the inactive monomer, the entire assembly of linker and NxN flap up to residue 120 and the entire GRD are so flexible that they cannot be seen in the cryo-EM volumes. We expect that the GRD behaves largely as a rigid body whose displacements relative to the core domain become larger in the dATP-bound form.

Intriguingly, binding of dTTP to the s-sites at the dimer interface induces two distinct conformations of loop 2 (Fig. 8c). In the active monomer, loop 2 is projected towards the aforementioned network of interactions between NxN flap, GRD and substrate, packing against the flap, though no polar interactions are apparent. However, loop 2 also makes extensive interactions with its equivalent in the inactive monomer, which forces the latter into a conformation where it would sterically clash with an ordered NxN flap (Fig. 8d). Furthermore, Phe412, which stacks on the substrate base in the active monomer, and Arg305, which is projected from the substrate-distal side of loop 2, sterically prevent the substrate binding in the inactive monomer. These may be important contributing factors to the disorder of the NxN flap and GRD. Finally, Gln309, which interacts with the substrate in the active monomer (Figure 8 - figure supplement 2), is oriented away from the substrate through its interactions with the “active” loop 2.

Cryo-EM structure of the ATP-dGTP dimer

As a third insight into the ATP-bound forms of PcNrdD, we solved the structure of PcNrdD in complex with specificity effector dGTP in the s-site and ATP at a concentration where it would both act as allosteric effector at the a-site and as the cognate substrate for dGTP (Table 1; Table 1 - table supplement 3). Density for dGTP is clearly visible in the s-site (Figure 8 - figure supplement 1d) and for ATP in the active site (Figure 8 - figure supplement 2). Again, the two monomers can be divided into an active one and an inactive one. The conformations of loop 2 are once again asymmetric (Figure 8 - figure supplement 4). The conformation in the active monomer is well-defined, while loop 2 in the inactive monomer is more flexible than in the ATP-dTTP-GTP complex, but still compatible with the hypothesis that the asymmetric arrangement of loop 2 is responsible for disorder of the NxN-GRD-substrate network in the inactive monomer.

Cryo-EM structure of the dATP-bound dimer

In micrographs of PcNrdD samples in the presence of 0.5 mM dATP, we observed an approximately 1:1 mixture of dimeric and tetrameric particles, consistent with results from GEMMA. To resolve whether dATP inhibition was linked to oligomerisation, we resolved the structures of each of these oligomers separately. We made a reconstruction of the dATP-bound PcNrdD dimer from 1 009 021 particles to 2.7 Å resolution that was further improved by DeepEMhancer postprocessing. The core domain is very similar to that of the ATP-CTP complex, with an rmsd in Cα positions of 1.2 Å for 1089 Cα atoms considering each dimer as a whole (Table 1; Table 1 - table supplement 4). dATP is clearly visible in the specificity site (Figure 8 - figure supplement 1b).

Strikingly, in contrast to the ATP complexes, the entire GRD is disordered in both monomers of the dATP-bound dimer (Figure 8 - figure supplement 5). No structure is visible after residue 676 at the end of the last strand of the barrel. This suggests that inhibition by binding of dATP to the ATP-cone is coupled to disordering of the C-terminal domain. The NxN flap that covers the top of the substrate in the ATP/CTP-bound form is also disordered, as are the ATP-cones themselves.

Cryo-EM structure of the dATP-bound tetramer

From the same micrographs that revealed the dATP-bound dimers we also resolved dATP-bound tetramers. We obtained a map from cryoSPARC at 2.8 Å, which was again improved with DeepEMhancer (Fig. 9a). Two dATP molecules are bound to each ATP-cone and one at each of the specificity sites, giving a total of 12 visible dATPs per tetramer. The tetramer reveals an arrangement of two dimers (chains A/B and C/D respectively) in which the ATP-cones are all well-ordered and one ATP-cone from each pair mediates interactions between the dimers (Fig. 9b). The ATP-cones have an almost domain-swapped arrangement within each dimer relative to the core domains. This oligomeric arrangement has not previously been observed within the RNR family.

Structure of the dATP-bound PcNrdD tetramer.

a) cryo-EM reconstruction coloured by chain: dimer 1 light blue and pale blue; dimer 2 light green and pale green. The four ATP-cones are at the bottom of the figure. b) View rotated by 90° around a horizontal axis relative to a) and thus viewed from the bottom, showing the ATP-cones. c-d) Cartoon representation of the tetramer from the same angles as panels a) and b) respectively and with the same chain colouring. The two-fold symmetry axis relating the two dimers of the tetramer is shown as a yellow line. The 12 dATP molecules in the tetramer are shown as sticks. The dimer axis of the left-hand dimer’s core domains is shown as a dark blue line and the dimer axis of the ATP-cone pair as a black line. c) Details of the interaction of dATP molecules at the allosteric activity site in the ATP-cone. d) Close-up of the interaction area between the roof of the ATP-cone of one dimer and the core domain of the other dimer (marked with a black box in panel d). The most significant residues are labelled.

dATP binds to the specificity site as in the dimeric dATP-bound form (Figure 8 - figure supplement 1c). The two dATP molecules bound to each ATP-cone do so very similarly to the dATP-loaded ATP-cones of the RNR class I active site subunit from PaNrdA (PDB ID 5IM3) and the class I radical generating subunit from LbNrdB (5OLK) to which the PcNrdD ATP-cones have 29% and 33% sequence identity respectively (Figure 9 - figure supplement 1). The DALI server reveals an rmsd in Cα positions of 1.9 Å and 1.7 Å respectively. The next most similar ATP-cone is that of the RNR transcriptional regulator NrdR from Streptomyces coelicolor in its dodecameric form (Rozman Grinberg et al., 2022) binding two ATP molecules, with 82 aligned residues, rmsd 2.4 Å and sequence identity 17%. Structural similarity to other ATP-cones is listed in Figure 9 - figure supplement 1.

The two dATP molecules bind with their negatively charged triphosphate moieties oriented towards each other (Fig. 9c). No Mg2+ ion is observed between the triphosphate tails, but this may be an artefact of the cryo-EM method, as the PaNrdA and LbNrdB crystal structures suggest that this ion is necessary for charge neutralisation. As in the ATP-bound form, further charge compensation is achieved by extensive coordination by, or proximity of, basic side chains, e.g. Lys7, Arg8, Arg87 and Arg91. As previously observed in PaNrdA and LbNrdB, the adenosine base of the “inner” dATP (dATP1) fits into a pocket under the β-hairpin roof of the ATP-cone and the base of the “outer” nucleotide (dATP2) is sandwiched between an aromatic residue (Tyr83) and the side of the first α-helix in the domain. No direct contacts are seen between the dATP molecules bound to the ATP-cone and residues in the core domain.

Figure 9 - figure supplement 2 shows the differences in binding modes of dATP in the dATP-bound tetramer and ATP in the ATP-CTP dimer. The high mobility of the ATP-cone in the latter precludes a detailed analysis. However, it is clear that the triphosphate tails have very different orientations. In the ATP complex the tail of ATP2 is oriented away from the C-terminal helix towards the base of ATP1, such that the β-phosphate moiety may even H-bond to the base of ATP1. The side chain of Arg87 fills the space vacated by the γ-phosphate of ATP2. Instead of H-bonding to the γ-phosphate of ATP2, Arg91 swings away and interacts with the equivalent moiety of ATP2 in the ATP-cone of the other chain. The structures suggest that the additional 2’-OH group in ATP may cause conformational changes in the ribose ring that propagate to the phosphate tail, but further analysis requires improved resolution.

As in PaNrdA and LbNrdB, the ATP-cones in the dATP-bound tetramer self-interact through a pair of α-helices at the C-terminal end of the ATP-cone. Remarkably, the pair of ATP-cones is highly asymmetrically positioned with respect to the 2-fold axis of its core dimer, being displaced as a rigid body towards the other dimer (Fig. 9b). Both pairs of ATP-cones are displaced in the same way, giving the complex overall C2 symmetry around an axis bisecting the pair of dimers. Consequently, the 2-fold axes of the ATP-cones and the core dimer are not aligned (Fig. 9b). The asymmetric arrangement is associated with two distinct conformations of the linker 93-103 between the fourth helix of the ATP-cone (Figure 9 - figure supplement 3a,b) and the core domain. In chains A and C, a small kink leads from helix 4 into a short helical segment that packs antiparallel to helix 122-142 in the core domain. In contrast, in chains B and D, the short helix is parallel to helix 122-142 and is joined to helix 4 at a 45° angle. The rest of the polypeptide between residues 103-121, including the NxN flap, is disordered in both linkers.

Interestingly, all contacts between the two dimers of the tetramer are mediated by interactions between one of the ATP-cones in one dimer (B or D) and one of the core domains of the other dimer (C or A respectively). Figure 9d shows the interactions of the roof of the ATP-cone with two outer helices from the C-terminal half of the barrel domain (residues 458-485 and 550-569, preceding the sixth and seventh strands). The interactions are mostly hydrophobic but are reinforced by several H-bonds, e.g. between Gln3D-Gln458A, Ser53D–Gln458A, Arg11D-Asp468A, the main chain amide of Ile12D and Tyr557A. The amount of buried surface area is small: 717 Å2, or around 1% of the total surface area of each dimer. For comparison, the monomer-monomer interactions within each dimer bury 11.7% of the total area. The residues on the core domain of PcNrdD involved in interactions with the ATP-cone of the other dimer do not show high sequence conservation, even when the alignments are restricted to the most similar sequences.

Cryo-EM structures of PcNrdD-dATP complexes produced in the presence of CTP

To investigate whether substrate had any allosteric effect on the conformations of the ATP-cone, flap and GRD, we determined separate cryo-EM structures of tetrameric and dimeric PcNrdD, from the same grid, in the presence of 0.5 mM dATP and 0.5 mM CTP, achieving a resolution of 2.6 Å from 1 105 348 particles before post-processing for the tetramer and 2.6 Å from 1 132 695 particles for the dimer (Table 1; Table 1 - table supplement 5). Both forms were almost identical to those seen in the presence of dATP alone, and no CTP could be seen bound to the active site (Figure 6 - figure supplement 3). This strongly suggests that substrate binding is dependent on ordering of the flap region by binding of ATP to the ATP-cone.

Hydrogen-deuterium exchange mass spectrometry (HDX-MS)

To validate our structural and biochemical hypotheses, we further examined the degree of order of the different structural elements of PcNrdD in solution using HDX-MS. This method measures the degree of protection of peptides from H/D exchange in a deuterated solution, which reports on how exposed they are. Three different samples were analysed: apo-PcNrdD and the complexes with dATP-CTP and ATP-CTP.

The apo form shows high deuterium uptake over the entire ATP-cone, linker and NxN flap, from residues 1-117 (Figure 10a). Analysis of differences in uptake in the N-terminal region between the apo protein and the ATP or dATP complexes clearly confirms binding of both nucleotides to the ATP-cone (Figure 10b). Peptides spanning residues 2-84 exhibit clear protection from HDX in both nucleotide-bound states. Furthermore, bimodal deuterium uptake (James et al. 2022) was observed for N-terminal peptides spanning residues 2-23 and 83-99. For peptides very close to the N-terminus (2-23), the apo form only exhibits one high uptake (dynamic) conformational state while for the two ligand bound states there was a clear bimodal behaviour (EX1) with one high uptake and one protected state of an approximately equal size (Figure 10c and Figure 10 - figure supplement 1). Possible interpretations are that the bimodality is due to a mixture of occupied and empty ATP-cones, or to the mixture of dimers and tetramers.

Summary of the HDX-MS experiments. a) Heat map showing the deuterium uptake of the apo state for time points 30, 300 and 3000 s. Warm colours represent fast uptake and cold colours a low uptake. Both the N-terminal and the C-terminal regions have a high uptake already at 30s. For the N-terminal bimodal uptake, only the fast exchanging population (rightmost) is depicted. The influence of bimodal uptake and more detail can be seen in Figure 10 - figure supplements 1 and 2. b) Differential chiclet plot (Zhang et al. 2021) collapses the timepoints into a single maximum deuterium uptake difference for states apo-(ATP+CTP), apo-(dATP+CTP) and (dATP+CTP)-(ATP+CTP). Colours range from deep blue for a high degree of relative protection to deep red for high deprotection. Areas marked with dashed red boxes are peptides showing bimodal uptake, i.e. two conformational populations present simultaneously, and black dashed boxes denote areas with peptides having EX2 exchange (only one conformational population in the observed timespan). c) Deuterium uptake plots from dashed areas in b). For peptides 2-23 and 83-99, when there are two lines with the same colour, the top traces are for the fast-exchanging population and the bottom for a less dynamic/more protected population. Note that for peptide 83-99 the apo state has two populations at all timepoints, while the ATP and dATP converge at 3000s to one population (with the dATP state as most protected). For peptides 295-312 and 693-713 the apo curve has been removed for visual clarity.

Interestingly, peptides spanning amino acids 81-99 in the linker region have a bimodal uptake that is most pronounced in the apo state. For the nucleotide-bound forms there are also two populations where the dATP-CTP state converges into a single, more protected population at the longest time point, 3000s (Figure 10c). This suggests a mixture of states with flexible and locked linkers, consistent with the highly dynamic nature of residues 85-100 observed in the cryo-EM structures, where in the dATP bound tetramer the linker is ordered to residue ∼100, while in the dimer the ATP-cone and linker are highly dynamic and the first ordered structure is observed from residue 120.

The NxN flap at residues 101-117 has a high H/D exchange rate with both ligands and does not differ much between ATP and dATP, but the region 118-126 has a higher exchange rate in the dATP form, most likely because substrate is not bound to the active site.

Clear differences between the ATP and dATP complexes are seen in the region 178-204, which contains the β-hairpin of the s-site (Fig. 10c, panel 178-204) that is highly protected with dATP. This is consistent with the structures that show no ATP bound at the s-site but full occupancy of dATP. Peptides between 330-348, which are in contact with the β-hairpin, are also protected in the dATP complex. The region 295-312, i.e. loop 2, is deprotected in the dATP form relative to ATP. This is also in agreement with the cryo-EM structures, in which loop 2 is partially disordered in the dATP complex but ordered in the ATP complex in the monomer with bound CTP.

The HDX-MS results show a high degree of deuterium uptake in the GRD in all forms, indicating a high degree of mobility. This is consistent with the cryo-EM structures, in which the GRD overall has B-factors higher than the core domain (though the glycyl-radical containing loop has similar B-factors to the core, being buried). From the HDX-MS data it is clear that the GRD is highly dynamic, and while a trend of increased dynamics for some C-terminal peptides in the GRD can be seen for the dATP-CTP state, a detailed analysis is confounded by the limited time resolution and the complexity of the sample, which consists of a mixture of conformations.

Discussion

Ribonucleotide reductase is an essential enzyme for all free-living organisms. Its complex regulation at the levels of transcription, overall activity and substrate specificity makes it quite unique in Nature (Mathews, 2016, 2018). A fundamental understanding of this regulation could e.g. pave the way for better antibiotics. The ATP-cone is a small, genetically mobile domain frequently found at the N-terminus of RNR catalytic subunits of all major classes, as well as in the bacterial RNR-specific transcription factor NrdR (Rozman Grinberg et al., 2022). Its role in allosteric regulation of class I aerobic RNRs has been extensively studied. In the presence of dATP, the ATP-cone mediates the formation of a surprising variety of different oligomeric complexes that nevertheless have in common that they block the formation of an (observed or inferred) active complex between the catalytic and radical generating subunits that permits proton-coupled electron transfer on each catalytic cycle (reviewed in Martínez-Carranza et al., 2020).

Despite their catalytic subunits being evolutionarily related to those of class I, anaerobic class III RNRs have a fundamentally different activation mechanism that is likely common to all glycyl radical enzymes (Backman et al., 2017; Lundahl et al, 2022). Once generated on the catalytic subunit by encounter of the reductase with its radical SAM family activase, the glycyl radical can catalyse hundreds of cycles of turnover before being exhausted (Torrents et al., 2001). Allosteric inhibition could thus in principle proceed by two mechanisms: by blocking the initial encounter of the reductase and activase, or by preventing subsequent transfer of the radical to the substrate. The first of these hypotheses is intuitively more similar to the steric blocking mechanisms encountered in class I RNRs; the latter is more challenging to understand, as the glycine and cysteine are buried in close proximity in the active site.

In this work we present the first biochemical, biophysical and structural characterisation of an anaerobic, class III RNR containing an ATP-cone, providing the first picture of the structural basis for allosteric activity regulation in this large family. We show that the ATP-cone can bind two molecules of ATP or dATP, but not both simultaneously. HDX-MS confirms that binding of either nucleotide reduces the exposure of the ATP-cone to deuterium exchange, as well as confirming dATP binding to the s-site. As expected, the presence of an ATP-cone confers inhibition of enzyme activity at high dATP concentrations. GEMMA experiments show that PcNrdD exists in a mixture of dimeric and tetrameric states under most experimental conditions studied, with the equilibrium shifted towards dimers in the presence of ATP. Addition of inhibiting concentrations of dATP shifts the equilibrium towards tetramers. The equilibria are not affected by substrate or specificity effector. EPR spectroscopy clearly shows that the glycyl radical is formed even in the presence of dATP. Together, these results show that dATP inhibition in class III RNRs does not occur by blocking the initial encounter of reductase and activase. Instead, our nucleotide binding results show that dATP inhibition prevents binding of substrates.

The seven cryo-EM structures of PcNrdD in a variety of nucleotide complexes presented here suggest a mechanism in which binding of ATP or dATP modulates activity by affecting the dynamics of a tightly knit network of structural elements including the critical glycyl radical domain in the C-terminal region of the enzyme. A summary of the conformational states is shown in Figure 11. In the presence of ATP and substrate CTP, the GRD is fully ordered in one monomer and Gly711 is found at its expected position proximal to Cys416, to which it will deliver its radical to initiate each catalytic cycle and accept it back at the end of the cycle. The CTP substrate is bound in this monomer and a conserved but previously unobserved flap containing two Asn residues (NxN flap) forms over the top of the substrate, making H-bonding interactions with it. Loop 2, which mediates between the substrate specificity site and the active site, is fully ordered. In contrast, in the other monomer, no substrate is bound, loop 2 is disordered and the flap and C-terminus are more dynamic. At the same time, the ATP-cones themselves are relatively flexible. This is not unexpected for this type of ATP-cone, which binds two nucleotides and was first observed in the class I RNR PaNrdA (Johansson et al., 2016). SAXS studies showed that the ATP-cone in PaNrdA is flexible in the presence of ATP, while in its dATP-bound form it forms a symmetric interaction with an ATP-cone on another dimer, causing the enzyme to tetramerise. A similar behaviour was seen even when this type of ATP-cone was evolutionarily “transplanted” to the N-terminus of the radical-generating subunit, as seen in LbNrdB (Rozman Grinberg et al., 2018a).

Schematic showing the different structural states of PcNrdD described in this work. The GRD is coloured orange where present. Inactive and active monomers are labelled I and A respectively. The a-site in the ATP-cone is labelled with “a” on a green background, the s-site at the dimer interface with “s” on a red background and the active site with “c” on an orange background. Labels are only present when a nucleotide is observed in the structure. The ligands bound are listed, coloured according to the sites they bind to. In the dATP/dATP-CTP tetramer, one of the inactive monomers and its s-site are hidden behind the dimer in the foreground.

By careful isolation of a subset of particles, we were able to characterise an unusual dimer of ATP-cones in the ATP-CTP complex, where four ATP molecules bind with all their triphosphate tails pointing towards each other. Such an orientation is unprecedented for RNRs, but such tail-to-tail orientations, compensated by extensive positive charges, are found in e.g. adenylate kinases (Berry, 1994). The relevance of this arrangement remains to be elucidated. Interestingly, when dTTP was added to the s-site in the ATP-dTTP-CTP complex, the ATP-cones became even more dynamic and could only be observed in one monomer of the dimer as diffuse density. This behaviour was reproduced with dGTP in the s-site. Both complexes with s-site effector are functionally more relevant forms than the ATP-CTP complex, as in the cell, a form with empty s-site is unlikely to exist. The partitioning of monomers into active and inactive is amplified by the presence of s-site effector (dTTP or dGTP) in both complexes with ATP. Only one of the active sites is occupied by substrate, with ordered NxN flap and C-terminal GRD. This is correlated with two distinct conformations of loop 2 in the two monomers. In the inactive monomer, the NxN flap and GRD are disordered, as loop 2 blocks their active conformation and concomitantly substrate binding. These results point to only one of the PcNrdD monomers being active at a given time and also to a significant cross-talk between the a-site, s-site and active site. This is corroborated by the finding that an s-site effector is necessary for retention of ATP by the ATP-cone when PcNrdD is first incubated with these nucleotides then purified without them.

In the presence of inhibitory concentrations of dATP we were able to isolate structures of both dimers and tetramers from the same cryo-EM sample, in agreement with the GEMMA results that show a shift towards the tetrameric form. In both forms, the C-terminal GRD is highly mobile in all monomers such that it cannot be seen in the cryo-EM reconstructions. An exposed glycyl radical is incompatible with its stability, so it is possible that the dynamics of the GRD are exaggerated somewhat by the vitrification process. Consistent with this, the HDX-MS results show a small but noticeable deprotection of peptides from the GRD in the presence of dATP-CTP compared to ATP-CTP.

In the tetramers, pairs of ATP-cones asymmetrically disposed relative to the 2-fold axes of their respective PcNrdD dimers associate through the “roof” of the “inner” dATP binding site to the core domain of the other dimer. While this allowed us to build complete models for the ATP-cones and analyse dATP binding, the role of tetramerisation in the inhibition mechanism is unclear. The tetrameric structures appear to be facilitated by a dATP-induced “stiffening” of the pair of ATP-cones, similar to what was seen in PaNrdA and LbNrdB, and their interaction with a second dimer of PcNrdD, but the interaction area is small (1% of the total surface area) and the residues involved are not highly conserved in NrdDs, even from the most similar sequences. Therefore, further studies of NrdDs containing ATP-cones are required in order to determine whether oligomerization is a general phenomenon in the family. Furthermore, the GRD is mobile in all complexes with dATP in the a-site, whether dimeric or tetrameric. Since both the ATP-cone and the GRD are disordered in the dATP-bound dimers, the exact molecular mechanism by which dATP induces disorder of the GRD remains somewhat elusive.

Taken together, our results show a clear correlation between ATP or dATP binding to the ATP-cone and activity or inhibition respectively. This involves a previously uncharacterised but highly conserved structural element, the NxN flap, that folds over the substrate and furthermore acts as a structural bridge between the ATP-cone, the active site and the C-terminal GRD. An increase in dynamics of these elements prevents substrate binding and thus transfer of the radical from the GRD to the substrate. Interestingly, a similar flap is formed over the top of the active site in the E. coli active complex, but it is formed by part of the C-terminal region of the separate radical generating subunit (Kang, 2020). Thus, a locking of the substrate in the active site appears to be a conserved feature of active class I and class III RNRs. However, in class I RNR there is no tunnel allowing substrate access to the active site in the locked state. This is consistent with the fact that the class I subunits dissociate on each catalytic cycle, while in class III the glycyl radical can catalyse many substrate reductions before having to be regenerated. It remains enigmatic exactly how the tiny chemical difference between dATP and ATP (the 2’-OH group) results in such major conformational changes, as in the ATP-CTP complex, where the ATP-cones are visible, the a-site in the ATP-cone is separated from the NxN flap by at least 30 Å and in the most biologically relevant complexes with additional s-site effector, the ATP-cone is very disordered. Nevertheless, the present results give first insights into allosteric activity regulation in anaerobic RNRs and add yet another aspect to the surprisingly wide range of allosteric conformational changes that can be mediated by the binding of the two highly similar nucleotides ATP and dATP to a small, evolutionarily mobile protein domain.

Materials and Methods

Cloning of nrdD and nrdG from P. copri

The genes encoding the reductase (NrdD) and activase (NrdG) proteins from P. copri were synthesised by GenScript with codon optimization for E. coli and subcloned into the pBG102 plasmid (pET27 derivative) (Center for Structural Biology, Vanderbilt University) between the BamHI and EcoRI restriction sites to produce His6-SUMO-NrdD and His6-SUMO-NrdG protein constructs.

Overexpression and purification of PcNrdD and PcNrdG

Plasmids containing the nrdD gene or the nrdG gene were transformed into E. coli BL21 (DE3) star and E. coli BL21 (DE3) competent cells, respectively. Cells (6 L) were grown at 37 °C in Luria Broth supplemented with kanamycin (50 µg/mL) to an A600 of 1.2. Protein expression was then induced with 0.5 mM IPTG and incubation was extended overnight at 20 °C. After centrifugation, pellets were suspended in 60 mL of lysis buffer 1a (50 mM Tris-HCl pH 8, 500 mM KCl, 0.5 mM TCEP) for PcNrdD and lysis buffer 2a (50 mM Tris-HCl pH 8, 500 mM NaCl, 0.5 mM TCEP) for PcNrdG, supplemented with lysozyme (0.5 mg/mL), and disrupted by sonication. Cell debris was removed by ultracentrifugation at 210 000 x g for 1 h at 4 °C. The supernatant was then loaded on an immobilised metal affinity Ni-NTA column (HisTrap 5 mL; Cytiva) equilibrated with buffer 1b (50 mM Tris-HCl, pH 8, 300 mM KCl) or buffer 2b (50 mM Tris-HCl, pH 8, 300 mM NaCl), supplemented with 50 mM imidazole as appropriate. The column was washed extensively with the corresponding buffer containing 50 mM imidazole and the His6-SUMO tagged proteins were eluted using 300 mM imidazole. The proteins were collected and dialyzed overnight at 4 °C against buffer 1b or 2b supplemented with 1 mM 1,4-dithiothreitol (DTT) in the presence of the PreScission Protease (150 µM) to cleave the affinity-solubility tag. The GST–3C-protease (PreScission) was expressed using pGEX-2T recombinant plasmids. After induction at 25 °C with 0.1 mM IPTG for 20 h, the PreScission Protease was purified using glutathione–Sepharose chromatography. After the dialysis, the cleaved PcNrdD and PcNrdG proteins were centrifuged at 4 °C for 10 min and loaded onto a HisTrap column equilibrated with buffer 1b or 2b, supplemented with 50 mM imidazole and the flow-through containing the cleaved protein was collected. Following cleavage of the His6-SUMO tag, the proteins incorporated a non-native N-terminal Gly-Pro-Gly-Ser sequence. The purified preparations are called as-purified PcNrdD and as-purified PcNrdG, respectively.

A fraction of as-purified PcNrdD was precipitated by addition of HCl, then centrifuged at 15 000 x g for 5 min. UV-visible spectra of the supernatant were recorded before and after protein precipitation to estimate the amount of nucleotide contamination. PcNrdD preparations with ≤5% nucleotide contamination are referred to as apo-PcNrdD.

Fractions containing apo-PcNrdD and as-purified PcNrdG were concentrated and loaded on a gel filtration column (Hiload 16/60 Superdex S200; Cytiva) in 25 mM Tris-HCl, pH 8, 250 mM NaCl, and 5 mM DTT. The purified protein was concentrated to 20 mg/mL with an Amicon ultrafiltration device (100 kDa cutoff for NrdD and 10 kDa cutoff for NrdG; Millipore), frozen in liquid nitrogen, and stored at −80 °C.

[Fe-S] cluster reconstitution and purification of holo-PcNrdG

The reconstitution of the [4Fe-4S] cluster and purification of PcNrdG containing an iron-sulfur cluster (holo-PcNrdG) were performed under strict anaerobic conditions in an Mbraun glove box kept at 18°C and containing less than 0.5 ppm O2. PcNrdG was treated with 5 mM DTT for 10 min then incubated for 3 h with a fivefold molar excess of ferrous ammonium sulphate and L-cysteine in the presence of 2 μM E. coli cysteine desulfurase CsdA. The holo-PcNrdG was then loaded onto a Superdex 200 10/300 gel filtration column (Cytiva) equilibrated in 25 mM Tris-HCl, 250 mM NaCl, and 5 mM DTT (Figure 1 - figure supplement 2). The peak containing the soluble holo-PcNrdG was then concentrated to 10 mg/mL on a Vivaspin concentrator (10 kDa cutoff).

In vitro enzymatic assay and HPLC analysis

Activity assays were performed under strict anaerobic conditions inside a glovebox in 25 mM Tris pH 8, 50 mM KCl, 10 mM DTT in a volume of 100 μL. A standard pre-reaction mixture contained 1 μM apo-PcNrdD, 2.5 μM holo-PcNrdG, 500 µM S-adenosyl-methionine (SAM), 5 mM MgCl2, 1 mM GTP (or 1 mM CTP), and different allosteric effectors dTTP (1 mM), ATP (0 - 5 mM) or dATP (0 - 1 mM). The pre-reaction mixture was incubated for 5 min at 37 °C before adding simultaneously 10 mM sodium formate and sodium dithionite (12.5 molar excess). When GTP was used as substrate, dTTP was used as s-site effector, and ATP or dATP were titrated to the ATP-cone (a-site). When CTP was used as substrate the s-site was empty at the beginning of the titration and ATP or dATP can plausibly bind both to the s-site and to the a-site in these experiments.

The reaction was incubated for 10 min at 37 °C and stopped by the addition of 2.5 μL of 3 M formic acid. Product formation was analysed by HPLC using an Agilent ZORBAX RR StableBond (C18, 4.6 x 150 mm, 3.5 µm pore size) equilibrated with buffer A (10% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Samples of 10 μL were injected and eluted at 0.5 mL/min with a step isocratic flow of 40%-100% buffer B (30% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Compound identification and product quantification were performed by external calibration and NrdD specific activity was determined.

From a direct plot of activity versus concentration of effector, the KL values (the concentration of an allosteric effector that gives half maximal enzyme activity) for binding of effectors to the a-site, were calculated in KaleidaGraph using the equation:

Ki for non-competitive dATP inhibition of GTP reduction was calculated in KaleidaGraph using the equation: and KL and Ki for dATP-dependent activation and non-competitive dATP inhibition of CTP reduction were calculated in KaleidaGraph using the equation:

Isothermal titration calorimetry experiments

ITC experiments were carried out at 20°C on a MicroCal ITC-200 system (Malvern Instruments Ltd) in ITC buffer containing 25 mM HEPES, pH 7.5, 100 mM KCl, 10 mM MgCl2 and 0.5 mM TCEP with a stirring speed of 1000 rpm. An initial injection volume was 0.6 µl over a duration of 1.2 s. ApoPcNrdD was prepared in the ITC buffer using SEC, followed by the addition of 1 mM dTTP and 1 mM GTP and incubation at room temperature for 5 minutes prior to loading to the cell. ATP and dATP ligands were directly diluted in the ITC buffer. Titration of ligand into buffer was also performed to control for heats of dilution and/or buffer mismatch. For dATP binding analysis, the concentration of PcNrdD in the cell was 50 to 100 µM and dATP concentration in the syringe 0.6 to 1 mM. For ATP titration into PcNrdD, protein concentration in the cell was 200 µM and ATP concentration in the syringe was 0.6 to 1.5 mM. All subsequent injection volumes were 1.7–2.5 µl over 3.4–5 s with a spacing of 160 s between the injections. Data for the initial injection were not considered. The data were analysed using the one set of sites model of the MicroCal ITC-200 analysis software (Malvern Panalytical). Standard deviations in thermodynamic parameters, N and KD were estimated from the fits of three different titrations.

Microscale thermophoresis

Both binding of nucleotides to the ATP-cone in PcNrdD and binding of substrates CTP and GTP to PcNrdD were assessed using MST. PcNrdD was labelled using Monolith Protein Labeling Kit RED-NHS 2nd generation according to the manufacturer’s protocol. MST was performed using the Monolith NT.115 instrument (NanoTemper Technologies GmbH) at room temperature. Binding of GTP and CTP was assayed in MST buffer containing 40 mM Hepes pH 7, 50 mM KCl, 50 and 100 mM MgCl2, 5 mM DTT, 0.1% Tween-20, 1 mM dTTP (only for GTP binding) and either 5 mM ATP or 1 mM dATP. The 16 reaction tubes were prepared by the addition of 2 μl MST buffer (concentrated 5 times), 5 μl GTP or CTP of the desired concentration and 1 μl RED dye-labelled NrdD in a total volume of 10 μl. Final NrdD concentration in the binding reaction was 11 nM while the binding partner concentrations were between 70 mM and 2 μM for CTP and 50 mM and 1.5 μM for GTP. Binding of ATP and dATP to the ATP-cone was assayed in MST buffer containing 25 mM Hepes pH 7.5, 100 mM KCl, 10 mM MgCl2, 0.5 mM TCEP, 0.1% Tween-20, 1 mM dTTP and 5 mM GTP. A 16-step dilution series of the binding partners was prepared by adding 5 μl buffer to 15 tubes. ATP or dATP was added to the first tube to the final concentration of 4 mM and 5 μl was transferred to the second tube and mixed by pipetting (1:1 dilution series). To each tube of the dilution series 5 μl of PcNrdD was added to reach a final concentration of 11 nM. The samples were transferred to Monolith™ NT.115 Series Premium Capillaries (NanoTemper Technologies GmbH), which were scanned using the MST instrument (100% excitation power, medium MST power). Obtained MST data was analysed and fitted using the MO. Affinity Analysis v2.3 software (NanoTemper Technologies) with default parameters. KD and standard deviation for GTP and CTP binding to PcNrdD in the presence of ATP and for ATP and dATP binding to the ATP-cone were calculated using fits from at least three individual titrations. KD values for CTP or GTP binding to PcNrdD in the presence of dATP were extremely high and could not be reliably determined since the titration curves did not reach a plateau and therefore are only estimates.

Nucleotide binding to the PcNrdD ATP-cone

PcNrdD was incubated with nucleotides at room temperature for 30 minutes and then desalted on a NAP-5 column. The desalted protein was boiled for 10 min, centrifuged, and the supernatant was loaded on a Agilent ZORBAX RR StableBond (C18, 4.6 x 150 mm, 3.5 µm pore size) HPLC column to evaluate the amount of nucleotides retained by the protein. Samples of 100 μL were injected on the column equilibrated with buffer A (10% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide) and eluted at 1 mL/min with a gradient of 40%-100% buffer B (30% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Compound identification and product quantification were performed by comparison with injected standards. Three different sets of experiments were performed, each set consisting of incubation with only ATP, only dATP, and a combination of ATP and dATP. In the first set of experiments 75 µM apo-NrdD was incubated with 1 mM ATP or 1 mM dATP, and when 1 mM each of ATP and dATP was used, dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer without nucleotides. In the second set of experiments 75 µM apo-NrdD in presence of 1 mM dTTP and 5 mM GTP was incubated with 3 mM ATP or 1 mM dATP, and when both nucleotides were combined dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer without nucleotides. In the third set of experiments 75 µM apo-NrdD in presence of 2 mM dTTP and 5 mM GTP was incubated with 3 mM ATP or 1 mM dATP, and when both nucleotides were combined dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer containing 0.1 mM dTTP and 1 mM GTP. It was possible to separate all phosphorylation levels of adenosine and deoxyadenosine nucleotides in the first set of experiments, but not in the second and third sets due to overlapping peaks of the s-site and substrate nucleotides (Figure 2 - figure supplement 1).

Gas-phase electrophoretic mobility molecular analysis

The GEMMA instrumental setup and general procedures were as described previously (Kaufman et al., 1996; Rofougaran et al, 2008). Apo-PcNrdD was equilibrated into a buffer containing 100 mM ammonium acetate pH 7.8, then applied onto a Sephadex G-25 chromatography column. In addition, 5 mM DTT was added to the PcNrdD protein solution to increase the protein stability. Prior to analysis, the protein sample was diluted to 2 µM in a buffer containing 100 mM ammonium acetate, pH 7.8, 0.005% (v/v) Tween 20, magnesium acetate (equimolar to the total nucleotide concentration) and the corresponding nucleotides (where indicated). The protein samples were then incubated for 2 min at room temperature, centrifuged and applied to the GEMMA instrument for data collection. Each sample was scanned five times and the signal was added together to obtain the traces presented. The experiments were performed with a flow rate driven by 2 psi to minimise noise signals that may appear with elevated nucleotide or protein concentrations at the manufacturer’s recommended flow rate driven by 3.7 psi.

Glycyl radical characterization by EPR

EPR samples were handled under strictly anaerobic conditions inside a glovebox in 25 mM Tris-HCl pH 8, 50 mM KCl, 5 mM DTT in a final volume of 100 μL. A standard pre-reaction mixture containing 75 μM holo-PcNrdG, 1 mM S-adenosyl-methionine, was treated with a 12.5 excess of dithionite and was incubated for 20 min at 37 °C. The pre-mixture was added to a mixture of 50 µM apo-PcNrdD, 10 mM sodium formate, supplemented or not with 1.5 mM ATP or 1 mM dATP and in the presence of 1 mM CTP. The final reaction was incubated for 20 min at 37 °C and 100 µL sample was transferred into a standard EPR tube (Wilmad LabGlass 707-SQ-250M) and stored in liquid nitrogen before recording the EPR spectrum.

X-band CW-EPR spectra of PcNrdD were recorded on a Bruker ELEXYS-E500 spectrometer operating at microwave frequency of 9.38 GHz, equipped with a SuperX EPR049 microwave bridge and a cylindrical TE011 ER 4122SHQE cavity in connection with an Oxford Instruments continuous flow cryostat. Measuring temperature at 30 K was achieved using liquid helium flow through an ITC 503 temperature controller (Oxford Instruments). For achieving non-saturating conditions, a microwave power of 10 µW was applied for all measurements. Other EPR settings were: modulation frequency of 100 kHz; modulation amplitude of 8 G; and accumulation of 16 scans for an optimal S/N ratio. Double integrated spectra were used for intensity analysis.

Cryo-EM sample preparation and data acquisition

The proteins for cryo-EM analysis were all prepared in the same manner. Apo-PcNrdD was incubated with 1 mM NTP or dNTP at room temperature for 5 min, then centrifuged before injection onto a Superdex S200 10/300 column pre-equilibrated with a buffer containing 25 mM HEPES-NaOH pH 7.5, 100 mM KCl, 0.5 mM TCEP, 10 mM MgCl2 and supplemented with varying combinations of 0.5 mM NTP or 0.2 mM dNTP. The relevant peak was collected, then diluted at 0.5 mg/mL and the corresponding nucleotides were added to different final concentrations (0.5-5 mM NTP or 0.5-1 mM dNTP) (Table 1).

The grids were glow-discharged for 60 s at 20 mA using a GloQube (Quorum) instrument, then prepared in the following manner: 3 µL of sample at 0.5 mg/ml were applied on Quantifoil 1.2, 1.3 300 mesh Cu grids or Quantifoil 2.1, 300 mesh Au grids and plunge-frozen in liquid nitrogen-cooled liquid ethane using a Vitrobot Mark IV (Thermo Fisher Scientific) with a blot force of 1 and followed by 5 s blot time, at 4 °C and 100% humidity. The grids used for each sample are specified in Table 1.

Grids were clipped and loaded into a 300 kV Titan Krios G2 microscope (Thermo Fisher Scientific, EPU 2.8.1 software) equipped with a Gatan BioQuantum energy filter and a K3 Summit direct electron detector (AMETEK). Grids were screened for quality based on particle distribution and density, and images from the best grid were recorded. Micrographs were recorded at a nominal magnification of 105,000. Details of the other data collection parameters used for each sample are given in Table 1.

Cryo-EM data processing

Data processing was performed in cryoSPARC (Punjani et al, 2017). The first steps in processing of all datasets were patch motion correction and patch CTF estimation, followed by curation of exposures based on ice thickness, poor defocus estimation, etc. Other steps are as detailed below. Resolution estimation in cryoSPARC was done using a gold standard Fourier shell correlation value of 0.143 (Rosenthal & Henderson, 2003). Map post-processing was done using DeepEMhancer (Sanchez-Garcia et al., 2021) and post-processing resolution estimation was performed using DeepRes (Ramirez-Aportela et al, 2019). Models were placed in the maps using either phenix.dock_in_map (Liebschner et al, 2019) or Molrep (Vagin & Teplyakov, 1997) in the CCP-EM package (Burnley et al, 2017). Model building was done by alternating rounds of manual building in Coot (Emsley et al, 2010) with real space refinement in phenix.refine (Liebschner et al., 2019). Secondary structure restraints were used throughout. The inclusion of riding hydrogen atoms reduced the number of bad contacts in all models. An AlphaFold2 (Jumper et al., 2021) model of PcNrdD was used in the later stages of building the first model to resolve some ambiguous loops. All final models have good correlations between map and model and good stereochemical properties (Table 1). All structures have been deposited in the Protein Databank and the corresponding volumes deposited in the Electron Microscopy Data Bank with the accession numbers listed in Table 1.

PcNrdD-ATP-CTP complex

Data processing: A total of 17 033 movies were used after curation. A low-pass filtered volume was made in Chimera (Pettersen et al, 2021) from the crystal structure of NrdD from T. maritima (TmNrdD, PDB ID 4COI). From this volume, a set of templates was created and used for template-based picking with a diameter of 100 Å. Due to the large number of micrographs, they were split into two sets. The largest set (16 219 micrographs) was analysed first. About 8.6 million particles were extracted with a 350-pixel box size and classified into 60 2D classes. Eight representative classes containing different orientations (∼3.3 million particles) were used to generate four ab-initio models without symmetry (C1) and particles were subjected to three-dimensional (3D) heterogeneous refinement using four classes. The most populated model containing ∼1 million particles was subjected to homogeneous and non-uniform refinement, which gave a map with an overall FSC resolution of 2.94 Å. A further 435 695 particles were then extracted from the remaining 810 micrographs. These were subjected to 2D classification, and the resulting classes were merged with the larger set, giving 1 199 575 particles. The combined particles were subjected to homogeneous, non-uniform refinement and local refinement with a predefined mask, which gave a map at 2.87 Å (Table 1 - table supplement 1).

Generation of map with better ATP-cone density: The particles were further classified into 50 2D classes, from which 23 classes having 1 040 266 particles were selected. Four ab-initio models were generated and 570 730 particles with slightly better ATP-cone density were refined without symmetry. The particles were then subjected to 3D classification using ten classes of ∼57 000 particles each. The classes having slightly better ATP-cone density were selected for further processing and final refinement without symmetry gave a map with an overall FSC resolution of 3.17 Å from 291 231 particles, which was post-processed using DeepEMhancer (Table 1 - table supplement 1).

Model building and refinement: A partially complete model for one dimer of the dATP complex (see below) was placed in the map using phenix.dock_in_map. The ATP-cones were built based on the ones from the dATP-only tetramers (see below).

PcNrdD-dATP complexes

Tetramers

Data processing for first part: A set of 11 161 movies was used after curation, template-based picking based on the structure of TmNrdD and particle extraction using a box size of 448 pixels gave 5 331 420 particles. Seven 2D classes having 1 171 839 particles were used to make ab-initio models, followed by heterogeneous refinement. The best class having 461 020 particles and volume was used for homogeneous refinement, which gave a map at 3.3 Å. Non-uniform refinement followed by local refinement with a pre-determined mask gave a map at 3.1 Å with no imposed symmetry.

Data processing for second part: A set of 16 667 movies was used after curation. Particles were picked and extracted using the same box size as in the first part, which gave 7 889 234 particles. Ten 2D classes having 1 881 057 particles were used to make two ab-initio models, followed by heterogeneous refinement. The best class with 1 083 657 particles was selected and subjected to homogeneous refinement followed by non-uniform refinement and a local new refinement with a pre-determined mask, which gave a map at 2.9 Å with no imposed symmetry.

Merging of datasets and CTF refinement: The best particles from the first dataset were classified into 80 2D classes from which 36 having 399 523 particles were selected. The particles from the second dataset resulting in the best volume were classified into 100 2D classes, of which 31 having 949 610 particles were selected. These particle sets were merged (giving 1 349 133 particles) and refined against the best 3D volume which gave a map at 2.78 Å. Iterative refinement and NU refinement jobs, ultimately using C2 symmetry, and local CTF refinement with a defocus search range of +/− 2000, gave a map at 2.8 Å with C2 symmetry that was post-processed using DeepEMhancer (Table 1 - table supplement 4).

Model building and refinement: Model building was initially done for one dimer by fitting a homology model generated using SwissModel with T4NrdD as a template to the map using phenix.dock_in_map. Similar homology models for the four ATP-cones were placed by hand into the density followed by rounds of real space refinement. After almost completely rebuilding one dimer, the second dimer was placed into the density and model building and refinement were continued.

Dimers

The first part of the data processing was shared with the PcNrdD-dATP tetramers. In the same micrographs there were a number of 2D class averages that represented dimers. Eight such class averages having 1 916 387 particles were used to make two ab-initio models. This was followed by rounds of heterogeneous refinement from where the best model having 1 009 021 particles was refined with no symmetry to give a map at 2.8 Å. A final non-uniform refinement with C1 symmetry gave a map at 2.6 Å that was post-processed using DeepEMhancer (Table 1 - table supplement 4).

PcNrdD-dATP-CTP complexes

Tetramers

Blob picking was carried out from 11 780 curated micrographs with a maximum diameter of 200 Å and a minimum diameter of 90. In total 4 646 479 particles were then extracted with a box size of 448 pixels. This was followed by 2D classification, ab-initio model generation and heterogeneous refinements. The best class with 1 105 348 particles was used for non-uniform refinement with C2 symmetry, which gave a final map at 2.59 Å resolution that was post-processed using DeepEMhancer (Table 1 - table supplement 5).

Dimers

The initial data processing steps were the same as above. Template based picking from 8 796 curated micrographs using a low-pass filtered volume of the dATP-only tetramer gave 6 836 668 particles after extraction with a box size of 300 pixels. This was followed by 2D classifications, selection of classes representing dimers, ab-initio model generations and refinement of the best class. The final round of non-uniform refinement with 1 132 695 particles with C1 symmetry gave a map of 2.53 Å resolution. The estimated median resolution after post-processing was 2.1 Å (Table 1 - table supplement 5). Model building used the coordinates of the dATP-only dimer as a starting model.

PcNrdD-ATP-dTTP-GTP complex

Data processing: A total of 12 501 movies were processed and 6 105 movies were used after curation. Template-based picking was used with a diameter of 150 pixels. About 3.9 million particles were extracted from 5 952 movies with a 400 pixel box size and used for two-dimensional (2D) classification into 100 classes. Eleven representative 2D classes containing different orientations (∼2.4 million particles) were used to generate an ab-initio model and particles were subjected to three-dimensional (3D) heterogeneous refinement using ten classes, resulting in the most populated model containing 437 866 particles. This model was subjected to homogeneous and non-uniform refinement, which gave a map with an overall FSC resolution of 2.47 Å. The particles were then subjected to 3D classification and the resulting classes were merged for further processing and final refinement without symmetry (C1) gave a map with an overall FSC resolution of 2.4 Å that was post-processed using DeepEMhancer (Table 1 - table supplement 2).

Model building and refinement: A complete model for the dimer of the ATP-CTP structure (see above) was placed in the map using phenix.dock_in_map. The ATP-cones were removed as they were not visible in the reconstruction. After model building and refinement, the final map-to-model correlation value was X. Model quality statistics are presented in Table 1.

PcNrdD-ATP-dGTP complex

Data processing: Template-based picking was carried out on 14 373 curated micrographs and 4 522 028 particles were extracted with a box size of 350 pixels. This was followed by multiple rounds of 2D classifications, ab-initio model generation and refinements. C1 symmetry was applied on the best volume which gave a map with 589 345 particles having a final resolution of 2.58 Å. Model building and refinement used the PcNrdD-ATP/CTP complex as a starting model. (Table 1 - table supplement 3)

Bioinformatics

To construct an HMM logo of NrdD sequences, a representative selection of NrdD sequences from all domains of life including viruses were collected and aligned with Clustal Omega (Sievers et al, 2011). Subsequently an HMM model was built with hmmbuild from the HMMER suite (Eddy, 2011) and a logo was created using the Skylign web service using the “Information Content - Above Background” option (Wheeler et al, 2014). Conserved parts of the logo were extracted and displayed here.

Hydrogen-deuterium exchange mass spectrometry (HDX-MS)

All chemicals were from Sigma Aldrich. pH measurements were made using a SevenCompact pH-meter equipped with an InLab Micro electrode (Mettler-Toledo). A 4-point calibration (pH 2,4,7,10) was made prior to all measurements. The HDX-MS analysis was made using automated sample preparation on a LEAP H/D-X PAL™ platform interfaced to an LC-MS system, comprising an Ultimate 3000 micro-LC coupled to an Orbitrap Q Exactive Plus MS.

HDX was performed on 2 mg/ml PcNrdD with and without ligands (5 mM ATP + 5 mM CTP or 1 mM dATP + 5 mM CTP) in 25 mM HEPES, 100 mM KCl, 20 mM MgCl2 and 0.5 mM TCEP, pH 7.5. The HDX-MS was performed in one continuous run, the apo state being run before both complexed states. For each labelling time point, 3 μl HDX samples were diluted with 27 μl labelling buffer (containing the ligands) of the same composition prepared in D2O, pH(read) 7.4. The HDX labelling was carried out at t = 0, 30, 300, and 3000 s at 20 °C. Each time point was run in 4-6 replicates. The labelling reaction was quenched by dilution of 30 μl labelled sample with 30 μl of 1% trifluoroacetic acid (TFA), 0.4 M TCEP, 4 M urea, pH 2.5 at 1 °C. 50 μl of the quenched sample was directly injected and subjected to online pepsin digestion at 4 °C on an in-house immobilized pepsin column (2.1 x 30 mm). The online digestion and trapping was performed for 4 min using a flow of 50 µL/min 0.1% formic acid (FA), pH 2.5. The peptides generated by pepsin digestion were subjected to on-line SPE on a PepMap300 C18 trap column (1 mm x 15 mm) and washed with 0.1% FA for 60 s. Thereafter, the trap column was switched in-line with a reversed-phase analytical column (Hypersil GOLD, particle size 1.9 µm, 1 x 50 mm). The mobile phases were (A) 0.1% FA and 95% acetonitrile / 0.1% FA (B) and separation was performed at 1 °C using a gradient of 5-50% B over 8 minutes and then from 50 to 90% B for 5 min. Following the separation, the trap and column were equilibrated at 5% organic content until the next injection. The needle port and sample loop were cleaned three times after each injection with mobile phase 5% MeOH / 0.1% FA, followed by 90% MeOH / 0.1% FA and a final wash of 5% MeOH / 0.1% FA. After each sample and blank injection, the pepsin column was washed by injecting 90 μl of pepsin wash solution 1% FA / 4 M urea / 5% MeOH. In order to minimize carry-over, a full blank was run between each sample injection. Separated peptides were analysed on a Q Exactive Plus MS, equipped with a HESI source operated at a capillary temperature of 250 °C with sheath gas 12, Aux gas 2 and sweep gas 1 (au). For HDX analysis, MS full scan spectra were acquired at 70K resolution, AGC 3e6, max IT 200 ms and scan range 300-2000. For identification of generated peptides, separate non-deuterated samples were analysed using data dependent MS/MS with HCD fragmentation. A summary of the HDX experimental detail is reported in Figure 10 - figure supplement 1. The mass spectrometry raw files have been deposited at the ProteomeXchange Consortium via the PRIDE partner repository (Perez-Riverol et al. 2022) with the dataset identifier PXD047943.

Data analysis

PEAKS Studio X Bioinformatics Solutions Inc. (BSI, Waterloo, Canada) was used for peptide identification after pepsin digestion of non-deuterated samples. The search was done on a FASTA file with only the NrdD sequence. The search criteria were a mass error tolerance of 15 ppm and a fragment mass error tolerance of 0.05 Da, allowing for fully unspecific cleavage by pepsin. Peptides identified by PEAKS with a peptide score value of log P > 25 and no modifications were used to generate a peptide list containing peptide sequence, charge state and retention time for the HDX analysis. HDX data analysis and visualisation was performed using HDExaminer, version 3.3 (Sierra Analytics Inc., Modesto, US). The analysis was made on the best charge state for each peptide, allowed only for EX2 (except for bimodal parts of the protein for which EX1 calculation of uptake was allowed) and the two first residues of a peptide were assumed unable to hold deuteration. Due to the comparative nature of the measurements, the deuterium incorporation levels for the peptic peptides were derived from the observed relative mass difference between the deuterated and non-deuterated peptides without back-exchange correction using a fully deuterated sample (Engen & Wales, 2015). As a full deuteration experiment was not made, full deuteration was set to 75% of maximum theoretical uptake. The presented deuteration data is the average of all high and medium confidence results. The allowed retention time window was ± 0.5 minute. The spectra for all time points were manually inspected; low scoring peptides, obvious outliers and any peptides where retention time correction could not be made consistent were removed. As bottom-up labelling HDX-MS is limited in structural resolution by the degree of overlap of the peptides generated by pepsin digestion, the peptide map overlap is shown for the respective state in Figure 10 - figure supplement 1.

Supporting information

eLife Assessment

This study advances our understanding of the allosteric regulation of anaerobic ribonucleotide reductases (RNRs) by nucleotides, providing valuable new structural insight into class III RNRs containing ATP cones. The cryo-EM structural characterization of the system is solid, but some open questions remain about the interpretation of activity/binding assays and the newly incorporated HDX-MS results. The work will be of interest to biochemists and structural biologists working on ribonucleotide reductases and other allosterically regulated enzymes.

Reviewer #1 (Public Review):

The goal of this study is to understand the allosteric mechanism of overall activity regulation in an anaerobic ribonucleotide reductase (RNR) that contains an ATP-cone domain. Through cryo-EM structural analysis of various nucleotide-bound states of the RNR, the mechanism of dATP inhibition is found to involve order-disorder transitions in the active site. These effects appear to prevent binding of substrate and a radical transfer needed to initiate the reaction.

Strengths of the manuscript include the comprehensive nature of the work - including both numerous structures of different forms of the RNR and detailed characterization of enzyme activity to establish the parameters of dATP inhibition. The manuscript has been improved in a revision by performing additional experiments to help corroborate certain aspects of the study. But these new experiments do not address all of the open questions about the structural basis for mechanism. Additionally, some questions about the strength of biochemical data and fit of binding or kinetic curves to data that were raised by other referees still remain. Some experimental observations are not consistent with the proposed model. For example, why does dATP enhance Gly radical formation when the proposed mechanism of dATP inhibition involves disorder in the Gly radical domain?

The work is impactful because it reports initial observations about a potentially new mode of allosteric inhibition in this enzyme class. It also sets the stage for future work to understand the molecular basis for this phenomenon in more detail.

Reviewer #3 (Public Review):

The manuscript by Bimai et al describes a structural and functional characterization of an anaerobic ribonucleotide reductase (RNR) enzyme from the human microbe, P. copri. More specifically, the authors aimed to characterize the mechanism by how (d)ATP modulates nucleotide reduction in this anaerobic RNR, using a combination of enzyme kinetics, binding thermodynamics, and cryo-EM structural determination, complemented by hydrogen-deuterium exchange (HDX). One of the principal findings of this paper is the ordering of a NxN 'flap' in the presence of ATP that promotes RNR catalysis and the disordering (or increased protein dynamics) of both this flap and the glycyl radical domain (GRD) when the inhibitory effector, dATP, binds. The latter is correlated with a loss of substrate binding, which is the likely mechanism for dATP inhibition. It is important to note that the GRD is remote (>30 Ang) from the binding site of the dATP molecule, suggesting long-range communication of the structural (dis)ordering. The authors also present evidence for a shift in oligomerization in the presence of dATP. The work does provide evidence for new insights/views into the subtle differences of nucleotide modulation (allostery) of RNR, in a class III system, through long-range interactions.

The strengths of the work are the impressive, in-depth structural analysis of the various regulated forms of PcRNR by (d)ATP using cryo-EM. The authors present seven different models in total, with striking differences in oligomerization and (dis)ordering of select structural features, including the GRD that is integral to catalysis. The authors present several, complementary biochemical experiments (ITC, MST, EPR, kinetics) aimed at resolving the binding and regulatory mechanism of the enzyme by various nucleotides. The authors present a good breadth of the literature in which the focus of allosteric regulation of RNRs has been on the aerobic orthologues.

The addition of hydrogen-deuterium exchange mass spectrometry (HDX-MS) complements the results originating from cryo-EM data. Most notably, is the observation of the enhanced exchange (albeit quite subtle) of the GRD domain in the presence of dATP that matches the loss of structural information in this region in the cryo-EM data. The most pronounced and compelling HDX results are seen in the form of dATP-induced protection of peptides immediately adjacent to the b-hairpin at the s-site, where dATP is expected to bind based on cryo-EM. It is clear that the presence of dATP increases the rigidity of this region.

Weaknesses: The discussion of the change in peptide mobility in the N-terminal region is complicated by the presence of bimodal mass spectral features and this may prevent detailed interpretation of the data, especially for select peptide region that shows opposite trends upon nucleotide association. Further, the HDX data in the NxN flap is unchanged upon nucleotide binding (ATP, dATP, or CTP), despite changes observed in the cryo-EM data.

Author response:

The following is the authors’ response to the previous reviews.

eLife assessment

This study advances our understanding of the allosteric regulation of anaerobic ribonucleotide reductases (RNRs) by nucleotides, providing valuable new structural insight into class III RNRs containing ATP cones. The cryo-EM structural characterization of the system is solid, but some open questions remain about the interpretation of activity/binding assays and the newly incorporated HDX-MS results. The work will be of interest to biochemists and structural biologists working on ribonucleotide reductases and other allosterically regulated enzymes.

Public Reviews:

Reviewer #1 (Public Review):

The goal of this study is to understand the allosteric mechanism of overall activity regulation in an anaerobic ribonucleotide reductase (RNR) that contains an ATP-cone domain. Through cryo-EM structural analysis of various nucleotide-bound states of the RNR, the mechanism of dATP inhibition is found to involve order-disorder transitions in the active site. These effects appear to prevent binding of substrate and a radical transfer needed to initiate the reaction.

Strengths of the manuscript include the comprehensive nature of the work - including both numerous structures of different forms of the RNR and detailed characterization of enzyme activity to establish the parameters of dATP inhibition. The manuscript has been improved in a revision by performing additional experiments to help corroborate certain aspects of the study. But these new experiments do not address all of the open questions about the structural basis for mechanism. Additionally, some questions about the strength of biochemical data and fit of binding or kinetic curves to data that were raised by other referees still remain. Some experimental observations are not consistent with the proposed model. For example, why does dATP enhance Gly radical formation when the proposed mechanism of dATP inhibition involves disorder in the Gly radical domain?

The work is impactful because it reports initial observations about a potentially new mode of allosteric inhibition in this enzyme class. It also sets the stage for future work to understand the molecular basis for this phenomenon in more detail.

We express our gratitude to the reviewer for dedicating time to review our work and for the overall favorable assessment. We agree that the question of exactly how much the glycyl radical domain becomes more mobile without losing the glycyl radical entirely is an unresolved one but we also think that our work sets a solid basis for future experiments by us and others.

Reviewer #3 (Public Review):

The manuscript by Bimai et al describes a structural and functional characterization of an anaerobic ribonucleotide reductase (RNR) enzyme from the human microbe, P. copri. More specifically, the authors aimed to characterize the mechanism by how (d)ATP modulates nucleotide reduction in this anaerobic RNR, using a combination of enzyme kinetics, binding thermodynamics, and cryo-EM structural determination, complemented by hydrogen-deuterium exchange (HDX). One of the principal findings of this paper is the ordering of a NxN 'flap' in the presence of ATP that promotes RNR catalysis and the disordering (or increased protein dynamics) of both this flap and the glycyl radical domain (GRD) when the inhibitory effector, dATP, binds. The latter is correlated with a loss of substrate binding, which is the likely mechanism for dATP inhibition. It is important to note that the GRD is remote (>30 Ang) from the binding site of the dATP molecule, suggesting long-range communication of the structural (dis)ordering. The authors also present evidence for a shift in oligomerization in the presence of dATP. The work does provide evidence for new insights/views into the subtle differences of nucleotide modulation (allostery) of RNR, in a class III system, through long-range interactions.

The strengths of the work are the impressive, in-depth structural analysis of the various regulated forms of PcRNR by (d)ATP using cryo-EM. The authors present seven different models in total, with striking differences in oligomerization and (dis)ordering of select structural features, including the GRD that is integral to catalysis. The authors present several, complementary biochemical experiments (ITC, MST, EPR, kinetics) aimed at resolving the binding and regulatory mechanism of the enzyme by various nucleotides. The authors present a good breadth of the literature in which the focus of allosteric regulation of RNRs has been on the aerobic orthologues.

The addition of hydrogen-deuterium exchange mass spectrometry (HDX-MS) complements the results originating from cryo-EM data. Most notably, is the observation of the enhanced exchange (albeit quite subtle) of the GRD domain in the presence of dATP that matches the loss of structural information in this region in the cryo-EM data. The most pronounced and compelling HDX results are seen in the form of dATP-induced protection of peptides immediately adjacent to the b-hairpin at the s-site, where dATP is expected to bind based on cryo-EM. It is clear that the presence of dATP increases the rigidity of this region.

We are happy that both reviewers find the HDX-MS experiments to be a valuable addition to the existing data.

Weaknesses:

The discussion of the change in peptide mobility in the N-terminal region is complicated by the presence of bimodal mass spectral features and this may prevent detailed interpretation of the data, especially for select peptide region that shows opposite trends upon nucleotide association.

Further, the HDX data in the NxN flap is unchanged upon nucleotide binding (ATP, dATP, or CTP), despite changes observed in the cryo-EM data.

We are grateful to the reviewer for the comprehensive feedback on the HDX-MS part and for identifying areas for improvement. The HDX analysis was of course undertaken with the intention of identifying differences in disorder of the NxN flap and GRD region. From an HDX perspective both regions were found to be highly susceptible to HDX regardless of state/ligand, due to surface accessibility and/or very fast dynamics. However, this does not mean that there is no difference in the degree of order of these regions upon ligand addition, simply that we with HDX-MS, in the limited time span of 30-3000 seconds, could not conclusively support an increased disorder. We have rephrased the discussion text to reflect this fact

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

On page 5 (and throughout the manuscript) there are some inconsistencies in how dissociation constants for effectors and inhibitors are described - for example, D in KD is sometimes subscripted and sometimes not.

Thank you for noticing these remaining errors. We hope that we have fixed all of them now.

Reviewer #3 (Recommendations For The Authors):

The authors addressed many of the initial concerns raised. The addition of the HDX-MS data in this revision is a welcomed contribution to the work and complements the cryo-EM data. In select cases, the data may be over-interpreted. This reviewer suggests that the authors revise the text in this section so that it is more consistent with the presented data.

Specific points:

(1) The bimodal mass spectral features in the N-terminal domain complicate the data interpretation. Specifically for peptides in 81-99 region, the fast exchanging feature shows protection in the presence of (d)ATP/CTP, but the opposite trend is observed for the slow exchanging species. It is therefore advisable to not make absolutes about the HDX results in this region, as the data are complicated.

As stated by the reviewer, it is not possible from the presented HDX data to deduce if this is a result of 50% loaded dimer or the oligomerization state of the protein. We have remedied this by removing mentions of a difference between the dATP and ATP in bimodality. Also, we have addressed this in the text by stating that the main reason is most likely the different oligomerization states present in solution. Nevertheless, it is clear from the HDX data that the N-terminal region and 81-99 are very interesting, and it was somewhat disappointing that due to the dynamics of the oligomerization it was not possible to SEC-purify pure dimer or tetramer samples for HDX-MS, in order to deconvolute the cause.

(2) Related to #1, the authors assign the bimodal HDX behavior to EX1 mechanism, but this is not necessarily (and unlikely) true based on the limited time points. The authors also state that it originates from the heterogeneity of the sample: "a mixture of states" which could reflect the mixture of oligomerization states. The authors should be careful assigning EX1 mechanism unless there are compelling results to support it.

We apologize for the unfortunate phrasing. It was not our intention to imply that the bimodality is due to true EX1 kinetics. See the above answer. The mention of EX1 has been removed from the discussion text.

(3) The deuterium uptake for peptide 118-126 is very small (~1Da) compared to the length of the peptide. The change in deuterium uptake (<0.25Da) from dATP is very small; the authors should proceed with caution when presenting interpretations of such small differences.

We agree with the reviewer that extra caution should be taken when dealing with such a small difference. However, the 118-126 peptide has been significance tested in both HDExaminer and Deuteros 2.0, and we also observed this for more than one run. The difference in uptake is small but increases to significance at the longer labelling times. The proximity to the NxN flap makes it interesting in context of an allosteric conformational change. i.e the dynamics of the NxN might be too fast so we can only see some secondary effects. We would like to keep the data in Figure 10 for reasons of transparency. In essence this is similar to the observed bimodality mentioned above: we cannot fully explain the observation but present the data as it was observed.

(4) On p. 22, the authors should consider revising the following statement: "confirming dATP binding to the s-site." Even though the HDX data are most compelling for the protection of peptides 178-204 and 330-348 that are adjacent to the beta-hairpin at the s-site, these data cannot "confirm" a binding site for a small molecule, such as dATP.

We appreciate that the reviewer has pointed out that the statement can be misleading, and we agree that the binding site of small molecules can’t be confirmed based solely on HDX data. The sentence reformulated to clarify that the binding site was confirmed based on the combined evidence of HDX data and the previously presented biochemical and structural data on the s-site.

Version 3

Abstract

Abstract

A small, nucleotide-binding domain, the ATP-cone, is found at the N-terminus of most ribonucleotide reductase (RNR) catalytic subunits. By binding ATP or dATP it regulates the enzyme activity of all classes of RNR. Functional and structural work on aerobic RNRs has revealed a plethora of ways in which dATP inhibits activity by inducing oligomerization and preventing a productive radical transfer from one subunit to the active site in the other. Anaerobic RNRs, on the other hand, store a stable glycyl radical next to the active site and the basis for their dATP-dependent inhibition is completely unknown. We present biochemical, biophysical and structural information on the effects of ATP and dATP binding to the anaerobic RNR from Prevotella copri. The enzyme exists in a dimer-tetramer equilibrium biased towards dimers when two ATP molecules are bound to the ATP-cone and tetramers when two dATP molecules are bound. In the presence of ATP, P. copri NrdD is active and has a fully ordered glycyl radical domain (GRD) in one monomer of the dimer. Binding of dATP to the ATP-cone results in loss of activity and increased dynamics of the GRD, such that it can not be detected in the cryo-EM structures. The glycyl radical is formed even in the dATP-bound form, but the substrate does not bind. The structures implicate a complex network of interactions in activity regulation that involve the GRD more than 30 Å away from the dATP molecules, the allosteric substrate specificity site and a conserved but previously unseen flap over the active site. Taken together, the results suggest that dATP inhibition in anaerobic RNRs acts by increasing the flexibility of the flap and GRD, thereby preventing both substrate binding and radical mobilisation.

Introduction

Ribonucleotide reductases (RNRs) are a family of enzymes with sophisticated radical chemistries and allosteric regulation. RNRs produce all four deoxyribonucleotides and are the only enzymes providing de novo building blocks for DNA replication and repair in all free-living organisms (Mathews, 2016). Virtually all RNRs possess an allosteric site regulating substrate specificity (the s-site), a crucial aspect of RNR that ultimately provides a balanced supply of dNTPs (Mathews, 2016, 2018). A second allosteric site (the a-site) is located in an N-terminal domain called the ATP-cone (Aravind et al, 2000), found in the majority of RNRs (Jonna et al, 2015). Whereas the s-site binds all deoxyribonucleoside triphosphates (dNTPs) and usually ATP, the a-site in the ATP-cone can only bind ATP and dATP. Binding of ATP activates RNR, and binding of dATP inhibits it (Hofer et al, 2012; Martínez-Carranza et al, 2020).

The RNR family consists of three evolutionarily related classes (I to III) with a common radical-based reaction mechanism but differing in the mode of radical generation and in quaternary structure (Högbom et al, 2020; Lundin et al, 2015). Class I, strictly aerobic RNRs are heterotetramers consisting of two α subunits (NrdA) with binding sites for substrates and allosteric nucleotides and two β subunits (NrdB) that harbour a radical initiator in the form of a stable radical or an oxidized metal cluster (Högbom et al, 2020). Class II RNRs are either dimers or monomers with an α subunit (NrdJ) that, in addition to binding sites for substrate and allosteric nucleotides, also harbours the radical initiator cofactor adenosylcobalamin. Class III RNRs, which only function anaerobically, are dimers of an α subunit (NrdD) with a stable radical close to the active site, and binding sites for allosteric nucleotides. The radical is introduced via encounter with a specific radical-SAM enzyme called NrdG, and once activated NrdD can perform multiple turnovers in the absence of NrdG (Backman et al, 2017; Torrents et al, 2001).

The dATP-dependent inhibition has been biochemically characterised in representatives of all three RNR classes, but only structurally studied in class I. A common denominator of all inhibited class I RNRs is oligomerization, leading to disturbed radical transfer between the α and β subunits. Hitherto four different inhibitory oligomerization mechanisms have been identified: heterooctameric α4β4 complexes in Escherichia coli, Clostridium botulinum and Neisseria gonorrhoeae (Ando et al, 2011; Martínez-Carranza et al, 2020; Wei et al, 2014), an α4 complex in Pseudomonas aeruginosa (Johansson et al, 2016), α6 complexes in the eukaryotes Homo sapiens, Saccharomyces cerevisiae and Dictyostelium discoideum (Ando et al, 2016; Brignole et al, 2018; Crona et al, 2013; Fairman et al, 2011), and a β4 complex in the bacterium Leeuwenhoekiella blandensis (Rozman Grinberg et al, 2018a). All these class I oligomers involve protein-protein interactions mediated by the ATP-cone domain.

The ATP-cone is in essence restricted to RNRs and the RNR-specific repressor NrdR (Rozman Grinberg et al, 2022). Among RNRs it is found in about 80% of class III enzymes and about 50% of class I enzymes, but less than 10% of class II enzymes (Jonna et al., 2015). The presence and function of the ATP-cone domain distinguish anaerobic RNRs from the other members of the large glycyl radical enzyme (GRE) family that are otherwise structurally related (Backman et al., 2017). The allosteric regulation mechanisms of class III RNRs (NrdDs) from E. coli, bacteriophage T4 and Lactococcus lactis were determined several decades ago (Andersson et al, 2000; Eliasson et al, 1994; Torrents et al, 2000). T4NrdD was the first class III RNR structure solved (Logan et al, 1999), and later, structures of Thermotoga maritima NrdD (TmNrdD) have been published (Aurelius et al, 2015; Wei et al, 2014). However, these two NrdDs lack an ATP-cone and the structural basis for allosteric activity regulation in the class III RNRs is still an outstanding question. Here, we present structural, biochemical and biophysical studies on the ATP-cone-containing NrdD from the human pathogen Prevotella copri (PcNrdD; Nii & al., 2023), showing that the binding of dATP causes increased flexibility of the C-terminal glycyl radical-bearing domain, as well as a flap-like loop that binds across the top of the active site in the active form. This increase in dynamics is coupled to enzymatic inactivation by inhibition of substrate binding and radical transfer. The final outcome of dATP inhibition is thus blocked radical transfer in both class I and class III RNRs, but it is achieved by completely different mechanisms, i.e. oligomerization prevents radical transfer between subunits in class I RNRs but long-range modulation of the flexibility of key structural elements prevents substrate binding in class III RNRs.

Results

Allosteric activity regulation by ATP and dATP

The amino acid sequence of PcNrdD suggests that it belongs to the formate-requiring class III RNRs (Burnim et al, 2022b; Wei et al., 2014), and initial optimization of the assay composition showed that this was the case (Figure 1 - figure supplement 1). ATP-cone mediated activation of PcNrdD enzyme activity by ATP or inhibition by dATP was tested with two different substrates. With GTP as substrate the s-site was filled with dTTP as effector, and in absence of any a-site effector the basal level of GTP reduction was 21±2 nmol/min/mg. An increasing concentration of ATP stimulated activity to a kcat of 1.3 s-1 (478±26 nmol/min/mg), with a KL of 0.67±0.12 mM (Fig. 1a). On the other hand, an increasing concentration of dATP resulted in an abrupt inhibition of enzyme activity with a Ki of 74±24 µM (Fig. 1b). Hence, the Ki value for dATP inhibition via the ATP-cone is approximately 10-fold lower compared to the KL value for ATP activation via the ATP-cone.

Activity assays of Prevotella copri NrdD in the presence of ATP or dATP.

A 2.5-fold excess of holoNrdG over apoNrdD was used to study the allosteric regulation effect of ATP and dATP on the a-site. GTP reduction was monitored with 1 mM dTTP as effector in the s-site and titrated with ATP (a) or dATP (b) in the a-site. CTP reduction assays were titrated with ATP (c) or dATP (d), in this case acting both as s-site effectors and a-site regulators. Experiments were performed in triplicate for (a), (c), and (d) and with four replicates for (b). Insets in panels (b) and (d) show the results plotted in log scale. Curve fits for calculation of KL and Ki used equations (1) and (2), respectively, given in Materials and Methods, and in (d) equation (3). Curve fits for (c) and (d) used a start activity of 50 nmol/min/mg, and in (c) only results for 0-3 mM ATP were used. The R values for curve fits in panels (a) - (d) were 0.99, 0.95, 0.83, and 0.96, respectively.

The activation/inhibition experiments were also carried out with CTP as substrate. Specificity effectors for CTP reduction are ATP or dATP, so in these experiments both the s- and a-sites were conceivably filled with ATP or dATP, respectively. It is obvious from Figure 1c that increasing concentrations of ATP initially stimulate activity and that high levels of ATP cause inhibition, plausibly by ATP also acting as substrate. We therefore used only results up to 3 mM ATP to calculate an approximate kcat of 1.1 s-1 (385±116 nmol/min/mg) with an apparent KL of 0.67±0.52 mM (Fig. 1c). Addition of dATP initially results in a stimulation of enzyme activity, conceivably when the s-site is filled with dATP, after which competing inhibition appears when dATP also binds to the ATP-cone (Fig. 1d). All in all, these results suggest that the ATP-cone in PcNrdD responds similarly to inhibition by dATP as has been observed before for several class I RNRs.

Binding of nucleotides to the ATP-cone

To confirm binding of ATP or dATP to the ATP-cone we used isothermal titration calorimetry (ITC) and microscale thermophoresis (MST) (Fig. 2). Substrate GTP and s-site effector dTTP were used to fill up the other nucleotide binding sites in PcNrdD. Binding of dATP occurred with a KD of 6 µM, whereas binding of ATP was 4-fold weaker with a KD of 26 µM (Fig. 2a-c). The number of bound dATP molecules observed with ITC was almost 2-fold higher for dATP compared to ATP, but both values were below 1.

Binding of nucleotides to the ATP-cone.

Binding assay using ITC: Representative thermograms of binding of ATP to 50 µM PcNrdD (a); and binding of dATP to 200 µM PcNrdD (b); (c) thermodynamic parameters of ligand binding. All titrations were performed in the presence of 1 mM GTP substrate and 1 mM s-site effector dTTP at 20 °C. Binding of nucleotides to the ATP-cone using MST: (d) Binding of ATP, (e) binding of dATP. All MST binding experiments were performed in the presence of 1 mM s-site effector dTTP and 5 mM GTP substrate at room temperature. Control experiments were done for ITC, by titrating the nucleotide into the buffer (not shown).

Nucleotide binding measured with MST confirmed our results for ATP (Fig. 2d) that had a KD of 25 µM, and dATP that with this method had a KD of 23 µM (Fig. 2e). However, in cryo-EM experiments, where higher nucleotide concentrations can be used, we show that the ATP-cone can bind two ATP or two dATP molecules (see below), as also expected based on its sequence. The KL and Ki values reported above for ATP-activation and dATP-inhibition of GTP reduction conceivably reflect the requirement for two bound nucleotide molecules to achieve these effects, whereas the KD measurements via ITC and MST are restricted by the component concentrations that can be used and plausibly only reflect the first nucleotide bound.

It is noteworthy that the KD values for ATP and dATP do not differ by orders of magnitude, as has been found for several class I RNRs (Ando et al, 2016; Birgander et al, 2004; Ormö & Sjöberg, 1990; Rozman Grinberg et al., 2018a; Rozman Grinberg et al, 2018b; Torrents et al, 2006). Collectively ITC and MST instead suggested that ATP and dATP may interact at two distinct sites in the ATP-cone for the functional readout of dATP inhibition, as has earlier been observed in the transcriptional repressor NrdR (Rozman Grinberg et al., 2022). To test this hypothesis, apo-PcNrdD was incubated with ATP only, dATP only or a combination of ATP and dATP, then desalted, boiled, centrifuged and the nucleotide content of the supernatant analysed by HPLC. Figure 2 - figure supplement 1 shows that only dATP was bound when PcNrdD was incubated with a combination of ATP and dATP. This was also the case when incubations were performed in the presence of s- site effector and substrate, and also when the desalting was performed in the presence of s-site allosteric effector and substrate. In incubations with only ATP, s-site effector and substrate were needed during the entire work-up procedure for ATP to be retained by the protein. In conclusion, the binding experiments suggest that the PcNrdD ATP-cone can bind either ATP or dATP but not both simultaneously. Importantly, the results point to allosteric communication between the s-site, the active site and the a-site in the ATP- cone, as dTTP and GTP are required for any ATP to be retained. Both of these observations are consistent with the structural analyses shown below.

Modulation of the oligomeric state by ATP and dATP in the presence of allosteric effectors

Binding of dATP to the ATP-cone in all class I RNRs studied to date results in formation of higher oligomers unable to perform catalysis (reviewed in Martínez-Carranza et al., 2020). We therefore next asked whether a similar mechanism would be valid for dATP-mediated inhibition of a class III RNR. To elucidate the oligomeric states of PcNrdD we used gas-phase electrophoretic mobility molecular analysis (GEMMA) (Kaufman et al, 1996), a method similar to electrospray mass spectrometry. It requires low protein and nucleotide concentrations as well as volatile buffers.

When loaded with ATP, PcNrdD was in a dimer-tetramer equilibrium shifted towards dimers (Fig. 3, left; Fig. 3 - figure supplement 1). Addition of the CTP substrate or the s- site effector dTTP resulted in a similar equilibrium. In contrast, the dimer-tetramer equilibrium of dATP-loaded PcNrdD was shifted towards tetramers, and this equilibrium was likewise not affected by addition of the CTP substrate (Fig. 3, right; Fig. 3 - figure supplement 1). Hence, dATP-inhibition of enzyme activity seems to work differently in PcNrdD than the clearcut and drastic change in oligomeric structure observed for class I RNRs, since both ATP-loaded and dATP-loaded forms of PcNrdD were in dimer-tetramer equilibria.

Oligomeric states of Prevotella copri NrdD in the presence of nucleotides determined by GEMMA.

(a) Apo-PcNrdD (2 µM) loaded with the activator ATP (50-100 µM) (black and red) in the presence of CTP (100 µM) as a substrate (cyan) or dTTP (100 µM) (orange) as the allosteric effector. (b) apo-PcNrdD (2 µM) loaded with the inhibitor dATP (50-100 µM) (black and red) in the presence of CTP (50 µM) (cyan) as the substrate. Each sample was scanned 5 times to increase signal-to-noise level. (c) Calculated numbers of monomers based on measured molecular weight and fractions of dimers versus tetramers after conversion of experiments (a) and (b) to mass concentrations.

Glycyl radical formation in presence of ATP or dATP

Next, we asked whether dATP inhibition was mediated by prevention of glycyl radical formation. Apo-PcNrdD was incubated with NrdG and S-adenosylmethionine in presence of ATP or dATP. Whereas reduction of substrate requires the presence of formate, the formation of a glycyl radical does not. We therefore incubated with or without formate as well as with or without substrate. Figure 4 shows that the glycyl radical was readily formed in the presence of dATP both with and without the CTP substrate, and the result was similar in the absence of formate (Figure 4 - figure supplement 1). Interestingly, under all conditions the glycyl radical content was higher in the presence of dATP compared to with ATP, and it was also more stable than in the absence of any effector nucleotides. In addition, the shape of the doublet EPR signal was similar under all conditions studied. Collectively, these results show that the glycyl radical is readily formed in the presence of a dATP concentration that inhibits RNR activity.

Glycyl radical formation after 20 min in presence of formate and ATP±CTP or dATP±CTP.

Nucleotide concentrations were: 1.5 mM ATP, 1 mM dATP, and 1 mM CTP. Traces are arbitrarily moved to increase visibility and scaled to identical units (Y-axes).

Substrate binding to PcNrdD

To study substrate binding to activated and inhibited PcNrdD we used MST with labelled protein and GTP or CTP as substrate. The results were clear-cut in both sets of experiments (Fig. 5). ATP-loaded PcNrdD bound GTP in the presence of s-site effector dTTP (Fig. 5a), and it bound CTP in the presence of only ATP (Fig. 5b), whereas dATP-loaded PcNrdD had extremely low affinity to GTP and CTP under similar conditions (Fig 5c,d). These results show that the dATP-inhibition of enzyme activity in a class III RNR is mediated by inhibited binding of substrate.

Binding of substrate to ATP- or dATP-loaded PcNrdD.

Binding of GTP (a), and binding of CTP (b) to ATP-loaded PcNrdD. (c) Binding of GTP, and (d) binding of CTP to dATP-loaded PcNrdD. No additional nucleotides were present in CTP binding experiments, whereas binding of GTP was performed in the presence of 1 mM S-site effector dTTP. Fitted KDs are 2.8 ± 0.5 and 11 ± 2.4 mM for GTP and CTP binding respectively in the presence of ATP. In the presence of dATP fitted KDs were ≥ 2833 mM and ≥ 803 mM for GTP and CTP binding, respectively. Experiments were performed at room temperature.

Cryo-EM structure of the ATP-CTP bound PcNrdD dimer

2D classes of particles extracted from micrographs of PcNrdD incubated with effector ATP and substrate CTP indicated only dimers, with no appreciable amounts of tetramer. From a set of 570 730 particles we obtained a map from cryoSPARC at 3.0 Å. Application of deep learning local sharpening in DeepEMhancer (Sanchez-Garcia et al, 2021) improved the quality of the map, and we were able to build a reliable model for the great majority of the structure. As the ATP-cone density still did not permit reliable modelling, 3D classification was carried out on this particle set, enabling the identification of a subset of 291 231 particles that gave a volume at 3.2 Å with better-defined ATP-cones (Figure 6a). This map was used for refinement of the final model (Table 1; Table 1 – table supplement 1).

Structure of the PcNrdD dimer in complex with effector ATP and substrate CTP.

a) cryo-EM map with C1 symmetry showing the best-ordered ATP-cones after post-processing using DeepEMhancer. For the right-hand (active) monomer, the ATP-cone (residues 1-91) and the linking helix (92-104) are coloured olive, the linker and NxN flap region pale green, the glycyl radical loop red and the C-terminal extended region orange. The loop in the middle of the α/β barrel containing the radical initiator cysteine Cys416 at its tip is yellow. b) Overview of the PcNrdD dimer with ATP and CTP. The two monomers of the dimer are coloured in different shades of grey. The ATP and CTP molecules are shown as sticks. c) Schematic of the domain organisation of PcNrdD with the same colour scheme as a) and b); d) Closeup view of the binding of four ATP molecules to the dimer of ATP-cones in PcNrdD. The view is from the bottom of the molecule as seen in a) and b). Disordered loops are shown as dotted lines. e) Closeup view of the active site including the cryo-EM map for CTP. Residues within 4 Å of CTP are shown as sticks and polar interactions as dotted lines. f) Sequence logos of NrdD sequence motifs. A) ATP-cone plus downstream interaction partners, B) central parts of sequence and C) C-terminal parts. The numbering is from PcNrdD. Segments were selected to illustrate amino acids discussed in the text.

PcNrdD is a compact dimer (Fig. 6b). Each monomer consists of 739 residues, of which residues 1-91 constitute the ATP-cone, 92-110 a linker region, 111-122 a flap that folds over the substrate in the active site, 123-671 are the core domain with its 10-stranded α/β barrel fold typical of the RNR/glycyl radical enzyme (GRE) family, and 676-739 are the C-terminal glycyl radical domain (GRD) that contains a structural Zn2+ site followed by the buried loop containing the radical glycine residue at position Gly711 and a C-terminal tail that extends across the surface of the protein (Figure 6b,c). The closest structural neighbours as identified by the DALI server (Holm, 2022) are the NrdDs from bacteriophage T4 (Logan et al., 1999) (1H7A, 2.5 Å root mean square deviation [rmsd] in Cα positions for 535 residues, 27% sequence identity) and T. maritima (Aurelius et al., 2015; Wei et al., 2014) (4U3E, 3.0 Å rmsd for 508 residues, 16% sequence identity; Figure 6 - figure supplement 1). The linker, flap and GRD are well-ordered in one of the monomers (dark blue in Fig. 6) and substrate CTP is bound (Figure 6 - figure supplement 2a,b), while in the other monomer this region is less well-ordered and the active site is empty (Figure 6 - figure supplement 2c,d). We will refer to these as the “active” and “inactive” monomers respectively.

At the concentration of ATP used (1 mM), ATP is not observed to bind to the s-sites near the dimer interface. The ATP-cones containing the a-site are flexible, but the reconstruction from a subset of particles allowed the modelling of two ATP molecules per ATP-cone and many of the most important side chains. The ATP molecules are bound similarly to the twin dATP molecules seen in the ATP-cones of P. aeruginosa NrdA (PaNrdA) and L. blandensis NrdB (LbNrdB) (Johansson et al, 2016; Rozman Grinberg et al., 2018a), and to dATP bound to PcNrdD itself (see below). However, the ATP-cones in PcNrdD exhibit a hitherto unobserved relative orientation in which the four triphosphate groups are projected towards each other. (Fig. 6b,d). The large negative charge is compensated for by the proximity of as many as 12 Arg and Lys side chains from both monomers. The ATP-cones interact by contacts involving the “roof” of the domain (residues 1-18). Arg11 of each ATP-cone reaches over to interact with the α-phosphate group of one of the ATP molecules in the other ATP-cone, and a salt bridge is formed between Asp9 at the end of the lid and Arg91 at the end of the last helix of the other ATP-cone. Strikingly, the dimer of ATP-cones is slightly offset from the dimer axis of the core protein, such that the ATP-cone of the inactive monomer approaches the core domain of the active monomer, while the ATP-cone of the active monomer is more distant from the core domain of the inactive one (Fig. 6a,b). However, the closest atom of any ATP molecule is over 30 Å from the substrate CTP in the active site.

The glycyl radical site is found at Gly711, at the tip of a loop projected from the C-terminal region of the enzyme into the barrel, where it approaches the radical initiator cysteine Cys416 at the tip of a loop stretching through the barrel from the other side (Fig. 7). The second cysteine necessary for the radical mechanism of PcNrdD is Cys171 on the first β-strand of the barrel (Fig. 6e). Consistent with the fact that PcNrdD belongs to the group of anaerobic RNRs that use formate as overall reductant (Burnim et al, 2022a; Mulliez et al, 1995; Wei et al., 2014), there is no third cysteine residue on the sixth β-strand, and its place is taken by Asn435. As usual for GREs, Gly711 is completely buried within the barrel, with no solvent-accessible surface area. The base, phosphate and most of the ribose of substrate CTP are well-defined but the density is weaker around the 5’ C-atom (Fig. 6e).

The intricate network of interactions between the NxN flap, C-terminal GRD, loop 2 and substrate in the ATP-CTP complex of PcNrdD.

The NxN flap is coloured light green, the GRD orange, loop 2 dark red and the Cys radical loop yellow. The Gly radical loop is semi-transparent for clarity. The CA atom of Gly711 is indicated by an orange sphere. Important hydrogen bonds are shown as dotted lines.

After the glycyl radical loop emerges from the α/β barrel, it traverses the top of the barrel, forming a short helix from residues 723-730 and ending in an extended tail. This conformation of the C-terminus is very similar to the one seen for TmNrdD in PDB entry 4U3E (Wei et al., 2014). This tail makes few specific interactions with other residues in the core.

When compared to previously determined NrdD structures, the first common structural element is a long α-helix in the core domain stretching from Thr123 to Leu141. Between this and the ATP-cone are the linker (92-110), followed by the flap over the substrate in the active site (111-122). Two residues from the flap make H-bonds to CTP: Asn114 to the γ-phosphate group and Asn116 to the amino group of the cytosine base. We call this region the “NxN flap”, as the NxN motif (114-116) is one of the most highly conserved sequence motifs in the NrdD family (Fig. 6f). This is the first time such interactions have been observed in an NrdD, as the flap corresponds to a disordered segment of 17 residues in the structures of TmNrdD and T4NrdD. However, an AlphaFold2 (Jumper et al, 2021) model of TmNrdD (entry Q9WYL) suggests that this linker potentially has the same conformation in TmNrdD and possibly all other NrdDs, whether or not they have an ATP-cone. As seen previously in TmNrdD, the triphosphate moiety is recognised through the conserved HxHD motif (157-160), H-bonds to the main chain atoms of a loop (577-581) following the third-last strand of the α/β barrel and the dipole of the following helix (Fig. 6f).

Significantly, the NxN flap forms the nexus of a network of interactions (Fig. 7) between the substrate, the linker to the ATP-cone, the C-terminal GRD, and “loop 2” (residues 304-311) that is responsible for communicating the substrate specificity signal from the s-site at the dimer interface to the active site (Aurelius et al., 2015; Larsson et al, 2001; Logan et al., 1999). Important interactions linking the NxN flap to the C-terminus include H-bonds from Tyr712, which immediately follows the radical Gly711, to Glu581, which bridges to both the substrate and Ala115 in the flap, as well as from Arg719 in the GRD to the main chain carbonyl groups of Asn114 and Met117 in the flap. Tyr712 is highly conserved in the NrdD family (Fig. 6f and 7). This nexus is crucial for allosteric activity regulation, as will be discussed later.

As the NxN flap appears to close off the top of the active site, we used the web server Caver Web v1.1 (Stourac et al, 2019) to analyse substrate access to the active site in the ATP-bound form, and identified a tunnel ∼12 Å long leading from the surface of the protein to the active site that is 5.6 Å in diameter at its narrowest point (Figure 7 - figure supplement 1). Thus, in principle a fresh substrate could diffuse into the active site on each catalytic cycle.

Cryo-EM structure of the ATP-dTTP-GTP dimer

In order to probe the generality of the conformations observed in the ATP-CTP complex, we also solved the structure of PcNrdD in complex with ATP in the a-site, specificity effector dTTP in the s-site, and substrate GTP in the active site (c-site). Again, 2D classes show a predominantly dimeric form with no more than ∼10% tetramers. From a set of 437 886 particles selected from 2D classes with the best apparent density for the ATP-cone region, we obtained a volume at 2.40 Å resolution for the dimer with C1 symmetry that was further improved using DeepEMhancer (Table 1; Table 1 - table supplement 2). dTTP is clearly visible in the s-site (Figure 8 - figure supplement 1a) as is GTP in the active site (Figure 8 - figure supplement 2). A Mg2+ ion can be modelled in the s-site, coordinated by Glu290 and all three phosphate groups of dTTP. A β-hairpin loop between residues 185-192 that is disordered in the ATP-CTP structure in the absence of an s-site nucleotide is here ordered, due to an H-bond to the 2’-OH group of dTTP. Readout of the s-site nucleotide’s identity is achieved through an H-bond to Asn298 (Figure 8 - figure supplement 1).

Structure of PcNrdD in complex with a-site effector ATP, s-site effector dTTP and substrate GTP.

a) Volume for the PcNrdD dimer in complex with ATP, dTTP and GTP. The volume is contoured at a low level to emphasise the weak density for the more ordered ATP-cone domain in the active monomer. b) Ribbon diagram of the ATP-dTTP-GTP complex coloured as in Figure 6b, except that the most N-terminal helix (olive) now extends from residues 85-104. c) Zoom in on panel b) to illustrate the highly asymmetrical loop 2 conformations in the active and inactive monomers. Loop 2 in the inactive monomer (light blue) is coloured wheat, while in the active monomer (dark blue) it is purple. The conformation of loop 2 that forms a complementary cradle for the guanosine base of the substrate GTP in the active monomer induces a conformation of the other loop 2 that precludes substrate binding. d) Severe steric clash of loop 2 of the inactive monomer with the substrate and NxN flap. The view is rotated approximately 180° from panels a-c. Loop 2 from the inactive monomer (wheat) is superimposed on the active site of the active monomer (dark blue). The molecular surface of the loop is shown to emphasise that this conformation is incompatible with an ordered NxN flap (light green) and substrate binding. The Cys radical loop is yellow and Cys416 is marked by a sphere. The Gly radical loop is omitted for clarity.

Like the ATP-CTP complex, the ATP-dTTP-GTP complex can also be partitioned into an active and an inactive monomer. The ATP-cone region is very flexible, but the last helix of the ATP-cone in the active monomer can be traced back to residue 85 (Fig. 8a,b). The helix is not kinked at residue 91 as in the ATP-CTP dimer but forms an uninterrupted helix between residues 85-104. The volume for the ATP-cone region lies entirely on one side of the PcNrdD dimer axis, as is also apparent in some of the 2D classes (Figure 8 – figure supplement 3). The ATP-cone cannot be modelled in detail, but if the volume is contoured at very low level, there is almost enough density for one ATP-cone. Taken together, this suggests that there is one partially disordered ATP-cone and one highly flexible in the ATP-dTTP-GTP complex. In the active monomer, the NxN flap and entire C-terminus are ordered while in the inactive monomer, the entire assembly of linker and NxN flap up to residue 120 and the entire GRD are so flexible that they cannot be seen in the cryo-EM volumes. We expect that the GRD behaves largely as a rigid body whose displacements relative to the core domain become larger in the dATP-bound form.

Intriguingly, binding of dTTP to the s-sites at the dimer interface induces two distinct conformations of loop 2 (Fig. 8c). In the active monomer, loop 2 is projected towards the aforementioned network of interactions between NxN flap, GRD and substrate, packing against the flap, though no polar interactions are apparent. However, loop 2 also makes extensive interactions with its equivalent in the inactive monomer, which forces the latter into a conformation where it would sterically clash with an ordered NxN flap (Fig. 8d). Furthermore, Phe412, which stacks on the substrate base in the active monomer, and Arg305, which is projected from the substrate-distal side of loop 2, sterically prevent the substrate binding in the inactive monomer. These may be important contributing factors to the disorder of the NxN flap and GRD. Finally, Gln309, which interacts with the substrate in the active monomer (Figure 8 - figure supplement 2), is oriented away from the substrate through its interactions with the “active” loop 2.

Cryo-EM structure of the ATP-dGTP dimer

As a third insight into the ATP-bound forms of PcNrdD, we solved the structure of PcNrdD in complex with specificity effector dGTP in the s-site and ATP at a concentration where it would both act as allosteric effector at the a-site and as the cognate substrate for dGTP (Table 1; Table 1 - table supplement 3). Density for dGTP is clearly visible in the s-site (Figure 8 - figure supplement 1d) and for ATP in the active site (Figure 8 - figure supplement 2). Again, the two monomers can be divided into an active one and an inactive one. The conformations of loop 2 are once again asymmetric (Figure 8 - figure supplement 4). The conformation in the active monomer is well-defined, while loop 2 in the inactive monomer is more flexible than in the ATP-dTTP-GTP complex, but still compatible with the hypothesis that the asymmetric arrangement of loop 2 is responsible for disorder of the NxN-GRD-substrate network in the inactive monomer.

Cryo-EM structure of the dATP-bound dimer

In micrographs of PcNrdD samples in the presence of 0.5 mM dATP, we observed an approximately 1:1 mixture of dimeric and tetrameric particles, consistent with results from GEMMA. To resolve whether dATP inhibition was linked to oligomerisation, we resolved the structures of each of these oligomers separately. We made a reconstruction of the dATP-bound PcNrdD dimer from 1 009 021 particles to 2.7 Å resolution that was further improved by DeepEMhancer postprocessing. The core domain is very similar to that of the ATP-CTP complex, with an rmsd in Cα positions of 1.2 Å for 1089 Cα atoms considering each dimer as a whole (Table 1; Table 1 - table supplement 4). dATP is clearly visible in the specificity site (Figure 8 - figure supplement 1b).

Strikingly, in contrast to the ATP complexes, the entire GRD is disordered in both monomers of the dATP-bound dimer (Figure 8 - figure supplement 5). No structure is visible after residue 676 at the end of the last strand of the barrel. This suggests that inhibition by binding of dATP to the ATP-cone is coupled to disordering of the C-terminal domain. The NxN flap that covers the top of the substrate in the ATP/CTP-bound form is also disordered, as are the ATP-cones themselves.

Cryo-EM structure of the dATP-bound tetramer

From the same micrographs that revealed the dATP-bound dimers we also resolved dATP-bound tetramers. We obtained a map from cryoSPARC at 2.8 Å, which was again improved with DeepEMhancer (Fig. 9a). Two dATP molecules are bound to each ATP-cone and one at each of the specificity sites, giving a total of 12 visible dATPs per tetramer. The tetramer reveals an arrangement of two dimers (chains A/B and C/D respectively) in which the ATP-cones are all well-ordered and one ATP-cone from each pair mediates interactions between the dimers (Fig. 9b). The ATP-cones have an almost domain-swapped arrangement within each dimer relative to the core domains. This oligomeric arrangement has not previously been observed within the RNR family.

Structure of the dATP-bound PcNrdD tetramer.

a) cryo-EM reconstruction coloured by chain: dimer 1 light blue and pale blue; dimer 2 light green and pale green. The four ATP-cones are at the bottom of the figure. b) View rotated by 90° around a horizontal axis relative to a) and thus viewed from the bottom, showing the ATP-cones. c-d) Cartoon representation of the tetramer from the same angles as panels a) and b) respectively and with the same chain colouring. The two-fold symmetry axis relating the two dimers of the tetramer is shown as a yellow line. The 12 dATP molecules in the tetramer are shown as sticks. The dimer axis of the left-hand dimer’s core domains is shown as a dark blue line and the dimer axis of the ATP-cone pair as a black line. c) Details of the interaction of dATP molecules at the allosteric activity site in the ATP-cone. d) Close-up of the interaction area between the roof of the ATP-cone of one dimer and the core domain of the other dimer (marked with a black box in panel d). The most significant residues are labelled.

dATP binds to the specificity site as in the dimeric dATP-bound form (Figure 8 - figure supplement 1c). The two dATP molecules bound to each ATP-cone do so very similarly to the dATP-loaded ATP-cones of the RNR class I active site subunit from PaNrdA (PDB ID 5IM3) and the class I radical generating subunit from LbNrdB (5OLK) to which the PcNrdD ATP-cones have 29% and 33% sequence identity respectively (Figure 9 – figure supplement 1). The DALI server reveals an rmsd in Cα positions of 1.9 Å and 1.7 Å respectively. The next most similar ATP-cone is that of the RNR transcriptional regulator NrdR from Streptomyces coelicolor in its dodecameric form (Rozman Grinberg et al., 2022) binding two ATP molecules, with 82 aligned residues, rmsd 2.4 Å and sequence identity 17%. Structural similarity to other ATP-cones is listed in Figure 9 – figure supplement 1.

The two dATP molecules bind with their negatively charged triphosphate moieties oriented towards each other (Fig. 9c). No Mg2+ ion is observed between the triphosphate tails, but this may be an artefact of the cryo-EM method, as the PaNrdA and LbNrdB crystal structures suggest that this ion is necessary for charge neutralisation. As in the ATP-bound form, further charge compensation is achieved by extensive coordination by, or proximity of, basic side chains, e.g. Lys7, Arg8, Arg87 and Arg91. As previously observed in PaNrdA and LbNrdB, the adenosine base of the “inner” dATP (dATP1) fits into a pocket under the β-hairpin roof of the ATP-cone and the base of the “outer” nucleotide (dATP2) is sandwiched between an aromatic residue (Tyr83) and the side of the first α-helix in the domain. No direct contacts are seen between the dATP molecules bound to the ATP-cone and residues in the core domain.

Figure 9 – figure supplement 2 shows the differences in binding modes of dATP in the dATP-bound tetramer and ATP in the ATP-CTP dimer. The high mobility of the ATP-cone in the latter precludes a detailed analysis. However, it is clear that the triphosphate tails have very different orientations. In the ATP complex the tail of ATP2 is oriented away from the C-terminal helix towards the base of ATP1, such that the β-phosphate moiety may even H-bond to the base of ATP1. The side chain of Arg87 fills the space vacated by the γ-phosphate of ATP2. Instead of H-bonding to the γ-phosphate of ATP2, Arg91 swings away and interacts with the equivalent moiety of ATP2 in the ATP-cone of the other chain. The structures suggest that the additional 2’-OH group in ATP may cause conformational changes in the ribose ring that propagate to the phosphate tail, but further analysis requires improved resolution.

As in PaNrdA and LbNrdB, the ATP-cones in the dATP-bound tetramer self-interact through a pair of α-helices at the C-terminal end of the ATP-cone. Remarkably, the pair of ATP-cones is highly asymmetrically positioned with respect to the 2-fold axis of its core dimer, being displaced as a rigid body towards the other dimer (Fig. 9b). Both pairs of ATP-cones are displaced in the same way, giving the complex overall C2 symmetry around an axis bisecting the pair of dimers. Consequently, the 2-fold axes of the ATP-cones and the core dimer are not aligned (Fig. 9b). The asymmetric arrangement is associated with two distinct conformations of the linker 93-103 between the fourth helix of the ATP-cone (Figure 9 – figure supplement 3a,b) and the core domain. In chains A and C, a small kink leads from helix 4 into a short helical segment that packs antiparallel to helix 122-142 in the core domain. In contrast, in chains B and D, the short helix is parallel to helix 122-142 and is joined to helix 4 at a 45° angle. The rest of the polypeptide between residues 103-121, including the NxN flap, is disordered in both linkers.

Interestingly, all contacts between the two dimers of the tetramer are mediated by interactions between one of the ATP-cones in one dimer (B or D) and one of the core domains of the other dimer (C or A respectively). Figure 9d shows the interactions of the roof of the ATP-cone with two outer helices from the C-terminal half of the barrel domain (residues 458-485 and 550-569, preceding the sixth and seventh strands). The interactions are mostly hydrophobic but are reinforced by several H-bonds, e.g. between Gln3D-Gln458A, Ser53D–Gln458A, Arg11D-Asp468A, the main chain amide of Ile12D and Tyr557A. The amount of buried surface area is small: 717 Å2, or around 1% of the total surface area of each dimer. For comparison, the monomer-monomer interactions within each dimer bury 11.7% of the total area. The residues on the core domain of PcNrdD involved in interactions with the ATP-cone of the other dimer do not show high sequence conservation, even when the alignments are restricted to the most similar sequences.

Cryo-EM structures of PcNrdD-dATP complexes produced in the presence of CTP

To investigate whether substrate had any allosteric effect on the conformations of the ATP-cone, flap and GRD, we determined separate cryo-EM structures of tetrameric and dimeric PcNrdD, from the same grid, in the presence of 0.5 mM dATP and 0.5 mM CTP, achieving a resolution of 2.6 Å from 1 105 348 particles before post-processing for the tetramer and 2.6 Å from 1 132 695 particles for the dimer (Table 1; Table 1 - table supplement 5). Both forms were almost identical to those seen in the presence of dATP alone, and no CTP could be seen bound to the active site (Figure 6 - figure supplement 3). This strongly suggests that substrate binding is dependent on ordering of the flap region by binding of ATP to the ATP- cone.

Hydrogen-deuterium exchange mass spectrometry (HDX-MS)

To validate our structural and biochemical hypotheses, we further examined the degree of order of the different structural elements of PcNrdD in solution using HDX-MS. This method measures the degree of protection of peptides from H/D exchange in a deuterated solution, which reports on how exposed they are. Three different samples were analysed: apo- PcNrdD and the complexes with dATP-CTP and ATP-CTP.

The apo form shows high deuterium uptake over the entire ATP-cone, linker and NxN flap, from residues 1-117 (Figure 10a). Analysis of differences in uptake in the N-terminal region between the apo protein and the ATP or dATP complexes clearly validates binding of both nucleotides to the ATP-cone (Figure 10b). Peptides spanning residues 2-84 exhibit clear protection from HDX in both nucleotide-bound states. Furthermore, bimodal deuterium uptake (James et al. 2022) was observed for N-terminal peptides spanning residues 2-23 and 83-99. For peptides very close to the N-terminus (2-23), the apo form only exhibits one high uptake (dynamic) conformational state while for the two ligand bound states there was a clear bimodal behaviour with one high uptake and one protected state of an approximately equal ratio (Figure 10c and Figure 10 - figure supplement 1). In-depth analysis of the causes is difficult, but possible interpretations are that the bimodality is due to a mixture of occupied and empty ATP-cones, or more likely to the mixture of dimers and tetramers.

Summary of the HDX-MS experiments.

a) Heat map showing the deuterium uptake of the apo state for time points 30, 300 and 3000 s. Warm colours represent fast uptake and cold colours a low uptake. Both the N-terminal and the C-terminal regions have a high uptake already at 30s. For the N-terminal bimodal uptake, only the fast-exchanging population (rightmost) is depicted. The influence of bimodal uptake and more detail can be seen in Figure 10 - figure supplements 1 and 2. b) Differential chiclet plot (Zhang et al. 2021) collapses the timepoints into a single maximum deuterium uptake difference for states apo- (ATP+CTP), apo-(dATP+CTP) and (dATP+CTP)-(ATP+CTP). Colours range from deep blue for a high degree of relative protection to deep red for high deprotection. Areas marked with dashed red boxes are peptides showing bimodal uptake, i.e. two conformational populations present simultaneously, and black dashed boxes denote areas with peptides having EX2 exchange (only one conformational population in the observed timespan). c) Deuterium uptake plots from dashed areas in b). For peptides 2-23 and 83-99, when there are two lines with the same colour, the top traces are for the fast exchanging population and the bottom for a less dynamic/more protected population. Note that for peptide 83-99 the apo state has 2 populations at all timepoints, while the ATP and dATP converge at 3000s to one (dynamic) population. For peptides 295-312 and 693-713 the apo curve has been removed for visual clarity.

Interestingly, peptides spanning amino acids 81-99 in the linker region have a bimodal uptake that is most pronounced in the apo state. For the nucleotide-bound forms there are also two populations that converge into a single state at 3000s (Figure 10c). This suggests a mixture of conformational states with flexible and locked linkers, consistent with the highly dynamic nature of residues 85-100 observed in the cryo-EM structures, where in the dATP bound tetramer the linker is ordered to residue ∼100, while in the dimer the ATP-cone and linker are highly dynamic and the first ordered structure is observed from residue 120.

The NxN flap at residues 101-117 has a high H/D exchange rate already at the shortest labelling time, with no change at longer times for any state. While surprising in the context of the cryo-EM observations, this might be the result of very fast dynamics in the NxN flap. For the region 118-126, a higher exchange rate in the dATP form than in the ATP form at the longest time point, most likely because substrate is not bound to the active site.

Clear differences between the ATP and dATP complexes are seen in the region 178-204, which contains the β-hairpin of the s-site (Fig. 10c, panel 178-204) that is highly protected with dATP. This is consistent with the structures that show no ATP bound at the s-site but full occupancy of dATP. Peptides between 330-348, which are in contact with the β-hairpin, are also protected in the dATP complex. The region 295-312, i.e. loop 2, is deprotected in the dATP form relative to ATP. This is also in agreement with the cryo-EM structures, in which loop 2 is partially disordered in the dATP complex but ordered in the ATP complex in the monomer with bound CTP.

The HDX-MS results show a high degree of deuterium uptake in the GRD in all forms, indicating a high degree of mobility. This is consistent with the cryo-EM structures, in which the GRD overall has B-factors higher than the core domain (though the glycyl-radical containing loop has similar B-factors to the core, being buried). From the HDX-MS data it is clear that the GRD is highly dynamic, and while a trend of increased dynamics for some C-terminal peptides in the GRD can be seen for the dATP-CTP state, a detailed analysis is confounded by the limited time resolution and the complexity of the sample, which consists of a mixture of conformations.

Discussion

Ribonucleotide reductase is an essential enzyme for all free-living organisms. Its complex regulation at the levels of transcription, overall activity and substrate specificity makes it quite unique in Nature (Mathews, 2016, 2018). A fundamental understanding of this regulation could e.g. pave the way for better antibiotics. The ATP-cone is a small, genetically mobile domain frequently found at the N-terminus of RNR catalytic subunits of all major classes, as well as in the bacterial RNR-specific transcription factor NrdR (Rozman Grinberg et al., 2022). Its role in allosteric regulation of class I aerobic RNRs has been extensively studied. In the presence of dATP, the ATP-cone mediates the formation of a surprising variety of different oligomeric complexes that nevertheless have in common that they block the formation of an (observed or inferred) active complex between the catalytic and radical generating subunits that permits proton-coupled electron transfer on each catalytic cycle (reviewed in Martínez-Carranza et al., 2020).

Despite their catalytic subunits being evolutionarily related to those of class I, anaerobic class III RNRs have a fundamentally different activation mechanism that is likely common to all glycyl radical enzymes (Backman et al., 2017; Lundahl et al, 2022). Once generated on the catalytic subunit by encounter of the reductase with its radical SAM family activase, the glycyl radical can catalyse hundreds of cycles of turnover before being exhausted (Torrents et al., 2001). Allosteric inhibition could thus in principle proceed by two mechanisms: by blocking the initial encounter of the reductase and activase, or by preventing subsequent transfer of the radical to the substrate. The first of these hypotheses is intuitively more similar to the steric blocking mechanisms encountered in class I RNRs; the latter is more challenging to understand, as the glycine and cysteine are buried in close proximity in the active site.

In this work we present the first biochemical, biophysical and structural characterisation of an anaerobic, class III RNR containing an ATP-cone, providing the first picture of the structural basis for allosteric activity regulation in this large family. We show that the ATP-cone can bind two molecules of ATP or dATP, but not both simultaneously. HDX-MS confirms that binding of either nucleotide reduces the exposure of the ATP-cone to deuterium exchange, and supports the biophysical findings of dATP binding to the s-site. As expected, the presence of an ATP-cone confers inhibition of enzyme activity at high dATP concentrations. GEMMA experiments show that PcNrdD exists in a mixture of dimeric and tetrameric states under most experimental conditions studied, with the equilibrium shifted towards dimers in the presence of ATP. Addition of inhibiting concentrations of dATP shifts the equilibrium towards tetramers. The equilibria are not affected by substrate or specificity effector. EPR spectroscopy clearly shows that the glycyl radical is formed even in the presence of dATP. Together, these results show that dATP inhibition in class III RNRs does not occur by blocking the initial encounter of reductase and activase. Instead, our nucleotide binding results and structures show that dATP inhibition prevents binding of substrates.

The seven cryo-EM structures of PcNrdD in a variety of nucleotide complexes presented here suggest a mechanism in which binding of ATP or dATP modulates activity by affecting the dynamics of a tightly knit network of structural elements including the critical glycyl radical domain in the C-terminal region of the enzyme. A summary of the conformational states is shown in Figure 11. In the presence of ATP and substrate CTP, the GRD is fully ordered in one monomer and Gly711 is found at its expected position proximal to Cys416, to which it will deliver its radical to initiate each catalytic cycle and accept it back at the end of the cycle. The CTP substrate is bound in this monomer and a conserved but previously unobserved flap containing two Asn residues (NxN flap) forms over the top of the substrate, making H-bonding interactions with it. Loop 2, which mediates between the substrate specificity site and the active site, is fully ordered. In contrast, in the other monomer, no substrate is bound, loop 2 is disordered and the flap and C-terminus are more dynamic. At the same time, the ATP-cones themselves are relatively flexible. This is not unexpected for this type of ATP-cone, which binds two nucleotides and was first observed in the class I RNR PaNrdA (Johansson et al., 2016). SAXS studies showed that the ATP-cone in PaNrdA is flexible in the presence of ATP, while in its dATP-bound form it forms a symmetric interaction with an ATP-cone on another dimer, causing the enzyme to tetramerise. A similar behaviour was seen even when this type of ATP-cone was evolutionarily “transplanted” to the N-terminus of the radical-generating subunit, as seen in LbNrdB (Rozman Grinberg et al., 2018a).

Schematic showing the different structural states of PcNrdD described in this work.

The GRD is coloured orange where present. Inactive and active monomers are labelled I and A respectively. The a-site in the ATP-cone is labelled with “a” on a green background, the s-site at the dimer interface with “s” on a red background and the active site with “c” on an orange background. Labels are only present when a nucleotide is observed in the structure. The ligands bound are listed, coloured according to the sites they bind to. In the dATP/dATP-CTP tetramer, one of the inactive monomers and its s-site are hidden behind the dimer in the foreground.

By careful isolation of a subset of particles, we were able to characterise an unusual dimer of ATP-cones in the ATP-CTP complex, where four ATP molecules bind with all their triphosphate tails pointing towards each other. Such an orientation is unprecedented for RNRs, but such tail-to-tail orientations, compensated by extensive positive charges, are found in e.g. adenylate kinases (Berry, 1994). The relevance of this arrangement remains to be elucidated. Interestingly, when dTTP was added to the s-site in the ATP-dTTP-CTP complex, the ATP-cones became even more dynamic and could only be observed in one monomer of the dimer as diffuse density. This behaviour was reproduced with dGTP in the s-site. Both complexes with s-site effector are functionally more relevant forms than the ATP-CTP complex, as in the cell, a form with empty s-site is unlikely to exist. The partitioning of monomers into active and inactive is amplified by the presence of s-site effector (dTTP or dGTP) in both complexes with ATP. Only one of the active sites is occupied by substrate, with ordered NxN flap and C-terminal GRD. This is correlated with two distinct conformations of loop 2 in the two monomers. In the inactive monomer, the NxN flap and GRD are disordered, as loop 2 blocks their active conformation and concomitantly substrate binding. These results point to only one of the PcNrdD monomers being active at a given time and also to a significant cross-talk between the a-site, s-site and active site. This is corroborated by the finding that an s-site effector is necessary for retention of ATP by the ATP-cone when PcNrdD is first incubated with these nucleotides then purified without them.

In the presence of inhibitory concentrations of dATP we were able to isolate structures of both dimers and tetramers from the same cryo-EM sample, in agreement with the GEMMA results that show a shift towards the tetrameric form. In both forms, the C-terminal GRD is highly mobile in all monomers such that it cannot be seen in the cryo-EM reconstructions. An exposed glycyl radical is incompatible with its stability, so it is possible that the dynamics of the GRD are exaggerated somewhat by the vitrification process. Consistent with this, the HDX-MS results show a small but noticeable deprotection of peptides from the GRD in the presence of dATP-CTP compared to ATP-CTP.

In the tetramers, pairs of ATP-cones asymmetrically disposed relative to the 2-fold axes of their respective PcNrdD dimers associate through the “roof” of the “inner” dATP binding site to the core domain of the other dimer. While this allowed us to build complete models for the ATP-cones and analyse dATP binding, the role of tetramerisation in the inhibition mechanism is unclear. The tetrameric structures appear to be facilitated by a dATP-induced “stiffening” of the pair of ATP-cones, similar to what was seen in PaNrdA and LbNrdB, and their interaction with a second dimer of PcNrdD, but the interaction area is small (1% of the total surface area) and the residues involved are not highly conserved in NrdDs, even from the most similar sequences. Therefore, further studies of NrdDs containing ATP-cones are required in order to determine whether oligomerization is a general phenomenon in the family. Furthermore, the GRD is mobile in all complexes with dATP in the a-site, whether dimeric or tetrameric. Since both the ATP-cone and the GRD are disordered in the dATP-bound dimers, the exact molecular mechanism by which dATP induces disorder of the GRD remains somewhat elusive.

Taken together, our results show a clear correlation between ATP or dATP binding to the ATP-cone and activity or inhibition respectively. The dynamical transitions involve a previously uncharacterised but highly conserved structural element, the NxN flap, that folds over the substrate and furthermore acts as a structural bridge between the ATP-cone and the C-terminal GRD. An increase in dynamics of these elements prevents substrate binding and thus transfer of the radical from the GRD to the substrate. Interestingly, a similar flap is formed over the top of the active site in the E. coli active complex, but it is formed by part of the C-terminal region of the separate radical generating subunit (Kang, 2020). Thus, a locking of the substrate in the active site appears to be a conserved feature of active class I and class III RNRs. However, in class I RNR there is no tunnel allowing substrate access to the active site in the locked state. This is consistent with the fact that the class I subunits dissociate on each catalytic cycle, while in class III the glycyl radical can catalyse many substrate reductions before having to be regenerated. It remains enigmatic exactly how the tiny chemical difference between dATP and ATP (the 2’-OH group) results in such major conformational changes, as in the ATP-CTP complex, where the ATP-cones are visible, the a-site in the ATP-cone is separated from the NxN flap by at least 30 Å and in the most biologically relevant complexes with additional s-site effector, the ATP-cone is very disordered. Nevertheless, the present results give first insights into allosteric activity regulation in anaerobic RNRs and add yet another aspect to the surprisingly wide range of allosteric conformational changes that can be mediated by the binding of the two highly similar nucleotides ATP and dATP to a small, evolutionarily mobile protein domain.

Materials and Methods

Cloning of nrdD and nrdG from P. copri

The genes encoding the reductase (NrdD) and activase (NrdG) proteins from P. copri were synthesised by GenScript with codon optimization for E. coli and subcloned into the pBG102 plasmid (pET27 derivative) (Center for Structural Biology, Vanderbilt University) between the BamHI and EcoRI restriction sites to produce His6-SUMO-NrdD and His6-SUMO-NrdG protein constructs.

Overexpression and purification of PcNrdD and PcNrdG

Plasmids containing the nrdD gene or the nrdG gene were transformed into E. coli BL21 (DE3) star and E. coli BL21 (DE3) competent cells, respectively. Cells (6 L) were grown at 37 °C in Luria Broth supplemented with kanamycin (50 µg/mL) to an A600 of 1.2. Protein expression was then induced with 0.5 mM IPTG and incubation was extended overnight at 20 °C. After centrifugation, pellets were suspended in 60 mL of lysis buffer 1a (50 mM Tris-HCl pH 8, 500 mM KCl, 0.5 mM TCEP) for PcNrdD and lysis buffer 2a (50 mM Tris-HCl pH 8, 500 mM NaCl, 0.5 mM TCEP) for PcNrdG, supplemented with lysozyme (0.5 mg/mL), and disrupted by sonication. Cell debris was removed by ultracentrifugation at 210 000 x g for 1 h at 4 °C. The supernatant was then loaded on an immobilised metal affinity Ni-NTA column (HisTrap 5 mL; Cytiva) equilibrated with buffer 1b (50 mM Tris-HCl, pH 8, 300 mM KCl) or buffer 2b (50 mM Tris-HCl, pH 8, 300 mM NaCl), supplemented with 50 mM imidazole as appropriate. The column was washed extensively with the corresponding buffer containing 50 mM imidazole and the His6-SUMO tagged proteins were eluted using 300 mM imidazole. The proteins were collected and dialyzed overnight at 4 °C against buffer 1b or 2b supplemented with 1 mM 1,4-dithiothreitol (DTT) in the presence of the PreScission Protease (150 µM) to cleave the affinity-solubility tag. The GST–3C-protease (PreScission) was expressed using pGEX-2T recombinant plasmids. After induction at 25 °C with 0.1 mM IPTG for 20 h, the PreScission Protease was purified using glutathione–Sepharose chromatography. After the dialysis, the cleaved PcNrdD and PcNrdG proteins were centrifuged at 4 °C for 10 min and loaded onto a HisTrap column equilibrated with buffer 1b or 2b, supplemented with 50 mM imidazole and the flow-through containing the cleaved protein was collected. Following cleavage of the His6-SUMO tag, the proteins incorporated a non-native N-terminal Gly-Pro-Gly-Ser sequence. The purified preparations are called as-purified PcNrdD and as-purified PcNrdG, respectively.

A fraction of as-purified PcNrdD was precipitated by addition of HCl, then centrifuged at 15 000 x g for 5 min. UV-visible spectra of the supernatant were recorded before and after protein precipitation to estimate the amount of nucleotide contamination. PcNrdD preparations with ≤5% nucleotide contamination are referred to as apo-PcNrdD.

Fractions containing apo-PcNrdD and as-purified PcNrdG were concentrated and loaded on a gel filtration column (Hiload 16/60 Superdex S200; Cytiva) in 25 mM Tris-HCl, pH 8, 250 mM NaCl, and 5 mM DTT. The purified protein was concentrated to 20 mg/mL with an Amicon ultrafiltration device (100 kDa cutoff for NrdD and 10 kDa cutoff for NrdG; Millipore), frozen in liquid nitrogen, and stored at −80 °C.

[Fe-S] cluster reconstitution and purification of holo-PcNrdG

The reconstitution of the [4Fe-4S] cluster and purification of PcNrdG containing an iron-sulfur cluster (holo-PcNrdG) were performed under strict anaerobic conditions in an Mbraun glove box kept at 18°C and containing less than 0.5 ppm O2. PcNrdG was treated with 5 mM DTT for 10 min then incubated for 3 h with a fivefold molar excess of ferrous ammonium sulphate and L-cysteine in the presence of 2 μM E. coli cysteine desulfurase CsdA. The holo-PcNrdG was then loaded onto a Superdex 200 10/300 gel filtration column (Cytiva) equilibrated in 25 mM Tris-HCl, 250 mM NaCl, and 5 mM DTT (Figure 1 -figure supplement 2). The peak containing the soluble holo-PcNrdG was then concentrated to 10 mg/mL on a Vivaspin concentrator (10 kDa cutoff).

In vitro enzymatic assay and HPLC analysis

Activity assays were performed under strict anaerobic conditions inside a glovebox in 25 mM Tris pH 8, 50 mM KCl, 10 mM DTT in a volume of 100 μL. A standard pre-reaction mixture contained 1 μM apo-PcNrdD, 2.5 μM holo-PcNrdG, 500 µM S-adenosyl-methionine (SAM), 5 mM MgCl2, 1 mM GTP (or 1 mM CTP), and different allosteric effectors dTTP (1 mM), ATP (0 - 5 mM) or dATP (0 - 1 mM). The pre-reaction mixture was incubated for 5 min at 37 °C before adding simultaneously 10 mM sodium formate and sodium dithionite (12.5 molar excess). When GTP was used as substrate, dTTP was used as s-site effector, and ATP or dATP were titrated to the ATP-cone (a-site). When CTP was used as substrate the s-site was empty at the beginning of the titration and ATP or dATP can plausibly bind both to the s- site and to the a-site in these experiments.

The reaction was incubated for 10 min at 37 °C and stopped by the addition of 2.5 μL of 3 M formic acid. Product formation was analysed by HPLC using an Agilent ZORBAX RR StableBond (C18, 4.6 x 150 mm, 3.5 µm pore size) equilibrated with buffer A (10% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Samples of 10 μL were injected and eluted at 0.5 mL/min with a step isocratic flow of 40%-100% buffer B (30% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Compound identification and product quantification were performed by external calibration and NrdD specific activity was determined.

From a direct plot of activity versus concentration of effector, the KL values (the concentration of an allosteric effector that gives half maximal enzyme activity) for binding of effectors to the a-site, were calculated in KaleidaGraph using the equation: Ki for non-competitive dATP inhibition of GTP reduction was calculated in KaleidaGraph using the equation: and KL and Ki for dATP-dependent activation and non-competitive dATP inhibition of CTP reduction were calculated in KaleidaGraph using the equation:

Isothermal titration calorimetry experiments

ITC experiments were carried out at 20°C on a MicroCal ITC-200 system (Malvern Instruments Ltd) in ITC buffer containing 25 mM HEPES, pH 7.5, 100 mM KCl, 10 mM MgCl2 and 0.5 mM TCEP with a stirring speed of 1000 rpm. An initial injection volume was 0.6 µl over a duration of 1.2 s. ApoPcNrdD was prepared in the ITC buffer using SEC, followed by the addition of 1 mM dTTP and 1 mM GTP and incubation at room temperature for 5 minutes prior to loading to the cell. ATP and dATP ligands were directly diluted in the ITC buffer. Titration of ligand into buffer was also performed to control for heats of dilution and/or buffer mismatch. For dATP binding analysis, the concentration of PcNrdD in the cell was 50 to 100 µM and dATP concentration in the syringe 0.6 to 1 mM. For ATP titration into PcNrdD, protein concentration in the cell was 200 µM and ATP concentration in the syringe was 0.6 to 1.5 mM. All subsequent injection volumes were 1.7–2.5 µl over 3.4–5 s with a spacing of 160 s between the injections. Data for the initial injection were not considered. The data were analysed using the one set of sites model of the MicroCal ITC-200 analysis software (Malvern Panalytical). Standard deviations in thermodynamic parameters, N and KD were estimated from the fits of three different titrations.

Microscale thermophoresis

Both binding of nucleotides to the ATP-cone in PcNrdD and binding of substrates CTP and GTP to PcNrdD were assessed using MST. PcNrdD was labelled using Monolith Protein Labeling Kit RED-NHS 2nd generation according to the manufacturer’s protocol. MST was performed using the Monolith NT.115 instrument (NanoTemper Technologies GmbH) at room temperature. Binding of GTP and CTP was assayed in MST buffer containing 40LmM HEPES pH 7, 50LmM KCl, 50 and 100LmM MgCl2, 5 mM DTT, 0.1% Tween-20, 1 mM dTTP (only for GTP binding) and either 5 mM ATP or 1 mM dATP. The 16 reaction tubes were prepared by the addition of 2 μl MST buffer (concentrated 5 times), 5 μl GTP or CTP of the desired concentration and 1 μl RED dye-labelled NrdD in a total volume of 10 μl. Final NrdD concentration in the binding reaction was 11 nM while the binding partner concentrations were between 70 mM and 2 μM for CTP and 50 mM and 1.5 μM for GTP. Binding of ATP and dATP to the ATP-cone was assayed in MST buffer containing 25 mM HEPES pH 7.5, 100LmM KCl, 10LmM MgCl2, 0.5 mM TCEP, 0.1% Tween-20, 1 mM dTTP and 5 mM GTP. A 16-step dilution series of the binding partners was prepared by adding 5 μl buffer to 15 tubes. ATP or dATP was added to the first tube to the final concentration of 4 mM and 5 μl was transferred to the second tube and mixed by pipetting (1:1 dilution series). To each tube of the dilution series 5 μl of PcNrdD was added to reach a final concentration of 11 nM. The samples were transferred to Monolith™ NT.115 Series Premium Capillaries (NanoTemper Technologies GmbH), which were scanned using the MST instrument (100% excitation power, medium MST power). Obtained MST data was analysed and fitted using the MO. Affinity Analysis v2.3 software (NanoTemper Technologies) with default parameters. KD and standard deviation for GTP and CTP binding to PcNrdD in the presence of ATP and for ATP and dATP binding to the ATP-cone were calculated using fits from at least three individual titrations. KDs for CTP or GTP binding to PcNrdD in the presence of dATP were extremely high and could not be reliably determined since the titration curves did not reach a plateau and therefore are only estimates.

Nucleotide binding to the PcNrdD ATP-cone

PcNrdD was incubated with nucleotides at room temperature for 30 minutes and then desalted on a NAP-5 column. The desalted protein was boiled for 10 min, centrifuged, and the supernatant was loaded on a Agilent ZORBAX RR StableBond (C18, 4.6 x 150 mm, 3.5 µm pore size) HPLC column to evaluate the amount of nucleotides retained by the protein. Samples of 100 μL were injected on the column equilibrated with buffer A (10% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide) and eluted at 1 mL/min with a gradient of 40%-100% buffer B (30% methanol, 50 mM potassium phosphate buffer, pH 7, 10 mM tetrabutylammonium hydroxide). Compound identification and product quantification were performed by comparison with injected standards. Three different sets of experiments were performed, each set consisting of incubation with only ATP, only dATP, and a combination of ATP and dATP. In the first set of experiments 75 µM apo-NrdD was incubated with 1 mM ATP or 1 mM dATP, and when 1 mM each of ATP and dATP was used, dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer without nucleotides. In the second set of experiments 75 µM apo-NrdD in presence of 1 mM dTTP and 5 mM GTP was incubated with 3 mM ATP or 1 mM dATP, and when both nucleotides were combined dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer without nucleotides. In the third set of experiments 75 µM apo-NrdD in presence of 2 mM dTTP and 5 mM GTP was incubated with 3 mM ATP or 1 mM dATP, and when both nucleotides were combined dATP was added 15 min after addition of ATP. After 30 min incubation the mixtures were desalted in a buffer containing 0.1 mM dTTP and 1 mM GTP. It was possible to separate all phosphorylation levels of adenosine and deoxyadenosine nucleotides in the first set of experiments, but not in the second and third sets due to overlapping peaks of the s-site and substrate nucleotides (Figure 2 - figure supplement 1).

Gas-phase electrophoretic mobility molecular analysis

The GEMMA instrumental setup and general procedures were as described previously (Kaufman et al., 1996; Rofougaran et al, 2008). Apo-PcNrdD was equilibrated into a buffer containing 100 mM ammonium acetate pH 7.8, then applied onto a Sephadex G-25 chromatography column. In addition, 5 mM DTT was added to the PcNrdD protein solution to increase the protein stability. Prior to analysis, the protein sample was diluted to 2 µM in a buffer containing 100 mM ammonium acetate, pH 7.8, 0.005% (v/v) Tween 20, magnesium acetate (equimolar to the total nucleotide concentration) and the corresponding nucleotides (where indicated). The protein samples were then incubated for 2 min at room temperature, centrifuged and applied to the GEMMA instrument for data collection. Each sample was scanned 5 times and the signal was added together to obtain the traces presented. The experiments were performed with a flow rate driven by 2 psi to minimise noise signals that may appear with elevated nucleotide or protein concentrations at the manufacturer’s recommended flow rate driven by 3.7 psi.

Glycyl radical characterization by EPR

EPR samples were handled under strictly anaerobic conditions inside a glovebox in 25 mM Tris-HCl pH 8, 50 mM KCl, 5 mM DTT in a final volume of 100 μL. A standard pre-reaction mixture containing 75 μM holo-PcNrdG, 1 mM S-adenosyl-methionine, was treated with a 12.5 excess of dithionite and was incubated for 20 min at 37 °C. The pre-mixture was added to a mixture of 50 µM apo-PcNrdD, 10 mM sodium formate, supplemented or not with 1.5 mM ATP or 1 mM dATP and in the presence of 1 mM CTP. The final reaction was incubated for 20 min at 37 °C and 100 µL sample was transferred into a standard EPR tube (Wilmad LabGlass 707-SQ-250M) and stored in liquid nitrogen before recording the EPR spectrum.

X-band CW-EPR spectra of PcNrdD were recorded on a Bruker ELEXYS-E500 spectrometer operating at microwave frequency of 9.38 GHz, equipped with a SuperX EPR049 microwave bridge and a cylindrical TE011 ER 4122SHQE cavity in connection with an Oxford Instruments continuous flow cryostat. Measuring temperature at 30 K was achieved using liquid helium flow through an ITC 503 temperature controller (Oxford Instruments). For achieving non-saturating conditions, a microwave power of 10 µW was applied for all measurements. Other EPR settings were: modulation frequency of 100 kHz; modulation amplitude of 8 G; and accumulation of 16 scans for an optimal S/N ratio. Double integrated spectra were used for intensity analysis.

Cryo-EM sample preparation and data acquisition

The proteins for cryo-EM analysis were all prepared in the same manner. Apo-PcNrdD was incubated with 1 mM NTP or dNTP at room temperature for 5 min, then centrifuged before injection onto a Superdex S200 10/300 column pre-equilibrated with a buffer containing 25 mM HEPES-NaOH pH 7.5, 100 mM KCl, 0.5 mM TCEP, 10 mM MgCl2 and supplemented with varying combinations of 0.5 mM NTP or 0.2 mM dNTP. The relevant peak was collected, then diluted at 0.5 mg/mL and the corresponding nucleotides were added to different final concentrations (0.5-5 mM NTP or 0.5-1 mM dNTP) (Table 1).

The grids were glow-discharged for 60Ls at 20LmA using a GloQube (Quorum) instrument, then prepared in the following manner: 3 µL of sample at 0.5 mg/ml were applied on Quantifoil 1.2, 1.3 300 mesh Cu grids or Quantifoil 2.1, 300 mesh Au grids and plunge-frozen in liquid nitrogen-cooled liquid ethane using a Vitrobot Mark IV (Thermo Fisher Scientific) with a blot force of 1 and followed by 5 s blot time, at 4 °C and 100% humidity. The grids used for each sample are specified in Table 1.

Grids were clipped and loaded into a 300 kV Titan Krios G2 microscope (Thermo Fisher Scientific, EPU 2.8.1 software) equipped with a Gatan BioQuantum energy filter and a K3 Summit direct electron detector (AMETEK). Grids were screened for quality based on particle distribution and density, and images from the best grid were recorded. Micrographs were recorded at a nominal magnification of 105,000. Details of the other data collection parameters used for each sample are given in Table 1.

Cryo-EM data processing

Data processing was performed in cryoSPARC (Punjani et al, 2017). The first steps in processing of all datasets were patch motion correction and patch CTF estimation, followed by curation of exposures based on ice thickness, poor defocus estimation, etc. Other steps are as detailed below. Resolution estimation in cryoSPARC was done using a gold standard Fourier shell correlation value of 0.143 (Rosenthal & Henderson, 2003). Map post-processing was done using DeepEMhancer (Sanchez-Garcia et al., 2021) and post-processing resolution estimation was performed using DeepRes (Ramirez-Aportela et al, 2019). Models were placed in the maps using either phenix.dock_in_map (Liebschner et al, 2019) or Molrep (Vagin & Teplyakov, 1997) in the CCP-EM package (Burnley et al, 2017). Model building was done by alternating rounds of manual building in Coot (Emsley et al, 2010) with real space refinement in phenix.refine (Liebschner et al., 2019). Secondary structure restraints were used throughout. The inclusion of riding hydrogen atoms reduced the number of bad contacts in all models. An AlphaFold2 (Jumper et al., 2021) model of PcNrdD was used in the later stages of building the first model to resolve some ambiguous loops. All final models have good correlations between map and model and good stereochemical properties (Table 1). All structures have been deposited in the Protein Databank and the corresponding volumes deposited in the Electron Microscopy Data Bank with the accession numbers listed in Table 1.

PcNrdD-ATP-CTP complex

Data processing

A total of 17 033 movies were used after curation. A low-pass filtered volume was made in Chimera (Pettersen et al, 2021) from the crystal structure of NrdD from T. maritima (TmNrdD, PDB ID 4COI). From this volume, a set of templates was created and used for template-based picking with a diameter of 100 Å. Due to the large number of micrographs, they were split into two sets. The largest set (16 219 micrographs) was analysed first. About 8.6 million particles were extracted with a 350-pixel box size and classified into 60 2D classes. Eight representative classes containing different orientations (∼3.3 million particles) were used to generate four ab-initio models without symmetry (C1) and particles were subjected to three-dimensional (3D) heterogeneous refinement using four classes. The most populated model containing ∼1 million particles was subjected to homogeneous and non-uniform refinement, which gave a map with an overall FSC resolution of 2.94 Å. A further 435 695 particles were then extracted from the remaining 810 micrographs. These were subjected to 2D classification, and the resulting classes were merged with the larger set, giving 1 199 575 particles. The combined particles were subjected to homogeneous, non-uniform and local refinement with a predefined mask, which gave a map at 2.87 Å (Table 1 - table supplement 1).

Generation of map with better ATP-cone density

The particles were further classified into 50 2D classes, from which 23 classes having 1 040 266 particles were selected. Four ab-initio models were generated and 570 730 particles with slightly better ATP-cone density were refined without symmetry. The particles were then subjected to 3D classification using ten classes of ∼57 000 particles each. The classes having slightly better ATP-cone density were selected for further processing and final refinement without symmetry gave a map with an overall FSC resolution of 3.17 Å from 291 231 particles, which was post-processed using DeepEMhancer (Table 1 - table supplement 1).

Model building and refinement

A partially complete model for one dimer of the dATP complex (see below) was placed in the map using phenix.dock_in_map. The ATP-cones were built based on the ones from the dATP-only tetramers (see below).

PcNrdD-dATP complexes

Tetramers

Data processing for first part

A set of 11 161 movies was used after curation, template-based picking based on the structure of TmNrdD and particle extraction using a box size of 448 pixels gave 5 331 420 particles. Seven 2D classes having 1 171 839 particles were used to make ab-initio models, followed by heterogeneous refinement. The best class having 461 020 particles and volume was used for homogeneous refinement, which gave a map at 3.3 Å. Non-uniform refinement followed by local refinement with a pre-determined mask gave a map at 3.1 Å with no imposed symmetry.

Data processing for second part

A set of 16 667 movies was used after curation. Particles were picked and extracted using the same box size as in the first part, which gave 7 889 234 particles. Ten 2D classes having 1 881 057 particles were used to make two ab-initio models, followed by heterogeneous refinement. The best class with 1 083 657 particles was selected and subjected to homogeneous refinement followed by non-uniform refinement and a local new refinement with a pre-determined mask, which gave a map at 2.9 Å with no imposed symmetry.

Merging of datasets and CTF refinement

The best particles from the first dataset were classified into 80 2D classes from which 36 having 399 523 particles were selected. The particles from the second dataset resulting in the best volume were classified into 100 2D classes, of which 31 having 949 610 particles were selected. These particle sets were merged (giving 1 349 133 particles) and refined against the best 3D volume which gave a map at 2.78 Å. Iterative refinement and NU refinement jobs, ultimately using C2 symmetry, and local CTF refinement with a defocus search range of +/- 2000, gave a map at 2.8 Å with C2 symmetry that was post-processed using DeepEMhancer (Table 1 - table supplement 4).

Model building and refinement

Model building was initially done for one dimer by fitting a homology model generated using SwissModel with T4NrdD as a template to the map using phenix.dock_in_map. Similar homology models for the four ATP-cones were placed by hand into the density followed by rounds of real space refinement. After almost completely rebuilding one dimer, the second dimer was placed into the density and model building and refinement were continued.

Dimers

The first part of the data processing was shared with the PcNrdD-dATP tetramers. In the same micrographs there were a number of 2D class averages that represented dimers. Eight such class averages having 1 916 387 particles were used to make two ab-initio models. This was followed by rounds of heterogeneous refinement from where the best model having 1 009 021 particles was refined with no symmetry to give a map at 2.8 Å. A final non-uniform refinement with C1 symmetry gave a map at 2.6 Å that was post-processed using DeepEMhancer (Table 1 - table supplement 4).

PcNrdD-dATP-CTP complexes

Tetramers

Blob picking was carried out from 11 780 curated micrographs with a maximum diameter of 200 Å and a minimum diameter of 90. In total 4 646 479 particles were then extracted with a box size of 448 pixels. This was followed by 2D classification, ab-initio model generation and heterogeneous refinements. The best class with 1 105 348 particles was used for non-uniform refinement with C2 symmetry, which gave a final map at 2.59 Å resolution that was post-processed using DeepEMhancer (Table 1 - table supplement 5).

Dimers

The initial data processing steps were the same as above. Template based picking from 8 796 curated micrographs using a low-pass filtered volume of the dATP-only tetramer gave 6 836 668 particles after extraction with a box size of 300 pixels. This was followed by 2D classifications, selection of classes representing dimers, ab-initio model generations and refinement of the best class. The final round of non-uniform refinement with 1 132 695 particles with C1 symmetry gave a map of 2.53 Å resolution. The estimated median resolution after post-processing was 2.1 Å (Table 1 - table supplement 5). Model building used the coordinates of the dATP-only dimer as a starting model.

PcNrdD-ATP-dTTP-GTP complex

Data processing

A total of 12 501 movies were processed and 6 105 movies were used after curation. Template-based picking was used with a diameter of 150 pixels. About 3.9 million particles were extracted from 5 952 movies with a 400 pixel box size and used for two-dimensional (2D) classification into 100 classes. Eleven representative 2D classes containing different orientations (∼2.4 million particles) were used to generate an ab-initio model and particles were subjected to three-dimensional (3D) heterogeneous refinement using ten classes, resulting in the most populated model containing 437 866 particles. This model was subjected to homogeneous and non-uniform refinement, which gave a map with an overall FSC resolution of 2.47 Å. The particles were then subjected to 3D classification and the resulting classes were merged for further processing and final refinement without symmetry (C1) gave a map with an overall FSC resolution of 2.4 Å that was post-processed using DeepEMhancer (Table 1 - table supplement 2).

Model building and refinement

A complete model for the dimer of the ATP-CTP structure (see above) was placed in the map using phenix.dock_in_map. The ATP-cones were removed as they were not visible in the reconstruction. After model building and refinement, the final map-to-model correlation value was X. Model quality statistics are presented in Table 1.

PcNrdD-ATP-dGTP complex

Data processing: Template-based picking was carried out on 14 373 curated micrographs and 4 522 028 particles were extracted with a box size of 350 pixels. This was followed by multiple rounds of 2D classifications, ab-initio model generation and refinements. C1 symmetry was applied on the best volume which gave a map with 589 345 particles having a final resolution of 2.58 Å. Model building and refinement used the PcNrdD-ATP/CTP complex as a starting model. (Table 1 - table supplement 3)

Bioinformatics

To construct an HMM logo of NrdD sequences, a representative selection of NrdD sequences from all domains of life including viruses were collected and aligned with Clustal Omega (Sievers et al, 2011). Subsequently an HMM model was built with hmmbuild from the HMMER suite (Eddy, 2011) and a logo was created using the Skylign web service using the “Information Content - Above Background” option (Wheeler et al, 2014). Conserved parts of the logo were extracted and displayed here.

Hydrogen-deuterium exchange mass spectrometry (HDX-MS)

All chemicals were from Sigma Aldrich. pH measurements were made using a SevenCompact pH-meter equipped with an InLab Micro electrode (Mettler-Toledo). A 4-point calibration (pH 2,4,7,10) was made prior to all measurements. The HDX-MS analysis was made using automated sample preparation on a LEAP H/D-X PAL™ platform interfaced to an LC-MS system, comprising an Ultimate 3000 micro-LC coupled to an Orbitrap Q Exactive Plus MS.

HDX was performed on 2 mg/ml PcNrdD with and without ligands (5 mM ATP + 5 mM CTP or 1 mM dATP + 5 mM CTP) in 25 mM HEPES, 100 mM KCl, 20 mM MgCl2 and 0.5 mM TCEP, pH 7.5. The HDX-MS was performed in one continuous run, the apo state being run before both complexed states. For each labelling time point, 3 μl HDX samples were diluted with 27 μl labelling buffer (containing the ligands) of the same composition prepared in D2O, pH(read) 7.4. The HDX labelling was carried out at t = 0, 30, 300, and 3000 s at 20 °C. Each time point was run in 4-6 replicates. The labelling reaction was quenched by dilution of 30 μl labelled sample with 30 μl of 1% trifluoroacetic acid (TFA), 0.4 M TCEP, 4 M urea, pH 2.5 at 1 °C. 50 μl of the quenched sample was directly injected and subjected to online pepsin digestion at 4 °C on an in-house immobilized pepsin column (2.1 x 30 mm). The online digestion and trapping was performed for 4 min using a flow of 50 µL/min 0.1% formic acid (FA), pH 2.5. The peptides generated by pepsin digestion were subjected to on-line SPE on a PepMap300 C18 trap column (1 mm x 15 mm) and washed with 0.1% FA for 60 s. Thereafter, the trap column was switched in-line with a reversed-phase analytical column (Hypersil GOLD, particle size 1.9 µm, 1 x 50 mm). The mobile phases were (A) 0.1% FA and 95% acetonitrile / 0.1% FA (B) and separation was performed at 1 °C using a gradient of 5-50% B over 8 minutes and then from 50 to 90% B for 5 min. Following the separation, the trap and column were equilibrated at 5% organic content until the next injection. The needle port and sample loop were cleaned three times after each injection with mobile phase 5% MeOH / 0.1% FA, followed by 90% MeOH / 0.1% FA and a final wash of 5% MeOH / 0.1% FA. After each sample and blank injection, the pepsin column was washed by injecting 90 μl of pepsin wash solution 1% FA / 4 M urea / 5% MeOH. In order to minimize carry-over, a full blank was run between each sample injection. Separated peptides were analysed on a Q Exactive Plus MS, equipped with a HESI source operated at a capillary temperature of 250°C with sheath gas 12, Aux gas 2 and sweep gas 1 (au). For HDX analysis, MS full scan sspectra were acquired at 70K resolution, AGC 3e6, max IT 200 ms and scan range 300-2000. For identification of generated peptides, separate non-deuterated samples were analysed using data dependent MS/MS with HCD fragmentation. A summary of the HDX experimental detail is reported in Figure 10 -figure supplement 1. The mass spectrometry raw files have been deposited at the ProteomeXchange Consortium via the PRIDE partner repository (Perez-Riverol et al. 2022) with the dataset identifier PXD047943.

Data analysis

PEAKS Studio X Bioinformatics Solutions Inc. (BSI, Waterloo, Canada) was used for peptide identification after pepsin digestion of non-deuterated samples. The search was done on a FASTA file with only the NrdD sequence. The search criteria were a mass error tolerance of 15 ppm and a fragment mass error tolerance of 0.05 Da, allowing for fully unspecific cleavage by pepsin. Peptides identified by PEAKS with a peptide score value of log P > 25 and no modifications were used to generate a peptide list containing peptide sequence, charge state and retention time for the HDX analysis. HDX data analysis and visualisation was performed using HDExaminer, version 3.3 (Sierra Analytics Inc., Modesto, US). The analysis was made on the best charge state for each peptide, allowed only for EX2 (except for bimodal parts of the protein for which EX1 calculation of uptake was allowed) and the two first residues of a peptide were assumed unable to hold deuteration. Due to the comparative nature of the measurements, the deuterium incorporation levels for the peptic peptides were derived from the observed relative mass difference between the deuterated and non-deuterated peptides without back-exchange correction using a fully deuterated sample (Engen & Wales, 2015). As a full deuteration experiment was not made, full deuteration was set to 75% of maximum theoretical uptake. The presented deuteration data is the average of all high and medium confidence results. The allowed retention time window was ± 0.5 minute. The spectra for all time points were manually inspected; low scoring peptides, obvious outliers and any peptides where retention time correction could not be made consistent were removed. As bottom-up labelling HDX-MS is limited in structural resolution by the degree of overlap of the peptides generated by pepsin digestion, the peptide map overlap is shown for the respective state in Figure 10 - figure supplement 1.

Supporting information